### 2022-04-25 GompertzManuscript/elsarticle/elsarticle.tex
initial commit

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{float}
\usepackage{booktabs}       % professional-quality tables
\usepackage{graphicx}
\modulolinenumbers[5]

\journal{Chaos, Solitons \& Fractals}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
% \bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\insertPdfFig}[3]{
  \begin{figure}[H]
  \centering
  \includegraphics[width=12.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}

\newcommand{\insertSmallPdfFig}[3]{
  \begin{figure}[h]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}
\graphicspath{{../figures/}}
\begin{document}

\begin{frontmatter}

\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}

%% Group authors per affiliation:
% \author{Matz Andreas Haugen\fnref{myfootnote}}
% \address{Enerhauggata 1, 0651 Oslo, Norway}
% \fntext[myfootnote]{Since 1880.}

%% or include affiliations in footnotes:
\author[mymainaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
% \ead[url]{www.matzhaugen.com}
\ead{matzhaugen@gmail.com}

\author[mysecondaryaddress]{Dorothea Gilbert}

\address[mymainaddress]{Enerhauggata 1, 0651 Oslo, Norway}
\address[mysecondaryaddress]{University of Oslo, Oslo, Norway}

\begin{abstract}
Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. We show that the Gompertz curve is incompatible with the traditional communicable disease spreading hypothesis, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor. Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. Third, we show that the population acts like a coherent organism under growth/depletion. Finally, we connect the Susceptible-Infected-Recovered (SIR) model with our new theory and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
\end{abstract}

\begin{keyword}
Gompertz \sep Coherence \sep Covid \sep Coronavirus \sep Network Analysis \sep Stochastic
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}

Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. Following this line of reasoning it can be shown that in the early stages of an epidemic, growth follows an initial exponential growth under a constant ratio of the infection rate and recovery rate. After the initial period, a logistic-like growth curve emerges, for which analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately,carletti2020covid}. However, instead of showing logistic-like growth, observed cumulative mortality exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
\begin{equation}
\label{eq:GompertzODE}
\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
\end{equation} 
% Redundant definitions for full script
with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
\begin{equation}
\label{eq:gomp_solution}
Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}}
\end{equation}
$Y_{0}=Y(t = 0)$, $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$ (see Fig. \ref{fig:Gompertz_concept} for a conceptual image). Rarely do macroscopic observations reliant on social networks behave so closely to a model \citep{broido2019scale}.

\insertSmallPdfFig{Gompertz_concept}{A conceptual display of the discrepancy seen between the logistic growth and the observations, superimposed by a Gompertz curve. Data is transformed to the straight line domain of the Gompertz curve for visual clarity, $f(Y)=\ln{\ln{1/Y}}$.}

This phenomenon is recorded by \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves. What is causing such a discrepancy between reality and the current theory of communicable diseases \citep{castro2020turning}?

One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery ratio, $\phi(t)$. How would $\phi(t)$ have to behave? Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\\
\frac{dI}{dt}& = -\beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I,
\end{align}
omiting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the total population, $N \gg Y \ge I$, which seems well founded at the national level,
\begin{align}
\label{eq:linSIR}
\frac{dY}{dt}& = \beta I\\
\frac{dI}{dt}& = (\beta - \alpha) I = \alpha (\phi(t) - 1) I,
\end{align}
and where the number of recovered, $R(t)$, is under this linearization decoupled from the other variables.

They further assume the number of diseased is proportional to the number of cumulative infected, offset by a time lag, which allows us to use the same set of linearized equations to model the number of cumulative people diseased without loss of generality.

 Due to this linearlization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions. And since $I=I(Y)$, we combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
\begin{subequations}
\label{eq:rypdal}
\begin{align}
\frac{dY}{dt} =& \gamma(t) Y(t) \label{eq:rypdalODE}\\
\gamma(t) =& \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] \label{eq:rypdalGamma},
\end{align}
\end{subequations}
where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$. This parametrization is the commonly used Richard's growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth. Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}. Note that at $\theta=1$, the traditional logistic growth curve is obtained, while at $\theta\rightarrow \infty$ we recover the exponential (Malthusian) explosion.

The observed Gompertzian mortality curves are realized in the limit $\theta \rightarrow 0$, with the relative growth rate, 
\begin{equation}
\label{eq:rypdalLimit}
\gamma(t) = \lim_{\theta \rightarrow 0}\frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right]
= \gamma_{\infty}\ln{\frac{Y_{\infty}}{Y}}
\end{equation}

 At this limit the growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. One would rather expect a near-constant relative growth rate in the early times. \citet{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries. Perhaps a more likely scenario is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. On the other hand, an infection of peripheral nodes should cause immediate exponential growth. Either case seems likely, and subject to initial conditions. \citet{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.

But even without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain growth which is almost Gompertzian, i.e. a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow \citet{carletti2020covid} and consider the extented version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of ODEs (\ref{eq:SIR}), with a growth equation of the diseased group at some rate $d$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

Some algebraic manipulations reveal that the diseased group is described by a single ODE, viz.
\begin{equation}
\label{eq:SIRD_D}
\frac{dD}{dt} = \eta( 1 - e^{-\xi D}) - \kappa D,
\end{equation}
where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonomous to the initial susceptible pool of people. This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $d$ and $a$. 
When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.

Note here that at small values of D follow logistic growth,
\begin{equation}
\frac{dD}{dt} = b D (1 - D/K),
\end{equation}
obtained with a second order expansion of the exponential term in \ref{eq:SIRD_D} and with $b=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$. The SIR model exhibits similar properties and its discrepancy with initial observations has been noted by others \citep{vattay2020forecasting}.

With the full 3-parameter single ODE of the diseased group in the SIRD model, one can get almost arbitrarily close to the Gompertzian straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity. 
Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$, an effect which is manifested as a tapering of the initial straight line in the double-log domain.
These comparisons can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_ymax} and Supplemental Information).

Note also that the macroscopic SIRD model does not consider network effects but assumes rather all entities to be connected and communicating at each time-step, usually justified through mean-field approximations \citep{smilkov2014beyond}. 
It is however odd that observed mortality never exhibits a stage of near-exponential growth as this model predicts, but rather a constant slope in the double-log domain. We are thus prompted to look for another model which can explain the observations.

\insertPdfFig{Gompertz_vs_SIRD_lnln_ymax}{Cumulative diseased transformed by $f(Y)=\ln{\ln{({Y_{max} / Y(t))}}}$ plotted against number of days when $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text after normalizing, although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}). The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm. Each plot is annotated with an inferred transmission rate in the SIRD model, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

\section{An alternative theory for observed mortality}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability. This could explain the immediate dampening of mortality growth we see. 
Inspired by  \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term at the microscopic level along with the countering force of the immune system. 
As with all multiplicative processes, it is convenient to log-transform normalized mortality with variables $F(t)=Y(t)/Y_{\infty}$ and $Z=\ln{F}$, from which a natural perturbation model emerges,
\begin{equation}
\label{eq:microscopic}
dZ(t)= -\beta Z(t) dt + \sqrt{\sigma}dW(t),
\end{equation}
where $W(t)$ is a delta-correlated Wiener process\footnote{This perturbation model is also called an Ornstein–Uhlenbeck process}. Notice that this is just a continuous version of an AR(1) model. The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $K=\exp{(-\frac{\sigma}{2\beta})}$, viz. 
\begin{equation}
\label{eq:microscopicOriginal}
dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{K}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
\end{equation}

The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response. 
Note that mortality is treated here as dimensionless where the microscopic treatment would interpret $F$ as cumulative probability of mortality, and the macroscopic equivalent would be the cumulative mortality.

To go from the microscopic to the macroscopic domain, we take the temporal average of (\ref{eq:microscopic}), and use the property that the average of log-quantities is the logarithm of the median quantity, to obtain the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}). We now see that by comparing (\ref{eq:rypdalODE}), (\ref{eq:rypdalLimit}) and the temporal average of (\ref{eq:microscopicOriginal}) that 
\begin{equation}
\sigma = 2\gamma_{\infty}\ln{(Y_{\infty})},
\end{equation}
showing the relationship between the diffusion coefficient and the final conditions.

Thus, a more direct interpretation of the observations is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate \citep{boxenbaum2017hypotheses,neafsey1988gompertz}. Here, a geometric rate is an exponential rate in the log-transformed domain, which is the natural transformation for many processes in nature \citep{zhang1994log}. The log-normal distribution of the abundance of $F$ can be obtained by solving the microscopic stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}. In fact, \citet{sitaram1984statistical} showed using methods from theormodynamics that a variable undergoing Gompertz growth is log-normally distributed in its abundance (see also \citet{gunasekaran1982lon}). This result is intuitively seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.

\section{Unifying the SIR model and the Gompertz model}
The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,

\begin{align}
\label{eq:compare}
\begin{split}
\dot{F} = \frac{d}{dt}F(t) = \beta F (1-F) \quad&\quad \text{Logistic}\\ 
\frac{d}{dt}\ln{(F(t))} = -\beta \ln{(F(t)}) &\quad \text{Gompertz}
\end{split}
\end{align}

In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in an SIR model, but also as a linear interaction term between the two macroscopic states. Many studies do not consider microscopic effects at all, and the solution is obtained by solving the macroscopic (set of ODEs) \citep{vattay2020forecasting}.
This procedure is a mean field approximation with an implied average correlation between the variables. 
Thus, all dynamics are governed by macroscopic deterministic variables parametrized by a transmission rate.

Because the macroscopic logistic ODE solution never invokes the microscopic world, an imaginary microscopic solution could be modeled by splitting the system into $N$ units, $F(t) \rightarrow f_1(t), f_2(t),$ etc, from which macroscopic growth would be obtained by taking the average over the individual units,
\begin{equation}
F(t) = \frac{1}{N}\sum_i^N f_i(t).
\end{equation}

One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compare}), 
\begin{equation}
\label{eq:networkSIR}
\frac{d f_i}{dt} = \beta (1-f_i)\sum_j{a_{ij}}f_j \quad \forall i,
\end{equation}
with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise. Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below. Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
\begin{equation}
\frac{\partial^2 \dot{F}}{\partial f_i \partial f_j} = -\frac{\beta}{N} a_{i,j}.
\end{equation}
Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.

In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus correlated mortality growth, viz.
\begin{equation}
F(t) = \exp{\left[\frac{1}{N}\sum_i^N z_i(t)\right]} = \left[\prod_i^N f_i(t)\right]^{1/N},
\end{equation}
where the rightmost equality is due to the central limit theorem. 
Under this model, correlation between entities is present at all orders in the original domain\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - F(t))} = -b$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - F(t) = \exp{(-bt)}$ \citep{shklovskii2005simple}. In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.

The remarkable observation that we see the Gompertz curve emerge at the macroscopic level implies that the system is indeed correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic distributions \citep{zhang1994log}. 
We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$. \citet{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response. 
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
In the quantum mechanical world, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of the electromagnetic field.

However, it is possible to reconcile the SIR model with the Gompertz curve from yet another argument. 
Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
Then, we reverse the causality where the population of infected are now dependent of the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}). 
This latter change is a crucial change which is based on an environmental stressor rather than a communicable disease.
Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
\begin{equation}
\sum_i a_{ij} = 1 \quad \forall j.
\end{equation} 
If we let $s_i = 1 - f_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model becomes 
\begin{equation}
\frac{d f_i}{dt} = \beta f_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha f_i\quad \forall i,
\end{equation}

Now use the Taylor series $x+x^2/2+... = -\ln(1-x)$, viz.

\begin{equation}
\frac{d f_i}{dt} = \beta f_i\sum_j{a_{ij}}\ln{(1 - s_{j})} - \alpha = \beta f_i\sum_j{a_{ij}}\ln{f_{j}} - \alpha f_i \quad \forall i,
\end{equation}
As we are interested in the aggregate behavior, we take averages in the log domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to the familiar Gompertz curve,
\begin{equation}
\frac{d}{{dt}} \ln\left[\prod_i{f_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i{f_i}\right]^{\frac{1}{N}} - \alpha.
\end{equation}
In this exposition, all the nodes in the network are communicating instantaneously. Lastly, if $f_i \ll 1, \forall i$, i.e. in the beginning of growth, the constant on the right hand side can be omitted. Another interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second constant parameter. One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $f_i$.

\section{Conclusion}
In conclusion, we have shown that Gompertz growth follows from infinite interactions between the susceptible and infected states, and the perceived pathogen travels at infinite speed throughout the population rejecting the possibility of a disease propagation stage through a perceived social network. In this vein, Richard's parameter $\theta$ in his growth model paradigm can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

We further show that the observations can be explained by a model where the biological system is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which gradual return to pre-epidemic conditions are mediated. The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities are present at all orders. From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from the quantum world \citep{molski2003coherent}.

Thus, we see that we are at the intersection of a road: In one direction we find the non-collaborative growth models with parameterized linear interaction effects, and in the other direction we see a field of microscopic entities coherently sharing information much like quantum entangled particles. The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology. One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of diatomic molecules.

\section*{Supplement}

\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

\begin{table}[H]
\centering
\caption{Estimated values for the SIRD model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_ymax} using notation from \ref{eq:SIRD_D}. Note that due to normalization values will be different from \citet{carletti2020covid}.}
\begin{tabular}{llll}
\toprule
{} &           $\eta\times 10$ &      $\xi$ &            $\kappa\times 10^{2}$ \\
\midrule
Belgium        &    1.0(1) &  2.9(3) &   9.3(1) \\
Portugal       &   0.55(4) &  4.2(4) &   4.5(4) \\
Austria        &    1.3(3) &  2.3(4) &    11(2) \\
Switzerland    &   1.24(1) &  2.4(2) &  11.0(7) \\
Sweden         &    0.5(2) &    3(1) &     3(2) \\
Italy          &   0.43(1) &  5.6(2) &   3.7(1) \\
France         &   0.88(7) &  3.4(3) &   8.1(6) \\
Spain          &  0.525(1) &  6.2(2) &   4.8(1) \\
Germany        &   0.78(8) &  3.0(3) &   6.5(7) \\
Denmark        &   0.50(4) &  5.2(6) &   4.2(4) \\
United Kingdom &   0.54(2) &  4.7(2) &   4.3(2) \\
Netherlands    &   0.51(3) &  5.0(4) &   4.3(3) \\
Turkey         &   0.84(9) &  2.8(3) &   6.9(8) \\
Iran           &  0.322(7) &  7.2(3) &  2.51(9) \\
US             &  0.339(8) &  7.0(3) &   1.8(1) \\
China          &    3.8(9) &  1.2(2) &    26(4) \\
\bottomrule
\end{tabular}

\end{table}

\begin{table}[H]
\centering
\caption{Estimated values for the Gompertz model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_ymax} using notation from \ref{eq:GompertzODE}.}
\begin{tabular}{llll}
\toprule
{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
\midrule
Belgium        &   2.74(1) &   8.1(2) &  3.6(3) \\
Portugal       &  2.242(6) &  6.23(9) &  3.1(1) \\
Austria        &   2.74(1) &   8.1(2) &  3.6(2) \\
Switzerland    &  2.740(7) &   8.1(1) &  3.6(2) \\
Sweden         &   1.73(3) &   4.3(3) &  2.6(4) \\
Italy          &  2.107(6) &  5.88(9) &  3.0(1) \\
France         &  2.752(1) &   8.2(2) &  3.6(2) \\
Spain          &  2.556(8) &   7.5(1) &  3.4(2) \\
Germany        &  2.335(7) &   6.5(1) &  3.2(1) \\
Denmark        &   2.32(1) &   6.6(2) &  3.2(3) \\
United Kingdom &  2.333(6) &  6.54(9) &  3.2(1) \\
Netherlands    &  2.311(7) &   6.5(1) &  3.2(1) \\
Turkey         &  2.306(7) &   6.4(1) &  3.2(1) \\
Iran           &  1.862(7) &  5.04(9) &  2.7(1) \\
US             &  2.014(8) &   5.3(1) &  2.9(1) \\
China          &  3.155(8) &   9.4(2) &  3.6(2) \\
\bottomrule
\end{tabular}

\end{table}

% \section*{References}

\bibliography{../../references2}

\end{document}

### 2022-04-25 main.tex
initial commit

--- 2022-04-25 GompertzManuscript/elsarticle/elsarticle.tex
@@
-\documentclass[review]{elsarticle}
+\documentclass{article}
+% \documentclass[review]{elsarticle}
 
-\usepackage{lineno,hyperref}
-\usepackage{amsmath}
+\usepackage{arxiv}
+\usepackage{booktabs}
+\usepackage{siunitx}
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
 \usepackage{amsfonts}       % blackboard math symbols
 \usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{lipsum}   % Can be removed after putting your text content
+\usepackage{graphicx}
+\usepackage[square,numbers]{natbib}
+\usepackage{doi}
+\usepackage{amsmath}
 \usepackage{float}
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{graphicx}
-\modulolinenumbers[5]
-
-\journal{Chaos, Solitons \& Fractals}
-
-%%%%%%%%%%%%%%%%%%%%%%%
-%% Elsevier bibliography styles
-%%%%%%%%%%%%%%%%%%%%%%%
-%% To change the style, put a % in front of the second line of the current style and
-%% remove the % from the second line of the style you would like to use.
-%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Numbered
-% \bibliographystyle{model1-num-names}
-
-%% Numbered without titles
-%\bibliographystyle{model1a-num-names}
-
-%% Harvard
-%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
-
-%% Vancouver numbered
-%\usepackage{numcompress}\bibliographystyle{model3-num-names}
-
-%% Vancouver name/year
-%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
-
-%% APA style
-%\bibliographystyle{model5-names}\biboptions{authoryear}
-
-%% AMA style
-%\usepackage{numcompress}\bibliographystyle{model6-num-names}
-
-%% `Elsevier LaTeX' style
-\bibliographystyle{elsarticle-num}
-%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \centering
-  \includegraphics[width=12.4cm]{#1.pdf}
+  \includegraphics[width=8.4cm]{#1.pdf}
   \caption{#2}
@@
 }
-\graphicspath{{../figures/}}
+
+% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
+\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
+\graphicspath{{figures/}}
+%\date{September 9, 1985} % Here you can change the date presented in the paper title
+%\date{}          % Or removing it
+
+\author{Matz A. Haugen, \\
+Independent Reseacher, Oslo\\
+\texttt{matzhaugen@gmail.com} \\
+  \And
+  Dorothea Gilbert \\
+  University of Oslo\\
+  %% \AND
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+  %% \And
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+  %% \And
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+}
+
+% Uncomment to remove the date
+%\date{}
+
+% Uncomment to override  the `A preprint' in the header
+%\renewcommand{\headeright}{Technical Report}
+%\renewcommand{\undertitle}{Technical Report}
+\renewcommand{\shorttitle}{On the nature of the Gompertz}
+
+%%% Add PDF metadata to help others organize their library
+%%% Once the PDF is generated, you can check the metadata with
+%%% $ pdfinfo template.pdf
+\hypersetup{
+pdftitle={A template for the arxiv style},
+pdfsubject={q-bio.NC, q-bio.QM},
+pdfauthor={Matz A. Haugen},
+pdfkeywords={},
+}
+
 \begin{document}
-
-\begin{frontmatter}
-
-\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
-
-%% Group authors per affiliation:
-% \author{Matz Andreas Haugen\fnref{myfootnote}}
-% \address{Enerhauggata 1, 0651 Oslo, Norway}
-% \fntext[myfootnote]{Since 1880.}
-
-%% or include affiliations in footnotes:
-\author[mymainaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
-\cortext[mycorrespondingauthor]{Corresponding author}
-% \ead[url]{www.matzhaugen.com}
-\ead{matzhaugen@gmail.com}
-
-\author[mysecondaryaddress]{Dorothea Gilbert}
-
-\address[mymainaddress]{Enerhauggata 1, 0651 Oslo, Norway}
-\address[mysecondaryaddress]{University of Oslo, Oslo, Norway}
+\maketitle
 
@@
 
-\begin{keyword}
-Gompertz \sep Coherence \sep Covid \sep Coronavirus \sep Network Analysis \sep Stochastic
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
+% keywords can be removed
+\keywords{Gompertz \and Coherence \and Covid \and Coronavirus \and Network Analysis \and Stochastic}
 
 \section{Introduction}
+Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
+Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
+A large body of research has emerged that explores how these models fit the recent mortality seen due to the Coronavirus epidemic \citep{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
 
-Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission.
-Following this line of reasoning it can be shown that in the early stages of an epidemic, growth follows an initial exponential growth under a constant ratio of the infection rate and recovery rate.
-After the initial period, a logistic-like growth curve emerges, for which analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately,carletti2020covid}.
-However, instead of showing logistic-like growth, observed cumulative mortality exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
+However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
 \begin{equation}
@@
 \end{equation}
-$Y_{0}=Y(t = 0)$, $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$ (see Fig.
-\ref{fig:Gompertz_concept} for a conceptual image).
-Rarely do macroscopic observations reliant on social networks behave so closely to a model \citep{broido2019scale}.
+$Y_{0}=Y(t = 0)$, $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.
 
-\insertSmallPdfFig{Gompertz_concept}{A conceptual display of the discrepancy seen between the logistic growth and the observations, superimposed by a Gompertz curve.
+% \insertSmallPdfFig{Gompertz_concept}{A conceptual display of the discrepancy seen between the logistic growth and the observations, superimposed by a Gompertz curve.
 Data is transformed to the straight line domain of the Gompertz curve for visual clarity, $f(Y)=\ln{\ln{1/Y}}$.}
@@
 
-One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery ratio, $\phi(t)$.
+One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$.
 How would $\phi(t)$ have to behave?
@@
 \end{align}
-omiting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.
+omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.
 
@@
 
- Due to this linearlization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
+ Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
 And since $I=I(Y)$, we combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
@@
 where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$.
-This parametrization is the commonly used Richard's growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth.
+This parametrization is the commonly used Richard's growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \citep{wu2020generalized}.
 Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}.
@@
  The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning.
-One would rather expect a near-constant relative growth rate in the early times.
+ One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
  \citet{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
-Perhaps a more likely scenario is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth.
+
+ Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
  On the other hand, an infection of peripheral nodes should cause immediate exponential growth. 
-Either case seems likely, and subject to initial conditions.
  \citet{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. 
@@
 
-But even without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain growth which is almost Gompertzian, i.e.
+Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain from the communicable disease models growth which is almost Gompertzian, i.e.
 a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
@@
 with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$.
-To show this, we follow \citet{carletti2020covid} and consider the extented version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of ODEs (\ref{eq:SIR}), with a growth equation of the diseased group at some rate $d$ relative to the current infected group,
+To show this, we follow \citet{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
+This requires an addition to our original set of equations (\ref{eq:SIR}), with a growth equation of the diseased group at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 
-Some algebraic manipulations reveal that the diseased group is described by a single ODE, viz.
+Some algebraic manipulations reveal that the diseased group is described by a single equation, viz.
 \begin{equation}
@@
 \end{equation}
-where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonomous to the initial susceptible pool of people.
-This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $d$ and $a$.
+where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonymous to the initial susceptible pool of people.
+This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$.
 When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.
@@
 
-With the full 3-parameter single ODE of the diseased group in the SIRD model, one can get almost arbitrarily close to the Gompertzian straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity.
-Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$, an effect which is manifested as a tapering of the initial straight line in the double-log domain.
-These comparisons can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_ymax} and Supplemental Information).
+With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity.
+Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters obtainable through linear regression, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$, an effect which is manifested as a tapering of the initial straight line in the double-log domain.
+Thus, through linear regression estimates, the Gompertz model avoids the possibility of non-identifiability issues and unstable optima \citep{roda2020difficult}.
+Comparisons between the two models and the inevitable concavity of the SIRD model can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig.
+\ref{fig:Gompertz_vs_SIRD_lnln_ymax} and Supplemental Information).
 
-Note also that the macroscopic SIRD model does not consider network effects but assumes rather all entities to be connected and communicating at each time-step, usually justified through mean-field approximations \citep{smilkov2014beyond}.
-It is however odd that observed mortality never exhibits a stage of near-exponential growth as this model predicts, but rather a constant slope in the double-log domain.
-We are thus prompted to look for another model which can explain the observations.
+Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond}.
+However, this model does not consider network effects explicitly.
+Rather, all entities are connected and communicating instantaneously as shown by \citet{mombach2002mean}.
+However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.
+We are thus prompted to look for another model which can explain the observed mortality patterns.
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_ymax}{Cumulative diseased transformed by $f(Y)=\ln{\ln{({Y_{max} / Y(t))}}}$ plotted against number of days when $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_ymax}{Cumulative number of diseased, transformed by $g(Y)=\ln{\ln{({Y_{max} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD with a Gompertz model.
 Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text after normalizing, although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}).
@@
 An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.
-This could explain the immediate dampening of mortality growth we see.
-Inspired by  \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term at the microscopic level along with the countering force of the immune system. 
-As with all multiplicative processes, it is convenient to log-transform normalized mortality with variables $F(t)=Y(t)/Y_{\infty}$ and $Z=\ln{F}$, from which a natural perturbation model emerges,
+This could explain the immediate dampening of mortality growth observed. 
+It could also explain the lack of correlation between population density and mortality (or infection) rates observed by some \citep{Hamidi2020,Hamidi2020a,Carozzi2020,Arpino2020,khavarian2021high,barak2021urban}.
+
+Inspired by  \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term at the along with the countering force of the immune system. 
+As with all multiplicative processes, it is convenient to log-transformed mortality with dimensionless variables $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
 \begin{equation}
@@
 \end{equation}
-where $W(t)$ is a delta-correlated Wiener process\footnote{This perturbation model is also called an Ornstein–Uhlenbeck process}.
-Notice that this is just a continuous version of an AR(1) model.
+where $dW(t)$ is a delta-correlated Wiener process with zero mean, making $Z(t)$ a stochastic process.
+This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}.
+It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}.
 The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $K=\exp{(-\frac{\sigma}{2\beta})}$, viz.
@@
 The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response.
-Note that mortality is treated here as dimensionless where the microscopic treatment would interpret $F$ as cumulative probability of mortality, and the macroscopic equivalent would be the cumulative mortality.
 
-To go from the microscopic to the macroscopic domain, we take the temporal average of (\ref{eq:microscopic}), and use the property that the average of log-quantities is the logarithm of the median quantity, to obtain the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}).
-We now see that by comparing (\ref{eq:rypdalODE}), (\ref{eq:rypdalLimit}) and the temporal average of (\ref{eq:microscopicOriginal}) that
+To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}) and transform back to the original domain, 
 \begin{equation}
-\sigma = 2\gamma_{\infty}\ln{(Y_{\infty})},
+d\langle\ln{F(t)}\rangle = -\beta \langle\ln{F(t)}\rangle dt.
 \end{equation}
-showing the relationship between the diffusion coefficient and the final conditions.
+Then use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $m(t)$, 
+\begin{equation}
+\label{eq:medianGomp}
+d\ln{m(t)} = -\beta \ln{m(t)} dt,
+\end{equation}
+which corresponds with the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}) with $Y(t) = Y_\infty m(t)$. 
+Comparing the stochastic stressor $\sigma$ in \ref{eq:microscopicOriginal} with the deterministic growth equation in (\ref{eq:GompertzODE}) gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
+We also see that by comparing (\ref{eq:rypdalODE}) and (\ref{eq:rypdalLimit}) with the stochastic counterpart in (\ref{eq:microscopicOriginal}) that the final growth level is governed by the stressor magnitude,
+\begin{equation}
+\sigma = 2\gamma_{\infty}\ln{(Y_{\infty})}.
+\end{equation}
 
@@
 Here, a geometric rate is an exponential rate in the log-transformed domain, which is the natural transformation for many processes in nature \citep{zhang1994log}. 
-The log-normal distribution of the abundance of $F$ can be obtained by solving the microscopic stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}.
-In fact, \citet{sitaram1984statistical} showed using methods from theormodynamics that a variable undergoing Gompertz growth is log-normally distributed in its abundance (see also \citet{gunasekaran1982lon}).
-This result is intuitively seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
+The log-normal distribution of the abundance of $F(t)$ can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon}.
+Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
 
@@
 \begin{split}
-\dot{F} = \frac{d}{dt}F(t) = \beta F (1-F) \quad&\quad \text{Logistic}\\ 
-\frac{d}{dt}\ln{(F(t))} = -\beta \ln{(F(t)}) &\quad \text{Gompertz}
+\dot{F} = \frac{d}{dt}F(t) = \beta F(t) (1-F(t)) \quad&\quad \text{Logistic}\\ 
+\frac{d}{dt}\ln{(F(t))} = -\beta \ln{(F(t)}) &\quad \text{Gompertz},
 \end{split}
 \end{align}
+where $F$ is here deterministic.
 
 In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in an SIR model, but also as a linear interaction term between the two macroscopic states. 
-Many studies do not consider microscopic effects at all, and the solution is obtained by solving the macroscopic (set of ODEs) \citep{vattay2020forecasting}.
-This procedure is a mean field approximation with an implied average correlation between the variables. 
+As mentioned earlier, this procedure is a mean field approximation with an implied average interaction between the variables. 
 Thus, all dynamics are governed by macroscopic deterministic variables parametrized by a transmission rate.
 
-Because the macroscopic logistic ODE solution never invokes the microscopic world, an imaginary microscopic solution could be modeled by splitting the system into $N$ units, $F(t) \rightarrow f_1(t), f_2(t),$ etc, from which macroscopic growth would be obtained by taking the average over the individual units,
+A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $F(t) \rightarrow f_1(t), f_2(t),$ etc, from which macroscopic growth would be obtained by taking the average over the individual units,
 \begin{equation}
-F(t) = \frac{1}{N}\sum_i^N f_i(t).
+F(t) = \frac{1}{N}\sum_i^N f_i(t),
 \end{equation}
+where the lower case $f_i$ emphasize the microscopic quality of the variables, and is here interpreted as probability of infection.
 
-One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compare}), 
+One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compare})
 \begin{equation}
@@
 \end{equation}
-with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
+using the shorthanded $f_i=f_i(t)$ and with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
 Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below.
@@
 
-In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus correlated mortality growth, viz.
+In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus suggests correlated mortality growth, viz.
 \begin{equation}
@@
 \end{equation}
-where the rightmost equality is due to the central limit theorem. 
+where the rightmost equality is due to the central limit theorem where $z_i$ is a random variable as in \ref{eq:microscopic} while $f_i$ is here deterministic. 
 Under this model, correlation between entities is present at all orders in the original domain\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - F(t))} = -b$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - F(t) = \exp{(-bt)}$ \citep{shklovskii2005simple}.
@@
 
-The remarkable observation that we see the Gompertz curve emerge at the macroscopic level implies that the system is indeed correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic distributions \citep{zhang1994log}.
+Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities \citep{zhang1994log}.
 We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
-This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$.
+This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
 \citet{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
 This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
-In the quantum mechanical world, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of the electromagnetic field.
+In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields.
 
+\subsection{Unification through a modified SIR model}
 However, it is possible to reconcile the SIR model with the Gompertz curve from yet another argument. 
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
-Then, we reverse the causality where the population of infected are now dependent of the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}). 
-This latter change is a crucial change which is based on an environmental stressor rather than a communicable disease.
+Then, we reverse the causality where the population of infected are now dependent of the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
+\begin{equation}
+\label{eq:reverseNetworkSIR}
+\frac{d f_i}{dt} = \beta f_i\sum_j{a_{ij}}(1-f_j) \quad \forall i.
+\end{equation} 
+This crucial change is based on the environmental stressor hypothesis rather than a communicable disease assumption.
 Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
@@
 \end{equation}
-As we are interested in the aggregate behavior, we take averages in the log domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to the familiar Gompertz curve,
+As we are interested in the aggregate macroscopic behavior, we take averages in the log domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
 \begin{equation}
@@
 \end{equation}
-In this exposition, all the nodes in the network are communicating instantaneously.
-Lastly, if $f_i \ll 1, \forall i$, i.e.
-in the beginning of growth, the constant on the right hand side can be omitted.
-Another interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second constant parameter.
-One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $f_i$.
+If $\alpha\rightarrow 0$, then this equation will exhibit Gompertz growth in the geometric ensemble average of microscopic units.
+The interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
+One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $f_i$ to obtain \ref{eq:medianGomp},
+\begin{equation}
+\frac{d}{dt}\ln{m(t)} = -\beta \ln{m(t)}.
+\end{equation}
+Finally, in this exposition, all the nodes in the network are communicating instantaneously.
 
 \section{Conclusion}
-In conclusion, we have shown that Gompertz growth follows from infinite interactions between the susceptible and infected states, and the perceived pathogen travels at infinite speed throughout the population rejecting the possibility of a disease propagation stage through a perceived social network.
+In conclusion, we have shown that Gompertz growth follows from infinite interactions between the susceptible and infected states, and the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
 In this vein, Richard's parameter $\theta$ in his growth model paradigm can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
-We further show that the observations can be explained by a model where the biological system is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which gradual return to pre-epidemic conditions are mediated.
-The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities are present at all orders.
-From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from the quantum world \citep{molski2003coherent}.
+We further show that the observed mortality across countries can be explained by a model where the biological system is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which gradual return to pre-epidemic conditions are mediated.
+The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders.
+From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \citep{molski2003coherent}.
 
-Thus, we see that we are at the intersection of a road: In one direction we find the non-collaborative growth models with parameterized linear interaction effects, and in the other direction we see a field of microscopic entities coherently sharing information much like quantum entangled particles.
+Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles.
 The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology.
-One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of diatomic molecules.
+One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
 
+\newpage
 \section*{Supplement}
@@
 \end{table}
+\bibliographystyle{plainnat}
+\bibliography{../references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .
 
-% \section*{References}
+%%% Uncomment this section and comment out the \bibliography{references} line above to use inline references.
+% \begin{thebibliography}{1}
 
-\bibliography{../../references2}
+%   \bibitem{kour2014real}
+%   George Kour and Raid Saabne.
+%   \newblock Real-time segmentation of on-line handwritten arabic script.
+%   \newblock In {\em Frontiers in Handwriting Recognition (ICFHR), 2014 14th
+%       International Conference on}, pages 417--422. IEEE, 2014.
+
+%   \bibitem{kour2014fast}
+%   George Kour and Raid Saabne.
+%   \newblock Fast classification of handwritten on-line arabic characters.
+%   \newblock In {\em Soft Computing and Pattern Recognition (SoCPaR), 2014 6th
+%       International Conference of}, pages 312--318. IEEE, 2014.
+
+%   \bibitem{hadash2018estimate}
+%   Guy Hadash, Einat Kermany, Boaz Carmeli, Ofer Lavi, George Kour, and Alon
+%   Jacovi.
+%   \newblock Estimate and replace: A novel approach to integrating deep neural
+%   networks with existing applications.
+%   \newblock {\em arXiv preprint arXiv:1804.09028}, 2018.
+
+% \end{thebibliography}
 

### 2022-04-25 main.tex
add refs

@@
 \end{equation} 
-If we let $s_i = 1 - f_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model becomes 
+If we let $s_i = 1 - f_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
 \begin{equation}
@@
 \bibliographystyle{plainnat}
-\bibliography{../references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .
+\bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .
 

### 2022-04-27 main.tex
mod last par on sec 2 + equations

@@
 
-Thus, a more direct interpretation of the observations is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate \citep{boxenbaum2017hypotheses,neafsey1988gompertz}.
-Here, a geometric rate is an exponential rate in the log-transformed domain, which is the natural transformation for many processes in nature \citep{zhang1994log}. 
-The log-normal distribution of the abundance of $F(t)$ can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon}.
+Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}.
+Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon}.
 Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
@@
 
+\begin{subequations}
 \begin{align}
-\label{eq:compare}
-\begin{split}
-\dot{F} = \frac{d}{dt}F(t) = \beta F(t) (1-F(t)) \quad&\quad \text{Logistic}\\ 
-\frac{d}{dt}\ln{(F(t))} = -\beta \ln{(F(t)}) &\quad \text{Gompertz},
-\end{split}
+\dot{F} = \frac{d}{dt}F(t) & = \beta F(t) (1-F(t)) \quad \text{Logistic}\label{eq:compareLog}\\ 
+\frac{d}{dt}\ln{(F(t))} & = -\beta \ln{(F(t)}) \quad\quad\text{Gompertz}\label{eq:compareGom},
 \end{align}
+\end{subequations}
 where $F$ is here deterministic.
@@
 
-A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $F(t) \rightarrow f_1(t), f_2(t),$ etc, from which macroscopic growth would be obtained by taking the average over the individual units,
+A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $F(t) \rightarrow f_1(t), f_2(t), ..., f_N(t)$, where the lower case $f_i$ emphasize the microscopic quality of the variables, and is here interpreted as probability of infection.
+From these microscopic units, macroscopic growth would be obtained by taking their average,
 \begin{equation}
@@
 \end{equation}
-where the lower case $f_i$ emphasize the microscopic quality of the variables, and is here interpreted as probability of infection.
 
-One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compare})
+One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compareLog})
 \begin{equation}

### 2022-04-27 main.tex
log-normal distribution implies CLT

@@
 Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon}.
-Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
+Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$ and justifying a central limit theorem in this log-domain.
 
@@
 
-In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus suggests correlated mortality growth, viz.
+In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus implies correlated mortality growth by application of the central limit theorem in the log-domain, viz.
 \begin{equation}
@@
 \end{equation}
-where the rightmost equality is due to the central limit theorem where $z_i$ is a random variable as in \ref{eq:microscopic} while $f_i$ is here deterministic. 
+where $z_i$ is a random variable as in \ref{eq:microscopic} while $f_i$ is here deterministic. 
 Under this model, correlation between entities is present at all orders in the original domain\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - F(t))} = -b$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - F(t) = \exp{(-bt)}$ \citep{shklovskii2005simple}.
@@
 
+\section*{Code Availability}
+The computer code required to produce the results are given in the following Github repository,
+\begin{verbatim}
+https://github.com/matzhaugen/GompertzManuscript
+\end{verbatim}
+
 \newpage
+
 \section*{Supplement}

### 2022-04-29 main.tex
remove redundant folder

@@
 \label{eq:gomp_solution}
-Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}}
+Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
 \end{equation}
-$Y_{0}=Y(t = 0)$, $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.
+where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.
 
@@
 With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity.
-Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters obtainable through linear regression, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$, an effect which is manifested as a tapering of the initial straight line in the double-log domain.
+Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters obtainable through linear regression, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$ which is a parameter.
 Thus, through linear regression estimates, the Gompertz model avoids the possibility of non-identifiability issues and unstable optima \citep{roda2020difficult}.
 Comparisons between the two models and the inevitable concavity of the SIRD model can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig.
-\ref{fig:Gompertz_vs_SIRD_lnln_ymax} and Supplemental Information).
+\ref{fig:Gompertz_vs_SIRD_lnln_infty} and Supplemental Information).
 
@@
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_ymax}{Cumulative number of diseased, transformed by $g(Y)=\ln{\ln{({Y_{max} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD with a Gompertz model.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text after normalizing, although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}).
-The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD with a Gompertz model.
+Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
+Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}), it is here obtained using non-linear least squares to put both models on equal footing.
+The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
 Each plot is annotated with an inferred transmission rate in the SIRD model, assuming bi-normally distributed parameter estimates.
@@
 \end{subequations}
-where $F$ is here deterministic.
+where $F$ is in this section treated as deterministic.
 
@@
 
-In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus implies correlated mortality growth by application of the central limit theorem in the log-domain, viz.
+In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus implies correlated mortality growth by application of the central limit theorem in the log-domain and the continuous mapping theorem, viz.
 \begin{equation}
-F(t) = \exp{\left[\frac{1}{N}\sum_i^N z_i(t)\right]} = \left[\prod_i^N f_i(t)\right]^{1/N},
+\exp{\left[\frac{1}{N}\sum_i^N z_i(t)\right]} \xrightarrow{a.s.} \left[\prod_i^N f_i(t)\right]^{1/N} = F(t),
 \end{equation}
@@
 \centering
-\caption{Estimated values for the SIRD model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_ymax} using notation from \ref{eq:SIRD_D}.
+\caption{Estimated values for the SIRD model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
 Note that due to normalization values will be different from \citet{carletti2020covid}.}
@@
 \centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_ymax} using notation from \ref{eq:GompertzODE}.}
+\caption{Estimated values for the Gompertz model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.}
 \begin{tabular}{llll}

### 2022-04-29 main.tex
use unnormalized values in tables

@@
 \centering
-\caption{Estimated values for the SIRD model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
-Note that due to normalization values will be different from \citet{carletti2020covid}.}
+\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
+Standard deviation is given in the paranthesis corresponding to the last significant digit of the reported number.}
 \begin{tabular}{llll}
 \toprule
-{} &           $\eta\times 10$ &      $\xi$ &            $\kappa\times 10^{2}$ \\
+{} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
 \midrule
-Belgium        &    1.0(1) &  2.9(3) &   9.3(1) \\
-Portugal       &   0.55(4) &  4.2(4) &   4.5(4) \\
-Austria        &    1.3(3) &  2.3(4) &    11(2) \\
-Switzerland    &   1.24(1) &  2.4(2) &  11.0(7) \\
-Sweden         &    0.5(2) &    3(1) &     3(2) \\
-Italy          &   0.43(1) &  5.6(2) &   3.7(1) \\
-France         &   0.88(7) &  3.4(3) &   8.1(6) \\
-Spain          &  0.525(1) &  6.2(2) &   4.8(1) \\
-Germany        &   0.78(8) &  3.0(3) &   6.5(7) \\
-Denmark        &   0.50(4) &  5.2(6) &   4.2(4) \\
-United Kingdom &   0.54(2) &  4.7(2) &   4.3(2) \\
-Netherlands    &   0.51(3) &  5.0(4) &   4.3(3) \\
-Turkey         &   0.84(9) &  2.8(3) &   6.9(8) \\
-Iran           &  0.322(7) &  7.2(3) &  2.51(9) \\
-US             &  0.339(8) &  7.0(3) &   1.8(1) \\
-China          &    3.8(9) &  1.2(2) &    26(4) \\
+Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
+Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
+Austria        &     8(1) &    36(5) &    11(1) \\
+Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
+Sweden         &    19(5) &     8(2) &     3(1) \\
+Italy          &   137(3) &  1.77(5) &  3.71(1) \\
+France         &  240(10) &  1.24(7) &   8.1(4) \\
+Spain          &   145(2) &  2.27(5) &  4.75(8) \\
+Germany        &    61(4) &   3.9(3) &   6.5(5) \\
+Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
+United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
+Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
+Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
+Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
+US             &   312(5) &  0.76(2) &  1.83(7) \\
+China          &  130(30) &   3.5(5) &    26(4) \\
 \bottomrule
@@
 \centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.}
+\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.}
 \begin{tabular}{llll}
@@
 \midrule
-Belgium        &   2.74(1) &   8.1(2) &  3.6(3) \\
-Portugal       &  2.242(6) &  6.23(9) &  3.1(1) \\
-Austria        &   2.74(1) &   8.1(2) &  3.6(2) \\
-Switzerland    &  2.740(7) &   8.1(1) &  3.6(2) \\
-Sweden         &   1.73(3) &   4.3(3) &  2.6(4) \\
-Italy          &  2.107(6) &  5.88(9) &  3.0(1) \\
-France         &  2.752(1) &   8.2(2) &  3.6(2) \\
-Spain          &  2.556(8) &   7.5(1) &  3.4(2) \\
-Germany        &  2.335(7) &   6.5(1) &  3.2(1) \\
-Denmark        &   2.32(1) &   6.6(2) &  3.2(3) \\
-United Kingdom &  2.333(6) &  6.54(9) &  3.2(1) \\
-Netherlands    &  2.311(7) &   6.5(1) &  3.2(1) \\
-Turkey         &  2.306(7) &   6.4(1) &  3.2(1) \\
-Iran           &  1.862(7) &  5.04(9) &  2.7(1) \\
-US             &  2.014(8) &   5.3(1) &  2.9(1) \\
-China          &  3.155(8) &   9.4(2) &  3.6(2) \\
+Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
+Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
+Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
+Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
+Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
+Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
+France         &  10.2(1) &   8.2(1) &  11.0(8) \\
+Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
+Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
+Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
+United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
+Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
+Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
+Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
+US             &  7.53(7) &  5.32(7) &   8.4(5) \\
+China          &   9.8(1) &   9.4(2) &    10(1) \\
 \bottomrule

### 2022-05-02 main.tex
change working in 2 vs 3 parameters

@@
 obtained with a second order expansion of the exponential term in \ref{eq:SIRD_D} and with $b=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
-The SIR model exhibits similar properties and its discrepancy with initial observations has been noted by others \citep{vattay2020forecasting}.
+The SIR and SEIR model exhibits similar properties and its discrepancy with initial observations has been noted by others \citep{vattay2020forecasting}.
 
-With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity.
-Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters obtainable through linear regression, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$ which is a parameter.
-Thus, through linear regression estimates, the Gompertz model avoids the possibility of non-identifiability issues and unstable optima \citep{roda2020difficult}.
-Comparisons between the two models and the inevitable concavity of the SIRD model can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig.
+With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig.
 \ref{fig:Gompertz_vs_SIRD_lnln_infty} and Supplemental Information).
+Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in (\ref{eq:GompertzODE}), it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
+\begin{equation}
+\label{eq:GompertzODE2param}
+\frac{d}{dt}\ln{Y(t)} = -\beta(\ln{Y(t)} - \ln {Y_{\infty}}),
+\end{equation}
+following e.g. \citet{chakrabarti1996non}.
+Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \citep{roda2020difficult}.
 
@@
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD with a Gompertz model.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model with a Gompertz model for a variety of countries in the period Jan-May 2020.
 Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}), it is here obtained using non-linear least squares to put both models on equal footing.
+Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GompertzODE2param}), it is here obtained using non-linear least squares to put both models on equal footing.
 The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
@@
 Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}.
-Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon}.
+Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
 Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$ and justifying a central limit theorem in this log-domain.

### 2022-05-03 main.tex
change notation to use deterministic vars in last section

@@
 \section{An alternative theory for observed mortality}
+\label{seq:alt}
 An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.
@@
 \end{equation}
-Then use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $m(t)$, 
+Then use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $M(t)$, 
 \begin{equation}
 \label{eq:medianGomp}
-d\ln{m(t)} = -\beta \ln{m(t)} dt,
+d\ln{M(t)} = -\beta \ln{M(t)} dt,
 \end{equation}
-which corresponds with the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}) with $Y(t) = Y_\infty m(t)$. 
+which corresponds with the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}) with $Y(t) = Y_\infty M(t)$. 
 Comparing the stochastic stressor $\sigma$ in \ref{eq:microscopicOriginal} with the deterministic growth equation in (\ref{eq:GompertzODE}) gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
@@
 \begin{align}
-\dot{F} = \frac{d}{dt}F(t) & = \beta F(t) (1-F(t)) \quad \text{Logistic}\label{eq:compareLog}\\ 
-\frac{d}{dt}\ln{(F(t))} & = -\beta \ln{(F(t)}) \quad\quad\text{Gompertz}\label{eq:compareGom},
+\dot{M} = \frac{d}{dt}M(t) & = \beta M(t) (1-M(t)) \quad \text{Logistic}\label{eq:compareLog}\\ 
+\frac{d}{dt}\ln{(M(t))} & = -\beta \ln{(M(t)}) \quad\quad\text{Gompertz}\label{eq:compareGom},
 \end{align}
 \end{subequations}
-where $F$ is in this section treated as deterministic.
+where $M(t)$ is deterministic.
 
@@
 
-A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $F(t) \rightarrow f_1(t), f_2(t), ..., f_N(t)$, where the lower case $f_i$ emphasize the microscopic quality of the variables, and is here interpreted as probability of infection.
-From these microscopic units, macroscopic growth would be obtained by taking their average,
+A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $M(t) \rightarrow x_1(t), x_2(t), ..., x_N(t)$, where the lower case $x_i$ emphasize the microscopic quality of the variables, and is here interpreted as probability of infection.
+From these microscopic units, macroscopic growth could be obtained by taking their arithmetic average,
 \begin{equation}
-F(t) = \frac{1}{N}\sum_i^N f_i(t),
+M(t) = \frac{1}{N}\sum_i^N x_i(t),
 \end{equation}
@@
 \label{eq:networkSIR}
-\frac{d f_i}{dt} = \beta (1-f_i)\sum_j{a_{ij}}f_j \quad \forall i,
+\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j \quad \forall i,
 \end{equation}
-using the shorthanded $f_i=f_i(t)$ and with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
+using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
 Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below.
@@
 \begin{equation}
-\frac{\partial^2 \dot{F}}{\partial f_i \partial f_j} = -\frac{\beta}{N} a_{i,j}.
+\frac{\partial^2 \dot{M}}{\partial x_i \partial x_j} = -\frac{\beta}{N} a_{i,j}.
 \end{equation}
@@
 
-In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus implies correlated mortality growth by application of the central limit theorem in the log-domain and the continuous mapping theorem, viz.
+In contrast, as discussed in Section \ref{seq:alt}, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
+This implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
+
 \begin{equation}
-\exp{\left[\frac{1}{N}\sum_i^N z_i(t)\right]} \xrightarrow{a.s.} \left[\prod_i^N f_i(t)\right]^{1/N} = F(t),
+M(t) = \exp \left[\frac{1}{N}\sum_i \ln x_i(t)\right] = \left[\prod_i x_i (t)\right]^{\frac{1}{N}}
 \end{equation}
-where $z_i$ is a random variable as in \ref{eq:microscopic} while $f_i$ is here deterministic. 
-Under this model, correlation between entities is present at all orders in the original domain\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - F(t))} = -b$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - F(t) = \exp{(-bt)}$ \citep{shklovskii2005simple}.
+
+Under this model, correlation between entities is present at all orders in the original domain and all the nodes in the network communicate instantaneously\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - M(t))} = -\beta$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - M(t) = \exp{(-\beta t)}$ \citep{shklovskii2005simple}.
 In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.
 
-Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities \citep{zhang1994log}.
+Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
 We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
@@
 \subsection{Unification through a modified SIR model}
-However, it is possible to reconcile the SIR model with the Gompertz curve from yet another argument. 
+However, it is possible to reconcile the SIR model with the Gompertz curve. 
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
@@
 \label{eq:reverseNetworkSIR}
-\frac{d f_i}{dt} = \beta f_i\sum_j{a_{ij}}(1-f_j) \quad \forall i.
+\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(1-x_j) \quad \forall i.
 \end{equation} 
@@
 \end{equation} 
-If we let $s_i = 1 - f_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
+If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
 \begin{equation}
-\frac{d f_i}{dt} = \beta f_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha f_i\quad \forall i,
+\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
 \end{equation}
 
-Now use the Taylor series $x+x^2/2+... = -\ln(1-x)$, viz.
+Now use the Taylor series $s+s^2/2+... = -\ln(1-s)$, viz.
 
 \begin{equation}
-\frac{d f_i}{dt} = \beta f_i\sum_j{a_{ij}}\ln{(1 - s_{j})} - \alpha = \beta f_i\sum_j{a_{ij}}\ln{f_{j}} - \alpha f_i \quad \forall i,
+\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = \beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
 \end{equation}
@@
 \begin{equation}
-\frac{d}{{dt}} \ln\left[\prod_i{f_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i{f_i}\right]^{\frac{1}{N}} - \alpha.
+\frac{d}{{dt}} \ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} - \alpha.
 \end{equation}
@@
 The interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
-One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $f_i$ to obtain \ref{eq:medianGomp},
+One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain \ref{eq:medianGomp},
 \begin{equation}
-\frac{d}{dt}\ln{m(t)} = -\beta \ln{m(t)}.
+\frac{d}{dt}\ln{M(t)} = -\beta \ln{M(t)}.
 \end{equation}
-Finally, in this exposition, all the nodes in the network are communicating instantaneously.
 

### 2022-05-03 main.tex
add brackets around equations

@@
 \end{equation}
-obtained with a second order expansion of the exponential term in \ref{eq:SIRD_D} and with $b=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
+obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $b=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
 The SIR and SEIR model exhibits similar properties and its discrepancy with initial observations has been noted by others \citep{vattay2020forecasting}.
@@
 \label{eq:GompertzODE2param}
-\frac{d}{dt}\ln{Y(t)} = -\beta(\ln{Y(t)} - \ln {Y_{\infty}}),
+\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right],
 \end{equation}
@@
 \begin{equation}
-d\langle\ln{F(t)}\rangle = -\beta \langle\ln{F(t)}\rangle dt.
+d\langle\ln{F(t)}\rangle = -\beta \langle\ln{F(t)}\rangle dt,
 \end{equation}
+where the bra-ket notation signifies the averaging operation.
 Then use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $M(t)$, 
@@
 Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}.
-Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
+Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in (\ref{eq:microscopicOriginal}) \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
 Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$ and justifying a central limit theorem in this log-domain.
@@
 The interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
-One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain \ref{eq:medianGomp},
+One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain (\ref{eq:medianGomp}),
 \begin{equation}

### 2022-05-07 main.tex
remove chakrabarti due to error in equations
final reading 1

@@
 
-However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
+However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825,bajzer1997mathematical} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
 \begin{equation}
@@
 
-One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$.
+One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}.
 How would $\phi(t)$ have to behave?
-Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
+Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
 \begin{align}
@@
 A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
-both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the total population, $N \gg Y \ge I$, which seems well founded at the national level,
+both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the total population, $N \gg Y \ge I$, which seems well founded at the national level, viz.
 \begin{align}
@@
  Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
-And since $I=I(Y)$, we combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
+And since $I$ is a function of $Y$, we can combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
 \begin{subequations}
@@
 
+\subsection{The SIR model family is almost Gompertzian}
 Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain from the communicable disease models growth which is almost Gompertzian, i.e.
@@
 To show this, we follow \citet{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of equations (\ref{eq:SIR}), with a growth equation of the diseased group at some rate $\delta$ relative to the current infected group,
+This requires an addition to our original set of equations (\ref{eq:SIR}), with a extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 
-Note here that at small values of D follow logistic growth,
+Note here that small values of D follow logistic growth, viz.
 \begin{equation}
@@
 obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $b=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
-The SIR and SEIR model exhibits similar properties and its discrepancy with initial observations has been noted by others \citep{vattay2020forecasting}.
+The SIR and SEIR model exhibits similar properties and its discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
 
@@
 \label{eq:GompertzODE2param}
-\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right],
+\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right].
 \end{equation}
-following e.g. \citet{chakrabarti1996non}.
 Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \citep{roda2020difficult}.
 
-Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond}.
-However, this model does not consider network effects explicitly.
+Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly.
 Rather, all entities are connected and communicating instantaneously as shown by \citet{mombach2002mean}.
@@
 The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
-Each plot is annotated with an inferred transmission rate in the SIRD model, assuming bi-normally distributed parameter estimates.
+Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates.
 Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
@@
 
-Inspired by  \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term at the along with the countering force of the immune system. 
-As with all multiplicative processes, it is convenient to log-transformed mortality with dimensionless variables $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
+Inspired by \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
+As with all multiplicative processes, it is convenient to log-transform mortality and work in a dimensionless space such that $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
 \begin{equation}
@@
 where $dW(t)$ is a delta-correlated Wiener process with zero mean, making $Z(t)$ a stochastic process.
+The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response.
+
 This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}.
@@
 
-The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response.
-
 To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}) and transform back to the original domain, 
@@
 where the bra-ket notation signifies the averaging operation.
-Then use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $M(t)$, 
+Then we use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $M(t)$, 
 \begin{equation}
@@
 \begin{equation}
-\sigma = 2\gamma_{\infty}\ln{(Y_{\infty})}.
+\sigma = 2\gamma_{\infty}\ln{Y_{\infty}}.
 \end{equation}
@@
 Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in (\ref{eq:microscopicOriginal}) \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
-Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$ and justifying a central limit theorem in this log-domain.
+Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
+This implies that the log-domain is the domain in which the central limit theorem applies.
 
 \section{Unifying the SIR model and the Gompertz model}
-The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
+This remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
 
@@
 
-A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $M(t) \rightarrow x_1(t), x_2(t), ..., x_N(t)$, where the lower case $x_i$ emphasize the microscopic quality of the variables, and is here interpreted as probability of infection.
+A microscopic solution could be modeled by splitting the system into $N$ microscopic deterministic units, $M(t) \rightarrow x_1(t), x_2(t), ..., x_N(t)$, where the lower case $x_i$ emphasizes the microscopic quality of the variables, and is here interpreted as probability of infection.
 From these microscopic units, macroscopic growth could be obtained by taking their arithmetic average,
 \begin{equation}
-M(t) = \frac{1}{N}\sum_i^N x_i(t),
+M(t) = \frac{1}{N}\sum_i^N x_i(t).
 \end{equation}
+Naturally, this simple arithmetic average treats each microscopic unit as independent variables contributing to the macroscopic observable.
 
@@
 \end{equation}
-using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $b$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
+using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
 Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below.
@@
 In contrast, as discussed in Section \ref{seq:alt}, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
-This implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
+As mentioned earlier, this implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
 
@@
 In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields.
+The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is quite noteworthy.
+It suggests that both systems share commonalities and means of communication.
 
@@
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
-Then, we reverse the causality where the population of infected are now dependent of the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
+Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
 \begin{equation}
@@
 \begin{equation}
-\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = \beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
+\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
 \end{equation}
@@
 \end{equation}
-If $\alpha\rightarrow 0$, then this equation will exhibit Gompertz growth in the geometric ensemble average of microscopic units.
-The interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
+If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
+An interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
 One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain (\ref{eq:medianGomp}),
@@
 \section{Conclusion}
-In conclusion, we have shown that Gompertz growth follows from infinite interactions between the susceptible and infected states, and the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter $\theta$ in his growth model paradigm can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
+In this vein, Richard's parameter, $\theta$, in his growth model paradigm can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
@@
 Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles.
-The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology.
+The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology \citep{lambert2013quantum}.
 One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
@@
 
+\subsection{A note on estimation}
+
+% Estimating the parameters using real data requires some consideration around the variance of the growth variable as a function of time.
+This variance will in turn inform the estimation process by considering a weighted minimization of the residuals of the observations regressed onto the model.
+In the stochastic Gompertz process postulated in Section \ref{seq:alt}, the process under a log-transformation is the Ornstein-Uhlenbeck process which has a closed from expression for its transition probability, and thus its variance as a function of time.
+This variance can then be used in a weighted least squares estimation technique or any other weighted technique.
+In the original domain there is also a closed form expression available but it is somewhat more convoluted and given that observations in this domain are visually quite homoscedastic with equal variance across time (Fig \ref{fig:Gompertz_vs_SIRD} and Tables \ref{tab:gomp} and \ref{tab:sird}), this is expression and consideration is omitted.
+But for the log-domain, observations exibit changing variance with time.
+
+% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\citep{risken1996fokker},
+% \begin{equation}
+% P(z, t|z', t') = \sqrt{\frac{\beta}{2\pi \sigma (1 - e^{-2\beta (t-t')})}}\exp{\left[ -\frac{\beta (z - e^{-\beta(t-t')}z')^2}{2\sigma(1-e^{-2 \beta(t-t')})} \right]}.
+% \end{equation}
+
+% If we consider $t$ the time at which the stationary distribution is reached and that this stationary distribution is given as a delta function, variance of the growth will be given as
+% \begin{equation}
+% \text{Var[Z(t)|t']} = \sigma(1-e^{-2 \beta(t-t')}).
+% \end{equation}
+
+% Because weights on observations in a regression are invariant to multiplication we can let $\sigma=1$.
+The slope coefficient would have to be estimated either by a standard least squares or through the 3-parameter model in (\ref{eq:GompertzODE}).
+
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta (4) algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
+\ref{fig:Gompertz_vs_SIRD}).
+Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
+
+The transmission rate is modeled as the product of $\eta \xi$ is (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
+\begin{equation}
+\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
+\end{equation}
+
 \insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
+Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.}
+
+% \insertPdfFig{Gompertz_vs_SIRD_gompfit_log_domain}{Log transformed cumulative diseased plotted against its time derivative after $Y(t) / Y_{max}>0.005$, with a straight line fit regressing onto the $50^{th}$ quantile.}
 
 \begin{table}[H]
+\label{tab:sird}
 \centering
@@
 \begin{table}[H]
+\label{tab:gomp}
 \centering

### 2022-05-09 elsarticle/elsarticle.tex
update elsarticle and main + figures white background

--- 2022-05-07 main.tex
@@
-\documentclass{article}
-% \documentclass[review]{elsarticle}
+\documentclass[review]{elsarticle}
 
-\usepackage{arxiv}
-\usepackage{booktabs}
-\usepackage{siunitx}
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
+\usepackage{lineno,hyperref}
+\usepackage{amsmath}
 \usepackage{amsfonts}       % blackboard math symbols
 \usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{lipsum}   % Can be removed after putting your text content
+\usepackage{float}
+\usepackage{booktabs}       % professional-quality tables
 \usepackage{graphicx}
-\usepackage[square,numbers]{natbib}
-\usepackage{doi}
-\usepackage{amsmath}
-\usepackage{float}
+\modulolinenumbers[5]
+
+\journal{Chaos, Solitons \& Fractals}
+
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Elsevier bibliography styles
+%%%%%%%%%%%%%%%%%%%%%%%
+%% To change the style, put a % in front of the second line of the current style and
+%% remove the % from the second line of the style you would like to use.
+%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Numbered
+% \bibliographystyle{model1-num-names}
+
+%% Numbered without titles
+%\bibliographystyle{model1a-num-names}
+
+%% Harvard
+%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
+
+%% Vancouver numbered
+%\usepackage{numcompress}\bibliographystyle{model3-num-names}
+
+%% Vancouver name/year
+%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
+
+%% APA style
+% \bibliographystyle{model5-names}\biboptions{authoryear}
+
+%% AMA style
+\usepackage{numcompress}\bibliographystyle{model6-num-names}
+
+%% `Elsevier LaTeX' style
+\bibliographystyle{elsarticle-num}
+%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \centering
-  \includegraphics[width=8.4cm]{#1.pdf}
+  \includegraphics[width=12.4cm]{#1.pdf}
   \caption{#2}
@@
 }
+\graphicspath{{../figures/}}
+\begin{document}
 
-% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
+\begin{frontmatter}
+
 \title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
-\graphicspath{{figures/}}
-%\date{September 9, 1985} % Here you can change the date presented in the paper title
-%\date{}          % Or removing it
 
-\author{Matz A. Haugen, \\
-Independent Reseacher, Oslo\\
-\texttt{matzhaugen@gmail.com} \\
-  \And
-  Dorothea Gilbert \\
-  University of Oslo\\
-  %% \AND
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-  %% \And
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-  %% \And
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-}
+%% Group authors per affiliation:
+% \author{Matz Andreas Haugen\fnref{myfootnote}}
+% \address{Enerhauggata 1, 0651 Oslo, Norway}
+% \fntext[myfootnote]{Since 1880.}
 
-% Uncomment to remove the date
-%\date{}
+%% or include affiliations in footnotes:
+\author[mymainaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
+\cortext[mycorrespondingauthor]{Corresponding author}
+% \ead[url]{www.matzhaugen.com}
+\ead{matzhaugen@gmail.com}
 
-% Uncomment to override  the `A preprint' in the header
-%\renewcommand{\headeright}{Technical Report}
-%\renewcommand{\undertitle}{Technical Report}
-\renewcommand{\shorttitle}{On the nature of the Gompertz}
+\author[mysecondaryaddress]{Dorothea Gilbert}
 
-%%% Add PDF metadata to help others organize their library
-%%% Once the PDF is generated, you can check the metadata with
-%%% $ pdfinfo template.pdf
-\hypersetup{
-pdftitle={A template for the arxiv style},
-pdfsubject={q-bio.NC, q-bio.QM},
-pdfauthor={Matz A. Haugen},
-pdfkeywords={},
-}
-
-\begin{document}
-\maketitle
+\address[mymainaddress]{Enerhauggata 1, 0651 Oslo, Norway}
+\address[mysecondaryaddress]{University of Oslo, Oslo, Norway}
 
@@
 
-% keywords can be removed
-\keywords{Gompertz \and Coherence \and Covid \and Coronavirus \and Network Analysis \and Stochastic}
+\begin{keyword}
+Gompertz \sep Coherence \sep Covid \sep Coronavirus \sep Network Analysis \sep Stochastic
+\end{keyword}
+
+\end{frontmatter}
+
+\linenumbers
 
 \section{Introduction}
+
 Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
@@
 
-\section*{Code Availability}
+\section*{Methods}
+
+\subsection{Estimating model parameters}
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
+\ref{fig:Gompertz_vs_SIRD}).
+Fitting is done with Python-Scipy with specified Jacobians and a 6-day windowed average of observed deaths.
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
+
+The transmission rate is modeled as the product of $\eta \xi$ in (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
+\begin{equation}
+\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
+\end{equation}
+
+\subsection{Code Availability}
 The computer code required to produce the results are given in the following Github repository,
@@
 
-\newpage
-
 \section*{Supplement}
 
-\subsection{A note on estimation}
-
-% Estimating the parameters using real data requires some consideration around the variance of the growth variable as a function of time.
-This variance will in turn inform the estimation process by considering a weighted minimization of the residuals of the observations regressed onto the model.
-In the stochastic Gompertz process postulated in Section \ref{seq:alt}, the process under a log-transformation is the Ornstein-Uhlenbeck process which has a closed from expression for its transition probability, and thus its variance as a function of time.
-This variance can then be used in a weighted least squares estimation technique or any other weighted technique.
-In the original domain there is also a closed form expression available but it is somewhat more convoluted and given that observations in this domain are visually quite homoscedastic with equal variance across time (Fig \ref{fig:Gompertz_vs_SIRD} and Tables \ref{tab:gomp} and \ref{tab:sird}), this is expression and consideration is omitted.
-But for the log-domain, observations exibit changing variance with time.
-
-% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\citep{risken1996fokker},
-% \begin{equation}
-% P(z, t|z', t') = \sqrt{\frac{\beta}{2\pi \sigma (1 - e^{-2\beta (t-t')})}}\exp{\left[ -\frac{\beta (z - e^{-\beta(t-t')}z')^2}{2\sigma(1-e^{-2 \beta(t-t')})} \right]}.
-% \end{equation}
-
-% If we consider $t$ the time at which the stationary distribution is reached and that this stationary distribution is given as a delta function, variance of the growth will be given as
-% \begin{equation}
-% \text{Var[Z(t)|t']} = \sigma(1-e^{-2 \beta(t-t')}).
-% \end{equation}
-
-% Because weights on observations in a regression are invariant to multiplication we can let $\sigma=1$.
-The slope coefficient would have to be estimated either by a standard least squares or through the 3-parameter model in (\ref{eq:GompertzODE}).
-
-The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta (4) algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
-\ref{fig:Gompertz_vs_SIRD}).
+\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
+Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
 Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
-
-The transmission rate is modeled as the product of $\eta \xi$ is (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
-\begin{equation}
-\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
-\end{equation}
-
-\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.}
-
-% \insertPdfFig{Gompertz_vs_SIRD_gompfit_log_domain}{Log transformed cumulative diseased plotted against its time derivative after $Y(t) / Y_{max}>0.005$, with a straight line fit regressing onto the $50^{th}$ quantile.}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
 
@@
 \centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.}
+\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.
+Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
 \begin{tabular}{llll}
@@
 \end{table}
-\bibliographystyle{plainnat}
-\bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .
 
-%%% Uncomment this section and comment out the \bibliography{references} line above to use inline references.
-% \begin{thebibliography}{1}
+% \section*{References}
 
-%   \bibitem{kour2014real}
-%   George Kour and Raid Saabne.
-%   \newblock Real-time segmentation of on-line handwritten arabic script.
-%   \newblock In {\em Frontiers in Handwriting Recognition (ICFHR), 2014 14th
-%       International Conference on}, pages 417--422. IEEE, 2014.
-
-%   \bibitem{kour2014fast}
-%   George Kour and Raid Saabne.
-%   \newblock Fast classification of handwritten on-line arabic characters.
-%   \newblock In {\em Soft Computing and Pattern Recognition (SoCPaR), 2014 6th
-%       International Conference of}, pages 312--318. IEEE, 2014.
-
-%   \bibitem{hadash2018estimate}
-%   Guy Hadash, Einat Kermany, Boaz Carmeli, Ofer Lavi, George Kour, and Alon
-%   Jacovi.
-%   \newblock Estimate and replace: A novel approach to integrating deep neural
-%   networks with existing applications.
-%   \newblock {\em arXiv preprint arXiv:1804.09028}, 2018.
-
-% \end{thebibliography}
+\bibliography{../references2}
 

### 2022-05-09 main.tex
update elsarticle and main + figures white background

@@
 
-\section*{Code Availability}
+\newpage
+
+\section*{Methods}
+
+\subsection{Estimating model parameters}
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
+\ref{fig:Gompertz_vs_SIRD}).
+Fitting is done with Python-Scipy with specified Jacobians and a 6-day windowed average of observed deaths.
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
+
+The transmission rate is modeled as the product of $\eta \xi$ in (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
+\begin{equation}
+\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
+\end{equation}
+
+\subsection{Code Availability}
 The computer code required to produce the results are given in the following Github repository,
@@
 \end{verbatim}
-
-\newpage
 
@@
 
-The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta (4) algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
-\ref{fig:Gompertz_vs_SIRD}).
+\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
+Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
 Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
-
-The transmission rate is modeled as the product of $\eta \xi$ is (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
-\begin{equation}
-\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
-\end{equation}
-
-\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
 
@@
 \caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
-Standard deviation is given in the paranthesis corresponding to the last significant digit of the reported number.}
+Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
 \begin{tabular}{llll}
@@
 \end{table}
-\bibliographystyle{plainnat}
+\bibliographystyle{abbrvnat}
 \bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .

### 2022-05-15 main.tex
2nd readthrough + update small changes

@@
 A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
-both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the total population, $N \gg Y \ge I$, which seems well founded at the national level, viz.
+both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
 \begin{align}
@@
 \end{equation}
-where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonymous to the initial susceptible pool of people.
+where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is again synonymous to the initial susceptible pool of people.
 This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$.
@@
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model with a Gompertz model for a variety of countries in the period Jan-May 2020.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
 Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
@@
 \section{Unifying the SIR model and the Gompertz model}
-This remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
+The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
 
@@
 using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
-Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below.
+Notice that there is an implied causality going from the infected to susceptible, which will become relevant below.
 Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
@@
 In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields.
-The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is quite noteworthy.
-It suggests that both systems share commonalities and means of communication.
+The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.
 
@@
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter, $\theta$, in his growth model paradigm can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
@@
 \ref{fig:Gompertz_vs_SIRD}).
-Fitting is done with Python-Scipy with specified Jacobians and a 6-day windowed average of observed deaths.
+Fitting is done using Python-Scipy's non-linear least squares algorithm \verb|curve_fit| with specified Jacobians and a 6-day windowed average of observed deaths.
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
@@
 \begin{equation}
-\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
+\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\text{E}[\eta]^2 + \text{Var}[\eta] \text{E}[\xi]^2,
 \end{equation}
+where estimates are obtained by substituting the model parameters with those estimated through the technique outlined above.
 
 \subsection{Code Availability}
-The computer code required to produce the results are given in the following Github repository,
-\begin{verbatim}
-https://github.com/matzhaugen/GompertzManuscript
-\end{verbatim}
+
+Scripts used to produce the plots are downloadable in the form of a Jupyter notebook using Python from 
+\newline
+\verb|https://emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb|
+% The computer code required to produce the results are given in the following Github repository,
+% \begin{verbatim}
+% https://github.com/matzhaugen/GompertzManuscript
+% \end{verbatim}
 
 \section*{Supplement}
-
-\subsection{A note on estimation}
 
@@
 \centering
-\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
+\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq.
+(\ref{eq:SIRD_D}).
 Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
@@
 \centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.}
+\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq. (\ref{eq:GompertzODE}).}
 \begin{tabular}{llll}

### 2022-05-18 letterToEditor/letter.tex
add letter to editor

\documentclass[11pt,a4paper,roman]{moderncv}      
\usepackage[english]{babel}

\moderncvstyle{classic}                            
\moderncvcolor{black}

% character encoding
\usepackage[utf8]{inputenc}

% adjust the page margins
\usepackage[scale=0.70]{geometry}

% personal data
\name{Matz Andreas Haugen}{}
\email{matzhaugen@gmail.com}
\address{Enerhauggata 1, 0651, Oslo, Norway}

\begin{document}

\recipient{To}{Editor, \\
               Nature,\\ 
               }
\date{\today}
\opening{\textbf{Initial submission of manuscript titled: ``Unifying Gompertzian growth with the communicable disease spreading paradigm''}}
\closing{Sincerely yours, \vspace{-1em}}

\enclosure[Enclosed ]{
              \\ List of referees
%                  \hspace{0.0em} 5. GATE (2016) \hspace{2em} 9. Identity-Proof (Voter-Id)\\
%                  2. Diploma (MarkSheet/Certificate)
%                  \hspace{1.8em} 6. SLET (2021) \hspace{1.9em} 10. Resume\\        
%                  3. B.Tech (MarkSheet/Certificate)
%                  \hspace{2.3em} 7. UGC-NET (2022) \hspace{-0.1em} \\
%                  4. M.Tech (MarkSheet/Certificate)
%                  \hspace{2.1em} 8. Caste Certificate \\
                 }
\makelettertitle

Respected Editor,
\\
%references such as what and how you got this information
\vspace{1em}
We hereby submit a manuscript where we seek to connect the communicable disease paradigm with the recently observed mortality patterns of the initial Coronavirus pandemic in March 2021 that have been shown by many to exhibit Gompertzian growth. 
In the manuscript, we show that this type of growth pattern is incompatible with a traditional communicable disease spreading models, i.e. the SIR (Susceptible-Infected-Recovered) model family of Kermack and McKendrick. Instead, the observed patterns can cleanly be explained by a simpler model without the need for a disease propagating stage, but rather through a ubiquitous stressor which elicits a mutual stress response, amounting to a 2-parameter model. 
The mathematical thesis is based on a simple Stochastic Differential Equation where the stress response is a random process, interpretable both at the macroscopic or the microscopic level. 
We also show a remarkable connection between coherent behavior previously relegated to microscopic quantum domains, now exhibited in national mortality patterns.

Our results has the potential to shift the entire field of pathogenic disease modeling away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role. 
The arguments in the paper are based purely on mathematical and physical grounds as we do not wish to cover too large a scope, but clearly this paper fits into a larger discussion on the communicable disease paradigm which we feel is missing in the current scientific debate.

If you do decide to review this paper you may find the enclosed list of possible referees helpful. 
Some of them are listed in the references of the manuscript.

Thank you for your consideration.

%narrate why you want to apply

\makeletterclosing
\newpage
\textbf{List of possible referees}
\begin{enumerate}
\item N. El Karoui, mathematics, stochastic models
\item M. Molski, physics, Gompertz and quantum systems
\item W. Whitt, mathematics, stochastic models
\item N. C. Petroni, E. De Lauro, physics, Gompertz
\item E. B. Postnikov, physics, SIR models
\item G. Vattay, physics, SIR models
\item M. Levitt, computational biology, Gompertz models
\item E. Estrada, mathematics, graph theory, SIR models
\item T. Carletti, D. Fanelli, and F. Piazza, physics, SIR models
\item Z. Bajzer, biology, has a very good review paper on Gompertz modeling 
\item J. C. Mombach, physics, growth models
\item B. Shklovskii, physics, theory of Gompertz
\item D. Smilkov, C. or L. Kocarev, physics and mathematics, SIR models
\item C. G. Antonopoulos, physics, chaos and SIR models
\item K. Rypdal and M. Rypdal, mathematics, SIR models

\end{enumerate}

\end{document}

### 2022-05-18 main.tex
add letter to editor

@@
 \usepackage{float}
+\usepackage{lineno}
+\linenumbers
 
@@
 
-\author{Matz A. Haugen, \\
-Independent Reseacher, Oslo\\
-\texttt{matzhaugen@gmail.com} \\
-  \And
-  Dorothea Gilbert \\
-  University of Oslo\\
-  %% \AND
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-  %% \And
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-  %% \And
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-}
+% \author{Matz A. Haugen, \\
+% Independent Reseacher, Oslo\\
+% \texttt{matzhaugen@gmail.com} \\
+%   \And
+%   Dorothea Gilbert \\
+%   University of Oslo\\
+% }
 
@@
 \frac{dS}{dt}& = -\beta \frac{IS}{N}\\
-\frac{dI}{dt}& = -\beta \frac{IS}{N} - \alpha I.\\
+\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
 \frac{dR}{dt}& = \alpha I,
@@
 
- At this limit the growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
+ At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
  The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning.
@@
 \subsection{The SIR model family is almost Gompertzian}
-Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain from the communicable disease models growth which is almost Gompertzian, i.e.
+Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain Gompertzian growth from the communicable disease models, i.e.
 a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
@@
 \begin{equation}
-\frac{dD}{dt} = b D (1 - D/K),
+\frac{dD}{dt} = c D (1 - D/K),
 \end{equation}
-obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $b=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
-The SIR and SEIR model exhibits similar properties and its discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
+obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $c=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
+The SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
 
@@
 Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates.
-Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
+Fitting is done with Python-Scipy and a 6-day moving average of deaths.
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
@@
 which corresponds with the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}) with $Y(t) = Y_\infty M(t)$. 
-Comparing the stochastic stressor $\sigma$ in \ref{eq:microscopicOriginal} with the deterministic growth equation in (\ref{eq:GompertzODE}) gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
+Comparing the stochastic stressor $\sigma$ in (\ref{eq:microscopicOriginal}) with the deterministic growth equation in (\ref{eq:GompertzODE}) gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
 We also see that by comparing (\ref{eq:rypdalODE}) and (\ref{eq:rypdalLimit}) with the stochastic counterpart in (\ref{eq:microscopicOriginal}) that the final growth level is governed by the stressor magnitude,
@@
 Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
-This implies that the log-domain is the domain in which the central limit theorem applies.
+This implies that the central limit theorem applies to the log-domain.
 
@@
 
-In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in an SIR model, but also as a linear interaction term between the two macroscopic states. 
+In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in a SIR model, but also as a linear interaction term between the two macroscopic states. 
 As mentioned earlier, this procedure is a mean field approximation with an implied average interaction between the variables. 
@@
 \end{equation}
-Naturally, this simple arithmetic average treats each microscopic unit as independent variables contributing to the macroscopic observable.
+Naturally, this simple arithmetic average treats each microscopic unit as an independent variable contributing to the macroscopic observable.
 
@@
 \end{equation}
-using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
-Notice that there is an implied causality going from the infected to susceptible, which will become relevant below.
+using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth or infection rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
+Notice that there is an implied causality from the infected to the susceptible, which will become relevant below.
 Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
@@
 In contrast, as discussed in Section \ref{seq:alt}, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
-As mentioned earlier, this implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
+A log-normal distribution of abundance implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
 
@@
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
@@
 \newline
-\verb|https://emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb|
+\verb|https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb|
 % The computer code required to produce the results are given in the following Github repository,

### 2022-05-18 main.tex
nature latex template + competing interests + auth contrib

@@
 pdfsubject={q-bio.NC, q-bio.QM},
-pdfauthor={Matz A. Haugen},
+pdfauthor={},
 pdfkeywords={},
@@
 
+\section{Author contributions}
+X conceived the idea and wrote the paper while Y advised.
+
+\section{Competing interests}
+The authors declare no competing interests.
+
 \section*{Supplement}
@@
 \end{table}
-\bibliographystyle{abbrvnat}
+\bibliographystyle{bst/apa}
 \bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .

### 2022-05-18 main.tex
update references

@@
 \documentclass{article}
+% \documentclass{sn-jnl}
 % \documentclass[review]{elsarticle}
@@
 \usepackage{graphicx}
-\usepackage[square,numbers]{natbib}
+\usepackage[square,compress,numbers]{natbib}
 \usepackage{doi}
@@
 \linenumbers
+
+% \def\bibcommenthead{}%
 
@@
 
-This phenomenon is recorded by \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
+This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
 What is causing such a discrepancy between reality and the current theory of communicable diseases \citep{castro2020turning}?
@@
 \section{Author contributions}
-X conceived the idea and wrote the paper while Y advised.
+X conceived the idea and wrote the manuscript while Y advised.
 
@@
 \end{table}
-\bibliographystyle{bst/apa}
+\bibliographystyle{bst/abbrvnat}
 \bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .

### 2022-05-19 letterToEditor/letter.tex
update methods (heterosk) and letter

@@
 % adjust the page margins
-\usepackage[scale=0.70]{geometry}
+\usepackage[scale=0.73]{geometry}
 
@@
 \vspace{1em}
-We hereby submit a manuscript where we seek to connect the communicable disease paradigm with the recently observed mortality patterns of the initial Coronavirus pandemic in March 2021 that have been shown by many to exhibit Gompertzian growth.
-In the manuscript, we show that this type of growth pattern is incompatible with a traditional communicable disease spreading models, i.e.
+We hereby submit a manuscript where we seek to connect the communicable disease paradigm with the recently observed mortality patterns of the initial Coronavirus pandemic in March-April 2020 that have been shown by many to exhibit Gompertzian growth.
+In the manuscript, we show that this type of growth pattern is incompatible with traditional communicable disease spreading models, i.e.
 the SIR (Susceptible-Infected-Recovered) model family of Kermack and McKendrick.
-Instead, the observed patterns can cleanly be explained by a simpler model without the need for a disease propagating stage, but rather through a ubiquitous stressor which elicits a mutual stress response, amounting to a 2-parameter model.
-The mathematical thesis is based on a simple Stochastic Differential Equation where the stress response is a random process, interpretable both at the macroscopic or the microscopic level. 
+Instead, the observed patterns can cleanly be explained by a simpler model without the need for a disease propagating stage, but rather through a ubiquitous stressor which elicits an instantaneous and mutual stress response, amounting to a 2-parameter model.
+The mathematical thesis is based on a simple Stochastic Differential Equation where the stress response is a random process, interpretable both at the macroscopic and the microscopic level. 
 We also show a remarkable connection between coherent behavior previously relegated to microscopic quantum domains, now exhibited in national mortality patterns.
+In light of this, we equate one of the model parameters in the traditional disease models with the level of \emph{coherency} of growth, where coherency has been rigorously defined in the physics literature (see manuscript).
 
-Our results has the potential to shift the entire field of pathogenic disease modeling away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role. 
-The arguments in the paper are based purely on mathematical and physical grounds as we do not wish to cover too large a scope, but clearly this paper fits into a larger discussion on the communicable disease paradigm which we feel is missing in the current scientific debate.
+The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" calling for a paradigm shift away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role.
 
@@
 
+Further, as we wish to have a double-blind review process, our names are omitted in the manuscript.
+
 Thank you for your consideration.
-
-%narrate why you want to apply
 

### 2022-05-19 main.tex
update methods (heterosk) and letter

@@
 \end{equation}
-obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $c=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
+obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
 The SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
@@
 It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}.
-The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $K=\exp{(-\frac{\sigma}{2\beta})}$, viz.
+The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
 \begin{equation}
 \label{eq:microscopicOriginal}
-dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{K}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
+dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{C}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
 \end{equation}
@@
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
+
+On statistical grounds, one could argue that the daily mortality counts follow a Poisson distribution.
+However, in a growth setting, one might expect heteroskedasticity inherent in the biological process where e.g.
+growth has less variance in the later stages, perhaps dependent on proximity to a stationary state.
+This latter hypothesis could be incorporated in the stochastic framework presented in the text, but to keep the scope of this paper confined we use a simple non-linear least squares minimization as our results are not meaningfully different compared with a more complex variance model.
 

### 2022-05-25 PNAS/PNAS-main.tex
Add pnas draft

--- 2022-05-19 main.tex
@@
-\documentclass{article}
-% \documentclass{sn-jnl}
-% \documentclass[review]{elsarticle}
-
-\usepackage{arxiv}
-\usepackage{booktabs}
-\usepackage{siunitx}
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{lipsum}   % Can be removed after putting your text content
-\usepackage{graphicx}
-\usepackage[square,compress,numbers]{natbib}
-\usepackage{doi}
-\usepackage{amsmath}
-\usepackage{float}
-\usepackage{lineno}
-\linenumbers
-
-% \def\bibcommenthead{}%
-
+\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
+% Use the lineno option to display guide line numbers if required.
 \newcommand{\insertPdfFig}[3]{
-  \begin{figure}[H]
+\begin{figure*}%[tbhp]
 \centering
-  \includegraphics[width=8.4cm]{#1.pdf}
+\includegraphics[width=.8\linewidth]{#1.pdf}
 \caption{#2}
 \label{fig:#1}
-  \end{figure}
+\end{figure*}
 }
 
-\newcommand{\insertSmallPdfFig}[3]{
-  \begin{figure}[h]
-  \centering
-  \includegraphics[width=8.4cm]{#1.pdf}
-  \caption{#2}
-  \label{fig:#1}
-  \end{figure}
-}
+\templatetype{pnasresearcharticle} % Choose template 
+% {pnasresearcharticle} = Template for a two-column research article
+% {pnasmathematics} %= Template for a one-column mathematics article
+% {pnasinvited} %= Template for a PNAS invited submission
 
-% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
-\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
-\graphicspath{{figures/}}
-%\date{September 9, 1985} % Here you can change the date presented in the paper title
-%\date{}          % Or removing it
+\title{Unifying the communicable disease spreading paradigm with Gompertzian growth}
 
-% \author{Matz A. Haugen, \\
-% Independent Reseacher, Oslo\\
-% \texttt{matzhaugen@gmail.com} \\
-%   \And
-%   Dorothea Gilbert \\
-%   University of Oslo\\
-% }
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a,1]{Matz A. Haugen}
+\author[b]{Dorothea Gilbert}
 
-% Uncomment to remove the date
-%\date{}
+\affil[a]{Abel X\\Enerhauggata 1\\0651, Oslo}
+\affil[b]{University of Oslo}
 
-% Uncomment to override  the `A preprint' in the header
-%\renewcommand{\headeright}{Technical Report}
-%\renewcommand{\undertitle}{Technical Report}
-\renewcommand{\shorttitle}{On the nature of the Gompertz}
+% Please give the surname of the lead author for the running footer
+\leadauthor{Haugen} 
+\graphicspath{{../figures/}}
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{
+We seek to connect the communicable disease paradigm with the recently observed mortality patterns of the initial Coronavirus pandemic in March-April 2020 that have been shown by many to exhibit Gompertzian growth.
+In doing this, we show that this type of growth pattern is incompatible with traditional communicable disease spreading models, but rather with a model through which a ubiquitus and instanteneous stressor elicits a mutual stress response.
+The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" suggesting a paradigm shift away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role.}
 
-%%% Add PDF metadata to help others organize their library
-%%% Once the PDF is generated, you can check the metadata with
-%%% $ pdfinfo template.pdf
-\hypersetup{
-pdftitle={A template for the arxiv style},
-pdfsubject={q-bio.NC, q-bio.QM},
-pdfauthor={},
-pdfkeywords={},
-}
+% Please include corresponding author, author contribution and author declaration information
+\authorcontributions{M.A. Haugen and D. Gilbert conceived the idea and M.A. Haugen wrote the manuscript.}
+\authordeclaration{The authors declare no competing interests.}
+% \equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
+\correspondingauthor{\textsuperscript{1}E-mail: matzhaugen@gmail.com}
 
-\begin{document}
-\maketitle
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+\keywords{Gompertz $|$ Coherence $|$ Coronavirus $|$ Network Analysis $|$ Stochastic}
 
@@
 
-% keywords can be removed
-\keywords{Gompertz \and Coherence \and Covid \and Coronavirus \and Network Analysis \and Stochastic}
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
-\section{Introduction}
-Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
+\begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
+
+\dropcap{T}raditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
 Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
@@
 
-A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
+A line of reasoning employed by Rypdal and Rypdal \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
 both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
@@
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
- \citet{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
+ Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
 
@@
  On the other hand, an infection of peripheral nodes should cause immediate exponential growth. 
- \citet{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. 
+ Herrmann and Schwartz \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. 
  Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
@@
 
-\subsection{The SIR model family is almost Gompertzian}
+\subsection*{The SIR model family is almost Gompertzian}
 Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain Gompertzian growth from the communicable disease models, i.e.
@@
 with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$.
-To show this, we follow \citet{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
+To show this, we follow Carletti et al.
+\cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
 This requires an addition to our original set of equations (\ref{eq:SIR}), with a extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
@@
 Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly.
-Rather, all entities are connected and communicating instantaneously as shown by \citet{mombach2002mean}.
+Rather, all entities are connected and communicating instantaneously as shown by Mombach et al.
+\cite{mombach2002mean}.
 However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.
@@
 
-\section{An alternative theory for observed mortality}
+\section*{An alternative theory for observed mortality}
 \label{seq:alt}
@@
 
-Inspired by \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
+Inspired by De Lauro et al. \cite{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
 As with all multiplicative processes, it is convenient to log-transform mortality and work in a dimensionless space such that $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
@@
 
-\section{Unifying the SIR model and the Gompertz model}
+\section*{Unifying the SIR model and the Gompertz model}
 The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
@@
 \end{equation}
-Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.
+Still, no Gompertz curve will emerge at the onset of the growth process. Estrada and Bartesaghi \cite{estrada2022networked} provide further analysis on this topic.
 
-In contrast, as discussed in Section \ref{seq:alt}, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
+In contrast, as discussed above, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
 A log-normal distribution of abundance implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
@@
 We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
-This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
-\citet{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
+This feature of $\theta$ was also obtained by Petroni et al.
+\cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
+Molski and Konarski \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
 This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
-In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields.
+In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber \cite{glauber1963coherent} in the context of electromagnetic fields.
 The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
@@
 
-\newpage
-
-\section*{Methods}
-
-\subsection{Estimating model parameters}
-The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
-\ref{fig:Gompertz_vs_SIRD}).
-Fitting is done using Python-Scipy's non-linear least squares algorithm \verb|curve_fit| with specified Jacobians and a 6-day windowed average of observed deaths.
+\matmethods{
+\subsection*{Estimating model parameters}
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (See SI appendix for details).
+Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths. 
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
@@
 
-\subsection{Code Availability}
+\subsection*{Code Availability}
 
@@
 \newline
-\verb|https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb|
-% The computer code required to produce the results are given in the following Github repository,
-% \begin{verbatim}
-% https://github.com/matzhaugen/GompertzManuscript
-% \end{verbatim}
+\url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb}
+}
 
-\section{Author contributions}
-X conceived the idea and wrote the manuscript while Y advised.
+\showmatmethods{} % Display the Materials and Methods section
 
-\section{Competing interests}
-The authors declare no competing interests.
+% \acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
 
-\section*{Supplement}
+% \showacknow{} % Display the acknowledgments section
 
-% Estimating the parameters using real data requires some consideration around the variance of the growth variable as a function of time.
-This variance will in turn inform the estimation process by considering a weighted minimization of the residuals of the observations regressed onto the model.
-In the stochastic Gompertz process postulated in Section \ref{seq:alt}, the process under a log-transformation is the Ornstein-Uhlenbeck process which has a closed from expression for its transition probability, and thus its variance as a function of time.
-This variance can then be used in a weighted least squares estimation technique or any other weighted technique.
-In the original domain there is also a closed form expression available but it is somewhat more convoluted and given that observations in this domain are visually quite homoscedastic with equal variance across time (Fig \ref{fig:Gompertz_vs_SIRD} and Tables \ref{tab:gomp} and \ref{tab:sird}), this is expression and consideration is omitted.
-But for the log-domain, observations exibit changing variance with time.
-
-% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\citep{risken1996fokker},
-% \begin{equation}
-% P(z, t|z', t') = \sqrt{\frac{\beta}{2\pi \sigma (1 - e^{-2\beta (t-t')})}}\exp{\left[ -\frac{\beta (z - e^{-\beta(t-t')}z')^2}{2\sigma(1-e^{-2 \beta(t-t')})} \right]}.
-% \end{equation}
-
-% If we consider $t$ the time at which the stationary distribution is reached and that this stationary distribution is given as a delta function, variance of the growth will be given as
-% \begin{equation}
-% \text{Var[Z(t)|t']} = \sigma(1-e^{-2 \beta(t-t')}).
-% \end{equation}
-
-% Because weights on observations in a regression are invariant to multiplication we can let $\sigma=1$.
-The slope coefficient would have to be estimated either by a standard least squares or through the 3-parameter model in (\ref{eq:GompertzODE}).
-
-\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
-
-% \insertPdfFig{Gompertz_vs_SIRD_gompfit_log_domain}{Log transformed cumulative diseased plotted against its time derivative after $Y(t) / Y_{max}>0.005$, with a straight line fit regressing onto the $50^{th}$ quantile.}
-
-\begin{table}[H]
-\label{tab:sird}
-\centering
-\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq.
-(\ref{eq:SIRD_D}).
-Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
-\begin{tabular}{llll}
-\toprule
-{} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
-\midrule
-Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
-Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
-Austria        &     8(1) &    36(5) &    11(1) \\
-Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
-Sweden         &    19(5) &     8(2) &     3(1) \\
-Italy          &   137(3) &  1.77(5) &  3.71(1) \\
-France         &  240(10) &  1.24(7) &   8.1(4) \\
-Spain          &   145(2) &  2.27(5) &  4.75(8) \\
-Germany        &    61(4) &   3.9(3) &   6.5(5) \\
-Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
-United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
-Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
-Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
-Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
-US             &   312(5) &  0.76(2) &  1.83(7) \\
-China          &  130(30) &   3.5(5) &    26(4) \\
-\bottomrule
-\end{tabular}
-
-\end{table}
-
-\begin{table}[H]
-\label{tab:gomp}
-\centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq. (\ref{eq:GompertzODE}).}
-\begin{tabular}{llll}
-\toprule
-{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
-\midrule
-Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
-Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
-Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
-Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
-Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
-Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
-France         &  10.2(1) &   8.2(1) &  11.0(8) \\
-Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
-Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
-Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
-United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
-Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
-Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
-Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
-US             &  7.53(7) &  5.32(7) &   8.4(5) \\
-China          &   9.8(1) &   9.4(2) &    10(1) \\
-\bottomrule
-\end{tabular}
-
-\end{table}
-\bibliographystyle{bst/abbrvnat}
-\bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .
-
-%%% Uncomment this section and comment out the \bibliography{references} line above to use inline references.
-% \begin{thebibliography}{1}
-
-%   \bibitem{kour2014real}
-%   George Kour and Raid Saabne.
-%   \newblock Real-time segmentation of on-line handwritten arabic script.
-%   \newblock In {\em Frontiers in Handwriting Recognition (ICFHR), 2014 14th
-%       International Conference on}, pages 417--422. IEEE, 2014.
-
-%   \bibitem{kour2014fast}
-%   George Kour and Raid Saabne.
-%   \newblock Fast classification of handwritten on-line arabic characters.
-%   \newblock In {\em Soft Computing and Pattern Recognition (SoCPaR), 2014 6th
-%       International Conference of}, pages 312--318. IEEE, 2014.
-
-%   \bibitem{hadash2018estimate}
-%   Guy Hadash, Einat Kermany, Boaz Carmeli, Ofer Lavi, George Kour, and Alon
-%   Jacovi.
-%   \newblock Estimate and replace: A novel approach to integrating deep neural
-%   networks with existing applications.
-%   \newblock {\em arXiv preprint arXiv:1804.09028}, 2018.
-
-% \end{thebibliography}
+% Bibliography
+\bibliography{../references2.bib}
 

### 2022-05-25 PNAS/PNAS-template-main.tex
Add pnas draft

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\title{Template for preparing your research report submission to PNAS using Overleaf}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Matz A. Haugen}
\author[b,1,2]{Dorothea Gilbert} 
\author[a]{Author Three}

\affil[a]{Independent Researcher}
\affil[b]{University of Oslo}

% Please give the surname of the lead author for the running footer
\leadauthor{Haugen}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level under- standable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance State- ment is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself and is required for all research papers.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{X and Y conceived the idea and X wrote the manuscript.}
\authordeclaration{The authors declare no competing interests.}
% \equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: matzhaugen\@email.com}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Gompertz $|$ Coherence $|$ Covid $|$ Coronavirus $|$ Network Analysis $|$ Stochastic}

\begin{abstract}
Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{T}his PNAS journal template is provided to help you write your work in the correct journal format.  Instructions for use are provided below.

Note: please start your introduction without including the word ``Introduction'' as a section heading (except for math articles in the Physical Sciences section); this heading is implied in the first paragraphs.

\section*{Guide to using this template on Overleaf}

Please note that whilst this template provides a preview of the typeset manuscript for submission, to help in this preparation, it will not necessarily be the final publication layout. For more detailed information please see the \href{http://www.pnas.org/site/authors/format.xhtml}{PNAS Information for Authors}.

If you have a question while using this template on Overleaf, please use the help menu (``?'') on the top bar to search for \href{https://www.overleaf.com/help}{help and tutorials}. You can also \href{https://www.overleaf.com/contact}{contact the Overleaf support team} at any time with specific questions about your manuscript or feedback on the template.

\subsection*{Author Affiliations}

Include department, institution, and complete address, with the ZIP/postal code, for each author. Use lower case letters to match authors with institutions, as shown in the example. Authors with an ORCID ID may supply this information at submission.

\subsection*{Submitting Manuscripts}

All authors must submit their articles at \href{http://www.pnascentral.org/cgi-bin/main.plex}{PNAScentral}. If you are using Overleaf to write your article, you can use the ``Submit to PNAS'' option in the top bar of the editor window.

\subsection*{Format}

Many authors find it useful to organize their manuscripts with the following order of sections;  Title, Author Affiliation, Keywords, Abstract, Significance Statement, Results, Discussion, Materials and methods, Acknowledgments, and References. Other orders and headings are permitted.

\subsection*{Manuscript Length}

PNAS generally uses a two-column format averaging 67 characters, including spaces, per line. The maximum length of a Direct Submission research article is six pages and a Direct Submission Plus research article is ten pages including all text, spaces, and the number of characters displaced by figures, tables, and equations.  When submitting tables, figures, and/or equations in addition to text, keep the text for your manuscript under 39,000 characters (including spaces) for Direct Submissions and 72,000 characters (including spaces) for Direct Submission Plus.

\subsection*{References}

References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g. \cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric}. All references should be included in the main manuscript file.

\subsection*{Data Archival}

PNAS must be able to archive the data essential to a published article. Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the Information for Authors, is acceptable.

\subsection*{Language-Editing Services}
Prior to submission, authors who believe their manuscripts would benefit from professional editing are encouraged to use a language-editing service (see list at www.pnas.org/site/authors/language-editing.xhtml). PNAS does not take responsibility for or endorse these services, and their use has no bearing on acceptance of a manuscript for publication.

% \begin{figure}%[tbhp]
% \centering
% \includegraphics[width=.8\linewidth]{frog}
% \caption{Placeholder image of a frog with a long example caption to show justification setting.}
% \label{fig:frog}
% \end{figure}

% \begin{SCfigure*}[\sidecaptionrelwidth][t]
% \centering
% \includegraphics[width=11.4cm,height=11.4cm]{frog}
% \caption{This caption would be placed at the side of the figure, rather than below it.}\label{fig:side}
% \end{SCfigure*}

\subsection*{Digital Figures}

Only TIFF, EPS, and high-resolution PDF for Mac or PC are allowed for figures that will appear in the main text, and images must be final size. Authors may submit U3D or PRC files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format.  Color images must be in RGB (red, green, blue) mode. Include the font files for any text.

Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

Figure \ref{fig:frog} shows an example of how to insert a column-wide figure. To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment. Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide. Use \verb|\begin{SCfigure*}...\end{SCfigure*}| for a wide figure with side captions.

\subsection*{Tables}
In addition to including your tables within this manuscript file, PNAS requires that each table be uploaded to the submission separately as a “Table” file.  Please ensure that each table .tex file contains a preamble, the \verb|\begin{document}| command, and the \verb|\end{document}| command. This is necessary so that the submission system can convert each file to PDF.

\subsection*{Single column equations}

Authors may use 1- or 2-column equations in their article, according to their preference.

To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.

Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.

\begin{figure*}[bt!]
\begin{align*}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
       &=x^3+3x^2y+3xy^3+x^3. 
\end{align*}
\end{figure*}

\begin{table}%[tbhp]
\centering
\caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
\begin{tabular}{lrrr}
Species & CBS & CV & G3 \\
\midrule
1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
\bottomrule
\end{tabular}

\addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
\end{table}

\subsection*{Supporting Information (SI)}

Authors should submit SI as a single separate PDF file, combining all text, figures, tables, movie legends, and SI references.  PNAS will publish SI uncomposed, as the authors have provided it.  Additional details can be found here: \href{http://www.pnas.org/page/authors/journal-policies}{policy on SI}.  For SI formatting instructions click \href{https://www.pnascentral.org/cgi-bin/main.plex?form_type=display_auth_si_instructions}{here}.  The PNAS Overleaf SI template can be found \href{https://www.overleaf.com/latex/templates/pnas-template-for-supplementary-information/wqfsfqwyjtsd}{here}.  Refer to the SI Appendix in the manuscript at an appropriate point in the text. Number supporting figures and tables starting with S1, S2, etc.

Authors who place detailed materials and methods in an SI Appendix must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the SI methods. If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.

\subsubsection*{SI Datasets}

Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.

\subsubsection*{SI Movies}

Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files and submit a brief legend for each movie in a Word or RTF file. All movies should be submitted at the desired reproduction size and length. Movies should be no more than 10 MB in size.

\subsubsection*{3D Figures}

Supply a composable U3D or PRC file so that it may be edited and composed. Authors may submit a PDF file but please note it will be published in raw format and will not be edited or composed.

\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required. Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.

\subsection*{Subsection for Method}
Example text for subsection.
}

\showmatmethods{} % Display the Materials and Methods section

\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{pnas-sample}

\end{document}

### 2022-05-25 letterToEditor/letter.tex
Add pnas draft

@@
 \recipient{To}{Editor, \\
-               Nature,\\ 
+               Proceedings of the National Academy of Sciences,\\ 
                }
@@
 
-The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" calling for a paradigm shift away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role.
+The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" suggesting a paradigm shift away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role.
 

### 2022-05-26 PNAS/PNAS-main.tex
move figures inside pnas folder
update letter w refereese + slight mod to manuscript

@@
 
-\affil[a]{Abel X\\Enerhauggata 1\\0651, Oslo}
-\affil[b]{University of Oslo}
+\affil[a]{Independent Researcher, Oslo}
+\affil[b]{Independent Researcher, Oslo}
 
@@
 \leadauthor{Haugen}
-\graphicspath{{../figures/}}
+
 % Please add here a significance statement to explain the relevance of your work
@@
 In doing this, we show that this type of growth pattern is incompatible with traditional communicable disease spreading models, but rather with a model through which a ubiquitus and instanteneous stressor elicits a mutual stress response.
-The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" suggesting a paradigm shift away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role.}
+The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" suggesting a paradigm shift away from that of a purely communicable disease paradigm to a hybrid where the environment plays a bigger role.}
 
@@
 
-\subsection{Unification through a modified SIR model}
-However, it is possible to reconcile the SIR model with the Gompertz curve. 
+\subsection*{Unification through a modified SIR model}
+It is also possible to reconcile the SIR model with the Gompertz curve. 
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
@@
 \begin{equation}
-\frac{d}{{dt}} \ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} - \alpha.
+\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}\sum_i^N x_i.
 \end{equation}
@@
 The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (See SI appendix for details).
-Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths. 
+Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths, similar to the technique by Carletti et al.
+\citep{carletti2020covid}.
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
 
-On statistical grounds, one could argue that the daily mortality counts follow a Poisson distribution.
+% On statistical grounds, one could argue that the daily mortality counts follow a Poisson distribution.
 However, in a growth setting, one might expect heteroskedasticity inherent in the biological process where e.g.

### 2022-05-26 letterToEditor/letter.tex
update letter w refereese + slight mod to manuscript

@@
 \enclosure[Enclosed ]{
-              \\ List of referees
+              \\ List of referees, NAS members and PNAS editorial board members
 %                  \hspace{0.0em} 5. GATE (2016) \hspace{2em} 9. Identity-Proof (Voter-Id)\\
@@
 \newpage
+
+\textbf{List of Editorial board members}
+\begin{enumerate}
+\item Levine, Herbert
+\item Levin, Simon A.
+\item Marquet, Pablo A.
+\end{enumerate}
+
+\textbf{List of NAS members (in order of relevance)} 
+\begin{enumerate}
+\item Levine, Herbert
+\item Levin, Simon A.
+\item Michael Levitt
+
+\item Nigel D. Goldenfeld 
+\item Evans, Steven N.
+\item Altshuler, Boris L.
+\end{enumerate}
 \textbf{List of possible referees}
 \begin{enumerate}
-\item N. El Karoui, mathematics, stochastic models
-\item M. Molski, physics, Gompertz and quantum systems
-\item W. Whitt, mathematics, stochastic models
-\item N. C. Petroni, E. De Lauro, physics, Gompertz
-\item E. B. Postnikov, physics, SIR models
-\item G. Vattay, physics, SIR models
-\item M. Levitt, computational biology, Gompertz models
-\item E. Estrada, mathematics, graph theory, SIR models
-\item T. Carletti, D. Fanelli, and F. Piazza, physics, SIR models
-\item Z. Bajzer, biology, has a very good review paper on Gompertz modeling 
-\item J. C. Mombach, physics, growth models
-\item B. Shklovskii, physics, theory of Gompertz
-\item D. Smilkov, C. or L. Kocarev, physics and mathematics, SIR models
-\item C. G. Antonopoulos, physics, chaos and SIR models
-\item K. Rypdal and M. Rypdal, mathematics, SIR models
+
+\item Nicola Cufaro Petroni, cufaro@ba.infn.it,
+Dipartimento di Matematica and TIRES, University of Bari (Ret).
+Has done extensive work on the Gompertz model in relation to the Richard's model, connecting it with stochastic growth models and the Logistic (Lotka-Volterra) model.
+We use their concept of growth rooted in stochastic phenomena.
+\item Salvatore De Martino,
+demartino247@gmail.com, 
+Dipartimento di Ingegneria dell’Informazione ed Elettrica.
+Same as above
+\item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
+\item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
+Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
+\item Ernesto Estrada, estrada@ifisc.uib-csic.es, Institute of Interdisciplinary Physics and Complex Systems , mathematics, graph theory, SIR models.
+Has written a recent paper on the Gompertz model in relation to communicable disease modelling including close connections with graph theoretical concepts used in our paper. 
+\item K. Rypdal, kristoffer.rypdal@uit.no, University of Tromsø, mathematics, SIR models.
+Has written a paper showing that SIR models become Gompertz in the extreme of the Richard's parameter. We also follow this paper's arguments in our paper.
+\item Michael Levitt (NAS Member), michael.levitt@stanford.edu, Stanford University, computational biology, Gompertz models.
+Has a preprint on the gompertz model in relation to covid where they leave the explanation of why mortality follows gompertz as an open question.
+\item Chris G. Antonopoulos, canton@essex.ac.uk , U. of Essex, physics, chaos and SIR models.
+Has written on the SIR model and chaotic dynamics.
+\item Gabor Vattay, vattay@elte.hu, Eötvös Loránd University Budapest, physics, SIR models
+Has written a paper on SIR models showing that there are ``initial transients'' in the observations which are explained by us in the Gompertz model.
+\item E. B. Postnikov, postnikov@kursksu.ru, Kursk State Univesity, physics, SIR models.
+Has done work on SIR models using a mean field approach, but has a broach research interest that could be a benefit when reviewing our paper.
+\item Both H. Levine and Levin A. in the NAS list would be quite appropriate.
+\item L. Kocarev, lkocarev@manu.edu.mk, Macedonian Academy of Sciences and Arts
+, physics and mathematics, SIR models.
+Has written papers on SIR models as they relate to non-markovian models.
 

### 2022-05-27 PNAS/PNAS-main.tex
fix small things

@@
 Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
-A large body of research has emerged that explores how these models fit the recent mortality seen due to the Coronavirus epidemic \citep{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
+A large body of research reveals how these models fit the recent mortality seen due to the Coronavirus epidemic \citep{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
 
@@
 A line of reasoning employed by Rypdal and Rypdal \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
-both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
+both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, are much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
 \begin{align}
@@
 \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of equations (\ref{eq:SIR}), with a extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
+This requires an addition to our original set of equations (\ref{eq:SIR}), with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 
-Some algebraic manipulations reveal that the diseased group is described by a single equation, viz.
+Algebraic manipulations reveal that the diseased group is described by a single equation, viz.
 \begin{equation}
@@
 where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is again synonymous to the initial susceptible pool of people.
-This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$.
-When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.
+This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$.
+However, when only modeling the diseased, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.
 
@@
 With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig.
-\ref{fig:Gompertz_vs_SIRD_lnln_infty} and Supplemental Information).
+\ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
 Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in (\ref{eq:GompertzODE}), it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
@@
 It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}.
-The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
+As argued by De Lauro et al.
+\cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation.
+This is seen if we recast this equation in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
 \begin{equation}
@@
 
-To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}) and transform back to the original domain, 
+To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}), 
 \begin{equation}
@@
 \subsection*{Estimating model parameters}
-The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (See SI appendix for details).
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (see SI Appendix for details).
 Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths, similar to the technique by Carletti et al.

### 2022-05-27 PNAS/PNAS-main.tex
change equation ref

@@
 \label{eq:SIR}
-\frac{dS}{dt}& = -\beta \frac{IS}{N}\\
+\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
 \frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
-\frac{dR}{dt}& = \alpha I,
+\frac{dR}{dt}& = \alpha I\nonumber,
 \end{align}
@@
  Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
-And since $I$ is a function of $Y$, we can combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
+And since $I$ is a function of $Y$, we can combine Eq.
+\ref{eq:linSIR} via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
 \begin{subequations}
@@
 Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain Gompertzian growth from the communicable disease models, i.e.
-a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
+a straight line under a double-log transform of Eq.
+\ref{eq:gomp_solution}, viz.
 \begin{equation}
@@
 \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of equations (\ref{eq:SIR}), with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
+This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$.
-However, when only modeling the diseased, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.
+However, when only modeling the diseased, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.
 
@@
 \end{equation}
-obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
+obtained with a second order expansion of the exponential term in Eq.
+\ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
 The SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
@@
 \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
-Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in (\ref{eq:GompertzODE}), it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
+Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq.
+\ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
 \begin{equation}
@@
 \insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GompertzODE2param}), it is here obtained using non-linear least squares to put both models on equal footing.
+Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text.
+Although the Gompertz curve can also be obtained with a simple linear fit using Eq.
+\ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing.
 The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
@@
 
-To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}), 
+To obtain the deterministic observable, we first take the average in the log-domain (Eq. \ref{eq:microscopic}), 
 \begin{equation}
@@
 where the bra-ket notation signifies the averaging operation.
-Then we use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $M(t)$, 
+Then we use the property that the average of log-quantities is the logarithm of the median quantity \cite{petroni2020gompertz}, where we denote the median of $F(t)$ as $M(t)$, 
 \begin{equation}
@@
 \end{equation}
-which corresponds with the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}) with $Y(t) = Y_\infty M(t)$. 
-Comparing the stochastic stressor $\sigma$ in (\ref{eq:microscopicOriginal}) with the deterministic growth equation in (\ref{eq:GompertzODE}) gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
-We also see that by comparing (\ref{eq:rypdalODE}) and (\ref{eq:rypdalLimit}) with the stochastic counterpart in (\ref{eq:microscopicOriginal}) that the final growth level is governed by the stressor magnitude,
+which corresponds with the familiar deterministic Gompertz differential equation in Eq. \ref{eq:GompertzODE} with $Y(t) = Y_\infty M(t)$. 
+Comparing the stochastic stressor $\sigma$ in Eq.
+\ref{eq:microscopicOriginal} with the deterministic growth equation in Eq.
+\ref{eq:GompertzODE} gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
+We also see that by comparing equations \ref{eq:rypdalODE} and \ref{eq:rypdalLimit} with the stochastic counterpart in Eq.
+\ref{eq:microscopicOriginal} that the final growth level is governed by the stressor magnitude,
 \begin{equation}
@@
 Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}.
-Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in (\ref{eq:microscopicOriginal}) \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
-Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
+Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
+Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (Eq.
+\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
 This implies that the central limit theorem applies to the log-domain.
@@
 
-One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compareLog})
+One could add network interaction necessitating a corresponding matrix version of Eq. \ref{eq:compareLog}
 \begin{equation}
@@
 This feature of $\theta$ was also obtained by Petroni et al.
-\cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
+\cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
+\ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
 Molski and Konarski \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
@@
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
-Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
+Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
 \begin{equation}
@@
 An interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
-One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain (\ref{eq:medianGomp}),
+One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq.
+\ref{eq:medianGomp},
 \begin{equation}
@@
 \subsection*{Estimating model parameters}
-The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (see SI Appendix for details).
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text (see SI Appendix for details).
 Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths, similar to the technique by Carletti et al.
@@
 
-The transmission rate is modeled as the product of $\eta \xi$ in (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
+The transmission rate is modeled as the product of $\eta \xi$ in Eq.
+\ref{eq:SIRD_D}, and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
 \begin{equation}

### 2022-05-27 PNAS/PNAS-main.tex
constrict log-median statement

@@
 
-This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}.
+This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}, where $Z(t)$ is Gaussian.
 It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}.
@@
 \cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation.
-This is seen if we recast this equation in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
+This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
 \begin{equation}
@@
 where the bra-ket notation signifies the averaging operation.
-Then we use the property that the average of log-quantities is the logarithm of the median quantity \cite{petroni2020gompertz}, where we denote the median of $F(t)$ as $M(t)$, 
+Then we use the property that for the log-normal quantity $F(t)$, the average of its logarithm is the logarithm of the \emph{median} quantity \cite{petroni2020gompertz}, where we denote the median of $F(t)$ as $M(t)$,
 \begin{equation}

### 2022-05-27 PNAS/PNAS-main.tex
typos

@@
 We seek to connect the communicable disease paradigm with the recently observed mortality patterns of the initial Coronavirus pandemic in March-April 2020 that have been shown by many to exhibit Gompertzian growth.
-In doing this, we show that this type of growth pattern is incompatible with traditional communicable disease spreading models, but rather with a model through which a ubiquitus and instanteneous stressor elicits a mutual stress response.
+In doing this, we show that this type of growth pattern is incompatible with traditional communicable disease spreading models, but rather with a model through which a ubiquitous and instantaneous stressor elicits a mutual stress response.
 The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" suggesting a paradigm shift away from that of a purely communicable disease paradigm to a hybrid where the environment plays a bigger role.}

### 2022-06-06 elsarticle/elsarticle.tex
update elsarticle

@@
 
-\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
+\title{Unifying the communicable disease spreading paradigm with Gompertzian growth}
 
@@
 
-\author[mysecondaryaddress]{Dorothea Gilbert}
+\author[mymainaddress]{Dorothea Gilbert}
 
-\address[mymainaddress]{Enerhauggata 1, 0651 Oslo, Norway}
-\address[mysecondaryaddress]{University of Oslo, Oslo, Norway}
+\address[mymainaddress]{Independent Researcher, Oslo, Norway}
+% \address[mysecondaryaddress]{Independent Researcher, Oslo, Norway}
 
@@
 \label{eq:SIR}
-\frac{dS}{dt}& = -\beta \frac{IS}{N}\\
-\frac{dI}{dt}& = -\beta \frac{IS}{N} - \alpha I.\\
-\frac{dR}{dt}& = \alpha I,
+\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
+\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
+\frac{dR}{dt}& = \alpha I\nonumber,
 \end{align}
@@
 
-A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
-both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the total population, $N \gg Y \ge I$, which seems well founded at the national level, viz.
+A line of reasoning employed by Rypdal and Rypdal \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
+both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, are much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
 \begin{align}
@@
 
- At this limit the growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
+ At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
  The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning.
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
- \citet{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
+ Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
 
@@
 \subsection{The SIR model family is almost Gompertzian}
-Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain from the communicable disease models growth which is almost Gompertzian, i.e.
-a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
+Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain Gompertzian growth from the communicable disease models, i.e.
+a straight line under a double-log transform of Eq.
+\ref{eq:gomp_solution}, viz.
 \begin{equation}
@@
 with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$.
-To show this, we follow \citet{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of equations (\ref{eq:SIR}), with a extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
+To show this, we follow Carletti et al.
+\cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
+This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 \end{equation}
-where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonymous to the initial susceptible pool of people.
-This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$.
-When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.
+where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is again synonymous to the initial susceptible pool of people.
+This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$.
+However, when only modeling the diseased, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.
 
@@
 \begin{equation}
-\frac{dD}{dt} = b D (1 - D/K),
+\frac{dD}{dt} = c D (1 - D/K),
 \end{equation}
-obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $b=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
-The SIR and SEIR model exhibits similar properties and its discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
+obtained with a second order expansion of the exponential term in Eq.
+\ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
+The SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
 
 With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig.
-\ref{fig:Gompertz_vs_SIRD_lnln_infty} and Supplemental Information).
-Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in (\ref{eq:GompertzODE}), it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
+\ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
+Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq.
+\ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
 \begin{equation}
@@
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model with a Gompertz model for a variety of countries in the period Jan-May 2020.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GompertzODE2param}), it is here obtained using non-linear least squares to put both models on equal footing.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
+Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text.
+Although the Gompertz curve can also be obtained with a simple linear fit using Eq.
+\ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing.
 The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
 Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates.
-Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
+Fitting is done with Python-Scipy and a 6-day moving average of deaths.
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
@@
 
-This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}.
+This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}, where $Z(t)$ is Gaussian.
 It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}.
-The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $K=\exp{(-\frac{\sigma}{2\beta})}$, viz.
+As argued by De Lauro et al.
+\cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation.
+This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
 \begin{equation}
 \label{eq:microscopicOriginal}
-dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{K}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
+dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{C}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
 \end{equation}
 
-To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}) and transform back to the original domain, 
+To obtain the deterministic observable, we first take the average in the log-domain (Eq. \ref{eq:microscopic}), 
 \begin{equation}
@@
 where the bra-ket notation signifies the averaging operation.
-Then we use the property that the average of log-quantities is the logarithm of the median quantity, where we denote the median of $F(t)$ as $M(t)$, 
+Then we use the property that for the log-normal quantity $F(t)$, the average of its logarithm is the logarithm of the \emph{median} quantity \cite{petroni2020gompertz}, where we denote the median of $F(t)$ as $M(t)$,
 \begin{equation}
@@
 \end{equation}
-which corresponds with the familiar deterministic Gompertz differential equation in (\ref{eq:GompertzODE}) with $Y(t) = Y_\infty M(t)$. 
-Comparing the stochastic stressor $\sigma$ in \ref{eq:microscopicOriginal} with the deterministic growth equation in (\ref{eq:GompertzODE}) gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
-We also see that by comparing (\ref{eq:rypdalODE}) and (\ref{eq:rypdalLimit}) with the stochastic counterpart in (\ref{eq:microscopicOriginal}) that the final growth level is governed by the stressor magnitude,
+which corresponds with the familiar deterministic Gompertz differential equation in Eq. \ref{eq:GompertzODE} with $Y(t) = Y_\infty M(t)$. 
+Comparing the stochastic stressor $\sigma$ in Eq.
+\ref{eq:microscopicOriginal} with the deterministic growth equation in Eq.
+\ref{eq:GompertzODE} gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
+We also see that by comparing equations \ref{eq:rypdalODE} and \ref{eq:rypdalLimit} with the stochastic counterpart in Eq.
+\ref{eq:microscopicOriginal} that the final growth level is governed by the stressor magnitude,
 \begin{equation}
@@
 Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}.
-Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in (\ref{eq:microscopicOriginal}) \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
-Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
-This implies that the log-domain is the domain in which the central limit theorem applies.
+Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
+Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (Eq.
+\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
+This implies that the central limit theorem applies to the log-domain.
 
 \section{Unifying the SIR model and the Gompertz model}
-This remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
+The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
 
@@
 \end{equation}
-Naturally, this simple arithmetic average treats each microscopic unit as independent variables contributing to the macroscopic observable.
+Naturally, this simple arithmetic average treats each microscopic unit as an independent variable contributing to the macroscopic observable.
 
-One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compareLog})
+One could add network interaction necessitating a corresponding matrix version of Eq. \ref{eq:compareLog}
 \begin{equation}
@@
 \end{equation}
-using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
-Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below.
+using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth or infection rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise.
+Notice that there is an implied causality from the infected to the susceptible, which will become relevant below.
 Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
@@
 \end{equation}
-Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.
+Still, no Gompertz curve will emerge at the onset of the growth process. Estrada and Bartesaghi \cite{estrada2022networked} provide further analysis on this topic.
 
-In contrast, as discussed in Section \ref{seq:alt}, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
-As mentioned earlier, this implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
+In contrast, as discussed above, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
+A log-normal distribution of abundance implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
 
@@
 We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
-This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
-\citet{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
+This feature of $\theta$ was also obtained by Petroni et al.
+\cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
+\ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
+Molski and Konarski \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
 This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
-In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields.
-The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is quite noteworthy.
-It suggests that both systems share commonalities and means of communication.
+In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber \cite{glauber1963coherent} in the context of electromagnetic fields.
+The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.
 
 \subsection{Unification through a modified SIR model}
-However, it is possible to reconcile the SIR model with the Gompertz curve. 
+It is also possible to reconcile the SIR model with the Gompertz curve. 
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
-Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
+Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
 \begin{equation}
@@
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter, $\theta$, in his growth model paradigm can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
@@
 
-\section*{Methods}
+\section{Methods}
 
 \subsection{Estimating model parameters}
-The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
-\ref{fig:Gompertz_vs_SIRD}).
-Fitting is done with Python-Scipy with specified Jacobians and a 6-day windowed average of observed deaths.
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text (see SI Appendix for details).
+Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths, similar to the technique by Carletti et al.
+\citep{carletti2020covid}.
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
 
-The transmission rate is modeled as the product of $\eta \xi$ in (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
+% On statistical grounds, one could argue that the daily mortality counts follow a Poisson distribution.
+However, in a growth setting, one might expect heteroskedasticity inherent in the biological process where e.g.
+growth has less variance in the later stages, perhaps dependent on proximity to a stationary state.
+This latter hypothesis could be incorporated in the stochastic framework presented in the text, but to keep the scope of this paper confined we use a simple non-linear least squares minimization as our results are not meaningfully different compared with a more complex variance model.
+
+The transmission rate is modeled as the product of $\eta \xi$ in Eq.
+\ref{eq:SIRD_D}, and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
 \begin{equation}
-\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
+\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\text{E}[\eta]^2 + \text{Var}[\eta] \text{E}[\xi]^2,
 \end{equation}
+where estimates are obtained by substituting the model parameters with those estimated through the technique outlined above.
 
 \subsection{Code Availability}
-The computer code required to produce the results are given in the following Github repository,
-\begin{verbatim}
-https://github.com/matzhaugen/GompertzManuscript
-\end{verbatim}
 
-\section*{Supplement}
+Scripts used to produce the plots are downloadable in the form of a Jupyter notebook using Python from 
+\newline
+\url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb}
 
-\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
+\section{Competing interests}
+The authors declare no competing interests.
+
+\section{Author Contributions}
+M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
+
+% \section{Supplement}
+
+% \insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
 Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
@@
 
-\begin{table}[H]
-\label{tab:sird}
-\centering
-\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
+% \begin{table}[H]
+% \label{tab:sird}
+% \centering
+% \caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
 Standard deviation is given in the paranthesis corresponding to the last significant digit of the reported number.}
-\begin{tabular}{llll}
-\toprule
-{} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
-\midrule
-Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
-Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
-Austria        &     8(1) &    36(5) &    11(1) \\
-Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
-Sweden         &    19(5) &     8(2) &     3(1) \\
-Italy          &   137(3) &  1.77(5) &  3.71(1) \\
-France         &  240(10) &  1.24(7) &   8.1(4) \\
-Spain          &   145(2) &  2.27(5) &  4.75(8) \\
-Germany        &    61(4) &   3.9(3) &   6.5(5) \\
-Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
-United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
-Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
-Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
-Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
-US             &   312(5) &  0.76(2) &  1.83(7) \\
-China          &  130(30) &   3.5(5) &    26(4) \\
-\bottomrule
-\end{tabular}
+% \begin{tabular}{llll}
+% \toprule
+% {} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
+% \midrule
+% Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
+% Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
+% Austria        &     8(1) &    36(5) &    11(1) \\
+% Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
+% Sweden         &    19(5) &     8(2) &     3(1) \\
+% Italy          &   137(3) &  1.77(5) &  3.71(1) \\
+% France         &  240(10) &  1.24(7) &   8.1(4) \\
+% Spain          &   145(2) &  2.27(5) &  4.75(8) \\
+% Germany        &    61(4) &   3.9(3) &   6.5(5) \\
+% Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
+% United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
+% Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
+% Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
+% Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
+% US             &   312(5) &  0.76(2) &  1.83(7) \\
+% China          &  130(30) &   3.5(5) &    26(4) \\
+% \bottomrule
+% \end{tabular}
 
-\end{table}
+% \end{table}
 
-\begin{table}[H]
-\label{tab:gomp}
-\centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.
+% \begin{table}[H]
+% \label{tab:gomp}
+% \centering
+% \caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.
 Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
-\begin{tabular}{llll}
-\toprule
-{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
-\midrule
-Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
-Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
-Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
-Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
-Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
-Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
-France         &  10.2(1) &   8.2(1) &  11.0(8) \\
-Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
-Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
-Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
-United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
-Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
-Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
-Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
-US             &  7.53(7) &  5.32(7) &   8.4(5) \\
-China          &   9.8(1) &   9.4(2) &    10(1) \\
-\bottomrule
-\end{tabular}
+% \begin{tabular}{llll}
+% \toprule
+% {} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
+% \midrule
+% Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
+% Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
+% Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
+% Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
+% Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
+% Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
+% France         &  10.2(1) &   8.2(1) &  11.0(8) \\
+% Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
+% Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
+% Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
+% United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
+% Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
+% Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
+% Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
+% US             &  7.53(7) &  5.32(7) &   8.4(5) \\
+% China          &   9.8(1) &   9.4(2) &    10(1) \\
+% \bottomrule
+% \end{tabular}
 
-\end{table}
+% \end{table}
 
-% \section*{References}
+% \section{References}
 

### 2022-06-09 letterToEditor/letter.tex
fix small things
updates

@@
 \recipient{To}{Editor, \\
-               Proceedings of the National Academy of Sciences,\\ 
+               Chaos, Solitons, and Fractals,\\ 
                }
@@
 \enclosure[Enclosed ]{
-              \\ List of referees, NAS members and PNAS editorial board members
+              \\ List of referees
 %                  \hspace{0.0em} 5. GATE (2016) \hspace{2em} 9. Identity-Proof (Voter-Id)\\
@@
 
-Further, as we wish to have a double-blind review process, our names are omitted in the manuscript.
-
 Thank you for your consideration.
@@
 
-\textbf{List of Editorial board members}
-\begin{enumerate}
-\item Levine, Herbert
-\item Levin, Simon A.
-\item Marquet, Pablo A.
-\end{enumerate}
+% \textbf{List of Editorial board members}
+% \begin{enumerate}
+% \item Levine, Herbert
+% \item Levin, Simon A.
+% \item Marquet, Pablo A.
+% \end{enumerate}
 
-\textbf{List of NAS members (in order of relevance)} 
-\begin{enumerate}
-\item Levine, Herbert
-\item Levin, Simon A.
-\item Michael Levitt
+% \textbf{List of NAS members (in order of relevance)} 
+% \begin{enumerate}
+% \item Levine, Herbert
+% \item Levin, Simon A.
+% \item Michael Levitt
 
-\item Nigel D. Goldenfeld 
-\item Evans, Steven N.
-\item Altshuler, Boris L.
-\end{enumerate}
+% \item Nigel D. Goldenfeld 
+% \item Evans, Steven N.
+% \item Altshuler, Boris L.
+% \end{enumerate}
 \textbf{List of possible referees}
@@
 Has done work on SIR models using a mean field approach, but has a broach research interest that could be a benefit when reviewing our paper.
-\item Both H. Levine and Levin A. in the NAS list would be quite appropriate.
 \item L. Kocarev, lkocarev@manu.edu.mk, Macedonian Academy of Sciences and Arts

### 2022-06-24 letterToEditor/letter.tex
update letter

@@
 \date{\today}
-\opening{\textbf{Initial submission of manuscript titled: ``Unifying Gompertzian growth with the communicable disease spreading paradigm''}}
+\opening{\textbf{Initial submission of manuscript titled: ``Unifying the communicable disease spreading paradigm with Gompertzian growth''}}
 \closing{Sincerely yours, \vspace{-1em}}
@@
 the SIR (Susceptible-Infected-Recovered) model family of Kermack and McKendrick.
-Instead, the observed patterns can cleanly be explained by a simpler model without the need for a disease propagating stage, but rather through a ubiquitous stressor which elicits an instantaneous and mutual stress response, amounting to a 2-parameter model.
+Instead, the observed patterns can be explained by a simpler model without the need for a disease propagating stage, but rather through a ubiquitous stressor which elicits an instantaneous and mutual stress response, amounting to a 2-parameter model.
 The mathematical thesis is based on a simple Stochastic Differential Equation where the stress response is a random process, interpretable both at the macroscopic and the microscopic level. 
@@
 Has written papers on SIR models as they relate to non-markovian models.
+\item Constantino Tsallis, tsallis@cbpf.br, Complexity Science Hub Vienna, Austria.
+Has done extensive work on long-range interaction and statistical mechanics.
+He might be able to see connections with his famous Tsallis entropy.
 

### 2022-06-24 elsarticle/elsarticle.tex
updates
add line re. wang
Add animal data

@@
 %% or include affiliations in footnotes:
-\author[mymainaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
+\author[mymainaddress, mysecondaryaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
 \cortext[mycorrespondingauthor]{Corresponding author}
@@
 
-\author[mymainaddress]{Dorothea Gilbert}
+\author[mysecondaryaddress]{Dorothea Gilbert}
 
-\address[mymainaddress]{Independent Researcher, Oslo, Norway}
-% \address[mysecondaryaddress]{Independent Researcher, Oslo, Norway}
+\address[mymainaddress]{Enerhauggata 1, 0651, Oslo, Norway}
+\address[mysecondaryaddress]{Independent Researcher, Oslo, Norway}
 
@@
  Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
+In fact, Wang et al.
+\cite{wang2012richards} showed that values of $\theta<1$ are not biologically realistic under a similar SIR model as the one given above.
 
@@
 An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.
-This could explain the immediate dampening of mortality growth observed. 
-It could also explain the lack of correlation between population density and mortality (or infection) rates observed by some \citep{Hamidi2020,Hamidi2020a,Carozzi2020,Arpino2020,khavarian2021high,barak2021urban}.
+This could explain the immediate dampening of mortality growth observed not explainable by a SIR model. 
+It could also explain the lack of correlation between population density and mortality (or infection) rates observed by some \citep{Hamidi2020,Hamidi2020a,Carozzi2020,Arpino2020,khavarian2021high,barak2021urban}, in addition to the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021,NINA2020}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
 

### 2022-07-14 ReviewerResponse.md

Reviewer responses
------------------

Reviewer 1:
-----------
The reviewer makes the following conjecture:
"The fact that the COVID-19 mortality can be accurately reproduced by the Gompertz curve does not seems to be very surprising taking into account that this dynamical dependence was initially developed to describe mortality curve and confirmed in this application by a variety of examples during almost two centuries"

While Gompertzian growth has been shown to exist in natural growth phenomena like cancer growth, plant leaf growth, etc., it has not previously been shown in the communicable disease paradigm before covid, e.g. at large spatial distances (e.g. across countries). This should indeed be surprising for the following reason: In smaller spatial scales, e.g. bacterial growth as the reviewer alludes to in Lavrova et al., the exponentially limiting growth factor is quite plausible. In the larger national scales, this exponential limiting factor is not plausible nor possible under the SIR/SIRD model, as we show in the introduction. The SIR/SIRD models imply a logistic growth curve, which is indeed not what we see. This conclusion is only possible due to our daily cadence of mortality data, not available for previous epidemics. In a weekly cadence, for example, the separation between the Gompertz and the Logistic is more tenuous and less convincing.

As we discuss in the paper and as has been discussed by references like Petroni et al., Gompertzian growth implies coherent or collaborating growth. In a communicable disease paradigm, we show that this collaboration can only result from infinite interactions and communication speed as exemplified in our proposed network in section 3.1. This is due to the constraint that people must be in contact to propagate the disease. In the reviewer's example from Lavrova, this constraint is not present and is indeed modeled as a ubiquitous oxygen concentration. This is a crucial difference that seems to have been missed by the reviewer.

Further regarding the reviewers comment:
"It also should be noted that the statement "a discrepancy between reality and the current theory of communicable diseases" does not seems to be correct because the classic (either SIR or SEIR)  theory of communicable diseases does not distinguish between "Removed" and "Recovered" persons as the variable "R". [...] Thus, a more motivated statement should be about developing a subdivision of an epidemiological model aimed at distinguishing between two outcomes."

This is exactly why we include a model that does distinguish between "Removed" and "Recovered" in the SIRD model and show that it leads to the logistic curve. Further, the premise of the paper is not to see what additional parameters must be added to the SIRD-like models to obtain Gompertzian growth, but rather to show that with only 2 parameters, under the environmental stressor model, the observations are accurately predicted. We do in fact leave it as an open question when we address potential networks that can lead to Gompertz, but do provide one such network in section 3. This seems much more "motivated" due to the parsimonious nature of the environmental stressor model, as opposed to adding to a model that already has more parameters and yet cannot accurately predict observations.

Regarding comment:
"Although the authors introduced SIRD model further in the text, the clear description of interplay about assumptions of both (SIR and SIRD) approaches should be discussed with the premises of both."

As we already discuss the SIRD and the SIR model based on the same premises, this comment is not clear. We introduce the "Dead" or "Removed" group for the exact reasons outlined by the reviewer himself above. A further elaboration of what kind of "premises" should be discussed is unclear.

"In addition, it should be pointed out that Ref. [5] already references the work Anastassopoulou at al. (2020). Data-based analysis, modelling and forecasting of the COVID-19 outbreak. PloS One, 15, e0230405 as that where SIRD model was introduced." 
It is not clear how this is relevant to the manuscript. We chose to follow Carletti due to the clarity of their exposition, not because they were the first ones to use the SIRD model.

Our modeling framework of the Gompertz does of course have wide implications with regards to the mechanisms of growth: We postulate that growth is taking place with instant communication between units. In the case of plant photosynthesis this is not so far fetched considering that electrons in the electron transport chain are already shown to be non-local, i.e. wavelike rather than particle like.  See for example G. Panitchayangkoon et al. "Long-lived quantum coherence in photosynthetic complexes at physiological temperature" PNAS 2010. This of course breaks down the chemical paradigm where molecules are localized in regular biological processes.

Although it was not the wish of the authors to present too much observational data that does not fit with the current virus-theory of covid, but rather focus on a mathematical model that does indeed explain this simple set of daily mortality data with a simple 2-parameter model, it may be illuminating to the reviewers and editor to note that many animals have been observed to die at anomalous rates during the spring of 2020.

- Fischer L, Peters M, Merbach S, Eydner M, Kuczka A, Lambertz J, Kum- merfeld M, Kahnt K, Weiss A, Petersen H. Increased mortality in wild tits in north rhine-westphalia (germany) in 2020 with a special focus on suttonella ornithocola and other infectious pathogens. European Journal of Wildlife Research 2021;67(3). doi:10.1007/s10344-021-01500-7.
- Peris ́e-Barrios AJ, Tomeo-Mart ́ın BD, G ́omez-Ochoa P, Delgado-Bonet P, Plaza P, Palau-Concejo P, Gonz ́alez J, Ortiz-D ́ıez G, Mel ́endez- Lazo A, Gentil M, Garc ́ıa-Castro J, Barbero-Fern ́andez A. Humoral responses to SARS-CoV-2 by healthy and sick dogs during the COVID- 19 pandemic in spain. Veterinary Research 2021;52(1). doi:10.1186/ s13567-021-00897-y. 
- Duff P, Fenemore C, Holmes P, Hopkins B, Jones J, He M, Everest D, Rocchi M. Disease surveillance in england and wales, december 2019. Vet- erinary Record 2020;186(1):15–6. doi:10.1136/vr.m68
- HuB,WeiH,FanZ,SongY,ChenM,QiuR,ZhuW,XuW,XueJ,Wang F. Emergence of rabbit haemorrhagic disease virus 2 in china in 2020. Veterinary Medicine and Science 2020;7(1):236–9. doi:10.1002/vms3.332.
- Fukui H, Shimoda H, Kadekaru S, Henmi C, Une Y. Rabbit hemorrhagic disease virus type 2 epidemic in a rabbit colony in japan. Journal of Vet- erinary Medical Science 2021;83(5):841–5. doi:10.1292/jvms.21-0007.
- van Aarde RJ, Pimm SL, Guldemond R, Huang R, Mar ́e C. The 2020 elephant die-off in botswana. PeerJ 2021;9:e10686. doi:10.7717/peerj. 10686.
- Kambayashi Y, Bannai H, Tsujimura K, Hirama A, Ohta M, Nemoto M. Outbreak of equine coronavirus infection among riding horses in tokyo, japan. Comparative Immunology, Microbiology and Infectious Diseases 2021;77:101668. doi:10.1016/j.cimid.2021.101668.
- King S, Rajko-Nenow P, Ashby M, Frost L, Carpenter S, Batten C. Out- break of african horse sickness in thailand, 2020. Transboundary and Emerg- ing Diseases 2020;doi:10.1111/tbed.13701.
- Lu G, Pan J, Ou J, Shao R, Hu X, Wang C, Li S. African horse sickness: Its emergence in thailand and potential threat to other asian countries. Transboundary and Emerging Diseases 2020;doi:10.1111/tbed.13625.
- Castillo-Olivares J. African horse sickness in thailand: Challenges of controlling an outbreak by vaccination. Equine Veterinary Journal 2020;53(1):9–14. doi:10.1111/evj.13353.

Taken as a whole, this cannot be explained by a virus. Indeed it would require the "virus" to spread to birds across the world simultaneously which seems quite implausible. A more plausible mechanism is an environmental stressor caused by a planetary change in the environment.

Minor text issues:
 (page 8): the sentence "where the bra-ket notation" could be misleading because it induces associations with quantum mechanical terms, inner products, etc. Here angle brackets denote simply standard statistical averaging.
 Noted, this will be changed to a simple expectation symbol "E[.]"

(page 9): what does it mean "planetary perturbation"?
This is the postulated cause of the environmental stressor. We can elaborate on this in the manuscript.

Reviewer 2:
-----------
This review is difficult to take seriously as it seems the reviewer has not understood the concepts introduced in the paper, as exemplified by
1. A lack of a reasoned discussion of why the paper should not be published.
2. The use of words like "esoteric". This word implied a something "mysterious" that cannot be explained, even though thorough explanations and even examples in section 3 are given.
3. The use of the word "kind of" implies a lack of understanding.
4. The reference given as Rodrigues et al. does not address the reason why Gompertz is observed, but merely shows that the observations can be fit with such a model, as has been cited by multiple papers in the manuscript. This reference in particular is a far cry from explaining the mechanistic underpinnings of the Gompertz model including its unification of the communicable disease paradigm based on a network of infected and susceptible people, thus obviating the statement "The present manuscript does not have an additional value." which also seems to confuse the reviewers value system with that of the general scientific community.
5. No discussion of the technical aspects of the manuscript.

Taken together it is clear that reviewer 2 did not engage with the material at the sufficient level to understand it.

To the Editor
-------------
I thank you for considering this manuscript for publication, but I must say that the quality of the reviews was not what I expected. This paper has the potential to have large ramifications in the scientific community as it not only postulates an alternative interpretation of observations with regards to covid, but also offers an alternative theory for Gompertzian growth in general and will thus affect our understanding of cancer growth, plant growth, etc.

I ask if you could suggest this paper to a few other reviewers that will hopefully engage more authentically with the material. Especially reviewer 2 does not seem to have the level of competence for this kind of job. Reviewer 1 (which I suspect is Postnikov) has some reasoned points, but are based on a misunderstanding in my view, and thus render me without tangible corrections I can make to improve the delivery of the manuscript's message.

### 2022-07-14 elsarticle/elsarticle.tex

@@
 
-This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
+This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, and are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
 What is causing such a discrepancy between reality and the current theory of communicable diseases \citep{castro2020turning}?
 
-One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}.
-How would $\phi(t)$ have to behave?
-Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
+We start by reviewing a basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
+First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
 \begin{align}
@@
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
- Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
+ Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries, particularly with an exponentially decreasing growth rate.
 In fact, Wang et al.
-\cite{wang2012richards} showed that values of $\theta<1$ are not biologically realistic under a similar SIR model as the one given above.
+\cite{wang2012richards} showed that any values of $\theta<1$ are not physically realistic under a similar SIR model as the one given above.
 
  Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
- On the other hand, an infection of peripheral nodes should cause immediate exponential growth. 
  Herrmann and Schwartz \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve.
+Estrada and Bartesaghi \cite{estrada2022networked} found that the Gompertz can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
  Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
@@
 
-\subsection{The SIR model family is almost Gompertzian}
-Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain Gompertzian growth from the communicable disease models, i.e.
-a straight line under a double-log transform of Eq.
-\ref{eq:gomp_solution}, viz.
+\subsection{The extended SIR model family is almost Gompertzian}
+Without the linearization and the $\theta$-parametrization of the Richard's model, one can still almost obtain Gompertzian growth from the communicable disease models.
+Under a double-log transform of Eq.
+\ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
 \begin{equation}
@@
 This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$.
-However, when only modeling the diseased, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.
+However, when only modeling the diseased as we do here, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.
 
-Note here that small values of D follow logistic growth, viz.
+In the beginning of an epidemic one can assume small values of D and thus the SIRD model follows logistic growth with
 \begin{equation}
@@
 \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
-The SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
+Both the SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
 
-With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig.
+Without this approximation, the full 3-parameters in Eq.
+\ref{eq:SIRD_D} of the diseased group can show a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig.
 \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
@@
 However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.
+
+Other more general observations that conflict with these communicable disease models is 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between intergenerational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
 We are thus prompted to look for another model which can explain the observed mortality patterns.
@@
 An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.
-This could explain the immediate dampening of mortality growth observed not explainable by a SIR model. 
-It could also explain the lack of correlation between population density and mortality (or infection) rates observed by some \citep{Hamidi2020,Hamidi2020a,Carozzi2020,Arpino2020,khavarian2021high,barak2021urban}, in addition to the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021,NINA2020}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
 

### 2022-07-15 elsarticle/elsarticle.tex
add finite term theory

@@
 \modulolinenumbers[5]
-
 \journal{Chaos, Solitons \& Fractals}
@@
 Third, we show that the population acts like a coherent organism under growth/depletion. 
-Finally, we connect the Susceptible-Infected-Recovered (SIR) model with our new theory and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+Fourth, we connect the Susceptible-Infected-Recovered (SIR) model with our new theory and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+We thus connect logistic and Gompertzian growth through a mechanism of higher-order interactions expressed in a closed form.
 \end{abstract}
@@
 
-Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly.
+The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly.
 Rather, all entities are connected and communicating instantaneously as shown by Mombach et al.
@@
 If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
-An interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
+An interpretation of $\alpha$ going to $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
 One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq.
@@
 \end{equation}
+
+Finally we are positioned to arrive at a closed form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
+Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with
+
+\begin{equation}
+\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
+\end{equation}
+Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
+\begin{eqnarray*}
+  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
+  &=&\int_s \frac{1-s^N}{1-s} ds\\
+  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
+\end{eqnarray*}
+where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. Putting this all together, the modified logistic function emerges as
+ \begin{equation}
+ \label{eq:modLogistic}
+\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} -\beta \ln{x(t)}.
+ \end{equation}
+ Note that for $N=1$, this becomes the standard logistic function, while higher order terms will move closer and closer to the Gompertz curve as the hypergerometric term goes to zero.
+This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig.
+\ref{fig:hypergeometric}).
+
+\insertPdfFig{hypergeometric}{Comparison relative growth rate in the Richard's $\theta$-logistic model (Eq.
+\ref{eq:rypdal}), Gompertz model (Eq.
+\ref{eq:medianGomp}), and the augmented causality-reversed SIR model (Eq.
+\ref{eq:modLogistic}), all as a function of the abundance variable.
+Both dependent and independent variables are normalized so that final abundance (or size) is set to unity.
+The Richard's $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}
 

### 2022-07-18 elsarticle/elsarticle.tex
update intro

@@
 This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, and are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
-What is causing such a discrepancy between reality and the current theory of communicable diseases \citep{castro2020turning}?
+Other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
+
+What is causing such a discrepancy between these observations and the current theory of communicable diseases \citep{castro2020turning}?
 
@@
 However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.
-
-Other more general observations that conflict with these communicable disease models is 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between intergenerational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
 We are thus prompted to look for another model which can explain the observed mortality patterns.

### 2022-07-21 elsarticle/elsarticle.tex
add some to the introduction

@@
 Third, we show that the population acts like a coherent organism under growth/depletion. 
-Fourth, we connect the Susceptible-Infected-Recovered (SIR) model with our new theory and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
-We thus connect logistic and Gompertzian growth through a mechanism of higher-order interactions expressed in a closed form.
+Fourth, we connect the Susceptible-Infected-Recovered (SIR) model with the Gompertz curve by augmenting the logistic growth equation and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+We thus find a novel parametrization bridging logistic and Gompertzian growth.
 \end{abstract}
@@
 
-This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, and are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
-Other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
-
+This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
+These observations are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
 What is causing such a discrepancy between these observations and the current theory of communicable diseases \citep{castro2020turning}?
 
-We start by reviewing a basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
+In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
+Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same time frame.
+While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model.
+We will then present a novel bridge between the these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.
+
+\section{Simple communicable disease models}
+We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
 First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
@@
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, Richard's parameter, $\theta$, related although not identical to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 

### 2022-07-30 elsarticle/elsarticle.tex
small corrections1

@@
 Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
-We show that the Gompertz curve is incompatible with the traditional communicable disease spreading hypothesis, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor.
+We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor.
 Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
-Third, we show that the population acts like a coherent organism under growth/depletion. 
-Fourth, we connect the Susceptible-Infected-Recovered (SIR) model with the Gompertz curve by augmenting the logistic growth equation and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
-We thus find a novel parametrization bridging logistic and Gompertzian growth.
+Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+Crucially, this augementation must be accompanied by causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people.
+We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
 \end{abstract}
@@
 In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
-Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same time frame.
+Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same timeframe.
 While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model.
-We will then present a novel bridge between the these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.
+We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.
 
-\section{Simple communicable disease models}
+\subsection{Simple communicable disease models}
 We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
@@
 
-A line of reasoning employed by Rypdal and Rypdal \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
+A line of reasoning employed by Rypdal and Rypdal (2020) \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
 both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, are much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
@@
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
- Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries, particularly with an exponentially decreasing growth rate.
+ Rypdal and Rypdal (2020) \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries, particularly with an exponentially decreasing growth rate.
 In fact, Wang et al.
-\cite{wang2012richards} showed that any values of $\theta<1$ are not physically realistic under a similar SIR model as the one given above.
+(2012) \cite{wang2012richards} showed that any values of $\theta<1$ are not physically realistic under a similar SIR model as the one given above.
 
  Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
- Herrmann and Schwartz \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve.
-Estrada and Bartesaghi \cite{estrada2022networked} found that the Gompertz can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
+ Herrmann and Schwartz (2020) \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve.
+Estrada and Bartesaghi (2022) \cite{estrada2022networked} found that the Gompertz can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
  Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
@@
 To show this, we follow Carletti et al.
-\cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
+(2020) \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
 This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
@@
 Rather, all entities are connected and communicating instantaneously as shown by Mombach et al.
-\cite{mombach2002mean}.
+(2002) \cite{mombach2002mean}.
 However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.
@@
 
-Inspired by De Lauro et al. \cite{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
+Inspired by De Lauro et al. (2014) \cite{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
 As with all multiplicative processes, it is convenient to log-transform mortality and work in a dimensionless space such that $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
@@
 As argued by De Lauro et al.
-\cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation.
+(2014) \cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation.
 This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
@@
 \end{equation}
-Still, no Gompertz curve will emerge at the onset of the growth process. Estrada and Bartesaghi \cite{estrada2022networked} provide further analysis on this topic.
+Still, no Gompertz curve will emerge at the onset of the growth process. Estrada and Bartesaghi (2022) \cite{estrada2022networked} provide further analysis on this topic.
 
@@
 
-Under this model, correlation between entities is present at all orders in the original domain and all the nodes in the network communicate instantaneously\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - M(t))} = -\beta$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - M(t) = \exp{(-\beta t)}$ \citep{shklovskii2005simple}.
-In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.
+Under this model, correlation between entities is present at all orders in the original domain and all the nodes in the network communicate instantaneously.
 
@@
 This feature of $\theta$ was also obtained by Petroni et al.
-\cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
+(2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
 \ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
-Molski and Konarski \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
+Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
 This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
-In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber \cite{glauber1963coherent} in the context of electromagnetic fields.
+In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields.
 The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.
@@
  \label{eq:modLogistic}
-\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} -\beta \ln{x(t)}.
+\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)}.
  \end{equation}
@@
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter, $\theta$, related although not identical to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, Richard's parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 

### 2022-07-30 letterToEditor/letter.tex
small corrections1

@@
 Some of them are listed in the references of the manuscript.
+Due to the conclusions of our paper, I request that this paper receives more reviews than your average, and at least 3 reviews.
 
@@
 Has written on the SIR model and chaotic dynamics.
-\item Gabor Vattay, vattay@elte.hu, Eötvös Loránd University Budapest, physics, SIR models
-Has written a paper on SIR models showing that there are ``initial transients'' in the observations which are explained by us in the Gompertz model.
-\item E. B. Postnikov, postnikov@kursksu.ru, Kursk State Univesity, physics, SIR models.
-Has done work on SIR models using a mean field approach, but has a broach research interest that could be a benefit when reviewing our paper.
-\item L. Kocarev, lkocarev@manu.edu.mk, Macedonian Academy of Sciences and Arts
-, physics and mathematics, SIR models.
-Has written papers on SIR models as they relate to non-markovian models.
 \item Constantino Tsallis, tsallis@cbpf.br, Complexity Science Hub Vienna, Austria.

### 2022-07-30 elsarticle/elsarticle.tex
change plot legend

@@
 where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$.
-This parametrization is the commonly used Richard's growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \citep{wu2020generalized}.
+This parametrization is the commonly used Richards growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \citep{wu2020generalized}.
 Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}.
@@
 \subsection{The extended SIR model family is almost Gompertzian}
-Without the linearization and the $\theta$-parametrization of the Richard's model, one can still almost obtain Gompertzian growth from the communicable disease models.
+Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models.
 Under a double-log transform of Eq.
@@
 Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
-We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
+We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
 This feature of $\theta$ was also obtained by Petroni et al.
@@
 
-\insertPdfFig{hypergeometric}{Comparison relative growth rate in the Richard's $\theta$-logistic model (Eq.
+\insertPdfFig{hypergeometric}{Comparison relative growth rate in the Richards $\theta$-logistic model (Eq.
 \ref{eq:rypdal}), Gompertz model (Eq.
@@
 Both dependent and independent variables are normalized so that final abundance (or size) is set to unity.
-The Richard's $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}
+The Richards $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}
 
@@
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 

### 2022-07-30 ScientificReports/main.tex
add sciRep manuscript

--- 2022-07-30 elsarticle/elsarticle.tex
@@
-\documentclass[review]{elsarticle}
-
-\usepackage{lineno,hyperref}
-\usepackage{amsmath}
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{float}
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{graphicx}
-\modulolinenumbers[5]
-\journal{Chaos, Solitons \& Fractals}
-
-%%%%%%%%%%%%%%%%%%%%%%%
-%% Elsevier bibliography styles
-%%%%%%%%%%%%%%%%%%%%%%%
-%% To change the style, put a % in front of the second line of the current style and
-%% remove the % from the second line of the style you would like to use.
-%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Numbered
-% \bibliographystyle{model1-num-names}
-
-%% Numbered without titles
-%\bibliographystyle{model1a-num-names}
-
-%% Harvard
-%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
-
-%% Vancouver numbered
-%\usepackage{numcompress}\bibliographystyle{model3-num-names}
-
-%% Vancouver name/year
-%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
-
-%% APA style
-% \bibliographystyle{model5-names}\biboptions{authoryear}
-
-%% AMA style
-\usepackage{numcompress}\bibliographystyle{model6-num-names}
-
-%% `Elsevier LaTeX' style
-\bibliographystyle{elsarticle-num}
-%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\insertPdfFig}[3]{
-  \begin{figure}[H]
-  \centering
-  \includegraphics[width=12.4cm]{#1.pdf}
-  \caption{#2}
-  \label{fig:#1}
-  \end{figure}
-}
-
+\documentclass[fleqn,10pt]{wlscirep}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \newcommand{\insertSmallPdfFig}[3]{
@@
 }
+\newcommand{\insertPdfFig}[3]{
+  \begin{figure}[h]
+  \centering
+  \includegraphics[width=12.4cm]{#1.pdf}
+  \caption{#2}
+  \label{fig:#1}
+  \end{figure}
+}
 \graphicspath{{../figures/}}
-\begin{document}
-
-\begin{frontmatter}
 
@@
 
-%% Group authors per affiliation:
-% \author{Matz Andreas Haugen\fnref{myfootnote}}
-% \address{Enerhauggata 1, 0651 Oslo, Norway}
-% \fntext[myfootnote]{Since 1880.}
+\author[1,*]{Matz Haugen}
+\author[2]{Dorothea Gilbert}
+\affil[1]{Abel X, Oslo, Norway}
+\affil[2]{Independent Researcher, Oslo, Norway}
 
-%% or include affiliations in footnotes:
-\author[mymainaddress, mysecondaryaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
-\cortext[mycorrespondingauthor]{Corresponding author}
-% \ead[url]{www.matzhaugen.com}
-\ead{matzhaugen@gmail.com}
+\affil[*]{matzhaugen@gmail.com}
 
-\author[mysecondaryaddress]{Dorothea Gilbert}
-
-\address[mymainaddress]{Enerhauggata 1, 0651, Oslo, Norway}
-\address[mysecondaryaddress]{Independent Researcher, Oslo, Norway}
+%\keywords{Keyword1, Keyword2, Keyword3}
 
@@
 \end{abstract}
+\begin{document}
 
-\begin{keyword}
-Gompertz \sep Coherence \sep Covid \sep Coronavirus \sep Network Analysis \sep Stochastic
-\end{keyword}
+\flushbottom
+\maketitle
+% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
+%
+%  Click the title above to edit the author information and abstract
+%
+\thispagestyle{empty}
 
-\end{frontmatter}
-
-\linenumbers
-
-\section{Introduction}
+\section*{Introduction}
 
 Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
-Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
-A large body of research reveals how these models fit the recent mortality seen due to the Coronavirus epidemic \citep{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
+Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
+A large body of research reveals how these models fit the recent mortality seen due to the Coronavirus epidemic \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
 
-However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825,bajzer1997mathematical} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
+However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \cite{Gompertz1825,bajzer1997mathematical} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
 \begin{equation}
@@
 
-This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
+This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
 These observations are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
-What is causing such a discrepancy between these observations and the current theory of communicable diseases \citep{castro2020turning}?
+What is causing such a discrepancy between these observations and the current theory of communicable diseases \cite{castro2020turning}?
 
-In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
+In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
 Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same timeframe.
@@
 
-\subsection{Simple communicable disease models}
-We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
+\subsection*{Simple communicable disease models}
+We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \cite{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
 First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
@@
 where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$.
-This parametrization is the commonly used Richards growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \citep{wu2020generalized}.
-Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}.
+This parametrization is the commonly used Richards growth curve \cite{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \cite{wu2020generalized}.
+Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \cite{petroni2020logistic}.
 Note that at $\theta=1$, the traditional logistic growth curve is obtained, while at $\theta\rightarrow \infty$ we recover the exponential (Malthusian) explosion.
@@
 
-\subsection{The extended SIR model family is almost Gompertzian}
+\subsection*{The extended SIR model family is almost Gompertzian}
 Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models.
@@
 \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
-Both the SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
+Both the SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \cite{vattay2020forecasting}.
 
@@
 \end{equation}
-Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \citep{roda2020difficult}.
+Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \cite{roda2020difficult}.
 
-The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly.
+The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly.
 Rather, all entities are connected and communicating instantaneously as shown by Mombach et al.
@@
 Fitting is done with Python-Scipy and a 6-day moving average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}
 
-\section{An alternative theory for observed mortality}
+\section*{An alternative theory for observed mortality}
 \label{seq:alt}
@@
 
-This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}, where $Z(t)$ is Gaussian.
-It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}.
+This perturbation model is called an Ornstein–Uhlenbeck process \cite{risken1996fokker}, where $Z(t)$ is Gaussian.
+It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \cite{akaike1970statistical}.
 As argued by De Lauro et al.
@@
 
-Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}.
-Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
+Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \cite{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \cite{zhang1994log}.
+Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \cite{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \cite{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
 Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (Eq.
@@
 
-Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
+Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \cite{zhang1994log}.
 We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
@@
 Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
-This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
+This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
 In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields.
@@
 
-\subsection{Unification through a modified SIR model}
+\subsection*{Unification through a modified SIR model}
 It is also possible to reconcile the SIR model with the Gompertz curve. 
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
@@
 The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders.
-From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \citep{molski2003coherent}.
+From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \cite{molski2003coherent}.
 
 Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles.
-The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology \citep{lambert2013quantum}.
+The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology \cite{lambert2013quantum}.
 One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
 
-\section{Methods}
+\section*{Methods}
 
-\subsection{Estimating model parameters}
+\subsection*{Estimating model parameters}
 The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text (see SI Appendix for details).
 Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths, similar to the technique by Carletti et al.
-\citep{carletti2020covid}.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
+\cite{carletti2020covid}.
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.
 
@@
 The transmission rate is modeled as the product of $\eta \xi$ in Eq.
-\ref{eq:SIRD_D}, and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
+\ref{eq:SIRD_D}, and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \cite{nadarajah2016distribution},
 \begin{equation}
@@
 
-\section{Competing interests}
+\section*{Competing interests}
 The authors declare no competing interests.
 
-\section{Author Contributions}
+\section*{Author Contributions}
 M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
-
-% \section{Supplement}
-
-% \insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
-
-% \begin{table}[H]
-% \label{tab:sird}
-% \centering
-% \caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}.
-Standard deviation is given in the paranthesis corresponding to the last significant digit of the reported number.}
-% \begin{tabular}{llll}
-% \toprule
-% {} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
-% \midrule
-% Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
-% Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
-% Austria        &     8(1) &    36(5) &    11(1) \\
-% Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
-% Sweden         &    19(5) &     8(2) &     3(1) \\
-% Italy          &   137(3) &  1.77(5) &  3.71(1) \\
-% France         &  240(10) &  1.24(7) &   8.1(4) \\
-% Spain          &   145(2) &  2.27(5) &  4.75(8) \\
-% Germany        &    61(4) &   3.9(3) &   6.5(5) \\
-% Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
-% United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
-% Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
-% Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
-% Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
-% US             &   312(5) &  0.76(2) &  1.83(7) \\
-% China          &  130(30) &   3.5(5) &    26(4) \\
-% \bottomrule
-% \end{tabular}
-
-% \end{table}
-
-% \begin{table}[H]
-% \label{tab:gomp}
-% \centering
-% \caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.
-Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
-% \begin{tabular}{llll}
-% \toprule
-% {} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
-% \midrule
-% Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
-% Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
-% Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
-% Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
-% Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
-% Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
-% France         &  10.2(1) &   8.2(1) &  11.0(8) \\
-% Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
-% Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
-% Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
-% United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
-% Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
-% Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
-% Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
-% US             &  7.53(7) &  5.32(7) &   8.4(5) \\
-% China          &   9.8(1) &   9.4(2) &    10(1) \\
-% \bottomrule
-% \end{tabular}
-
-% \end{table}
-
-% \section{References}
 

### 2022-07-30 arxiv/main.tex
add sciRep manuscript

--- 2022-05-19 main.tex
@@
 % \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
-\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
+\title{Unifying the communicable disease spreading paradigm with Gompertzian growth}
 \graphicspath{{figures/}}
@@
 \section{Author contributions}
-X conceived the idea and wrote the manuscript while Y advised.
+X and Y conceived the idea and X wrote the manuscript.
 
@@
 \end{tabular}
-
 \end{table}

### 2022-08-03 ScientificReports/main.tex
finalize for Sci Rep submission

@@
 Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
-Crucially, this augementation must be accompanied by causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people.
+Crucially, this augmentation must be accompanied by a causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people.
 We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
@@
 In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
-Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same timeframe.
+Many of these animal mortality spikes are not attributable to the Coronavirus, but the question still remains why these spikes all occurred within the same timeframe.
 While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model.
@@
 \end{equation} 
-If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
+If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model's infection term becomes 
 \begin{equation}
@@
 
-Finally we are positioned to arrive at a closed form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
+Finally we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
 Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with
@@
 \end{eqnarray*}
-where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. Putting this all together, the modified logistic function emerges as
+where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
  \begin{equation}

### 2022-08-04 ScientificReports/main.tex
respond to Thea's comments

@@
 In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
-Many of these animal mortality spikes are not attributable to the Coronavirus, but the question still remains why these spikes all occurred within the same timeframe.
-While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model.
-We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.
+Many of these animal mortality spikes are not attributable to a Coronavirus infection, but the question still remains why these spikes all occurred within the same time frame.
+While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model.
+This alternative theory represents an alternative paradigm in disease spreading dynamics.
+We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model as an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.
 
 \subsection*{Simple communicable disease models}
-We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \cite{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
+We start by reviewing the basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \cite{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
 First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
@@
 and where the number of recovered, $R(t)$, is under this linearization decoupled from the other variables.
+They further assume the number of deceased is proportional to the number of cumulative infected, offset by a time lag, which allows us to use the same set of linearized equations to model the number of cumulative people deceased without loss of generality.
 
-They further assume the number of diseased is proportional to the number of cumulative infected, offset by a time lag, which allows us to use the same set of linearized equations to model the number of cumulative people diseased without loss of generality.
-
- Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
+ Due to this linearization, the infection/recovery rate ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
 And since $I$ is a function of $Y$, we can combine Eq.
@@
  Herrmann and Schwartz (2020) \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve.
-Estrada and Bartesaghi (2022) \cite{estrada2022networked} found that the Gompertz can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
+Estrada and Bartesaghi (2022) \cite{estrada2022networked} found that the Gompertz curve can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
  Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
@@
 Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models.
-Under a double-log transform of Eq.
+Under a double-log-transform of Eq.
 \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
@@
 To show this, we follow Carletti et al.
-(2020) \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
+(2020) \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
+This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 
-Algebraic manipulations reveal that the diseased group is described by a single equation, viz.
+Algebraic manipulations reveal that the deceased group is described by a single equation, viz.
 \begin{equation}
@@
 This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$.
-However, when only modeling the diseased as we do here, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.
+However, when only modeling the deceased as we do here, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.
 
@@
 \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
-Both the SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \cite{vattay2020forecasting}.
+Both the SIR and SEIR model exhibit similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \cite{vattay2020forecasting}.
 
 Without this approximation, the full 3-parameters in Eq.
-\ref{eq:SIRD_D} of the diseased group can show a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig.
+\ref{eq:SIRD_D} of the deceased group can show a curve quite close to a straight line in the double-log-domain even in the initial observations, although there will always be a non-zero concavity (Fig.
 \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
 Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq.
-\ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
+\ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality \cite{Levitt2020},
 \begin{equation}
@@
 (2002) \cite{mombach2002mean}.
-However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.
+However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log-domain.
 We are thus prompted to look for another model which can explain the observed mortality patterns.
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
 Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text.
@@
 \label{seq:alt}
-An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.
+An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.
 
@@
 
-Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \cite{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \cite{zhang1994log}.
+Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a perturbation of the biosphere, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \cite{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \cite{zhang1994log}.
 Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \cite{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \cite{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
@@
 
-In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in a SIR model, but also as a linear interaction term between the two macroscopic states. 
+In the logistic model, we recognize the rightmost side of the logistic equation as the transmission term in an SIR model, but also as a linear interaction term between the two macroscopic states. 
 As mentioned earlier, this procedure is a mean field approximation with an implied average interaction between the variables. 
@@
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
-Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
+Then, we reverse the causality where the population of infected is now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
 \begin{equation}
@@
 \end{equation}
-As we are interested in the aggregate macroscopic behavior, we take averages in the log domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
+As we are interested in the aggregate macroscopic behavior, we take averages in the log-domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
 \begin{equation}
@@
 If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
-An interpretation of $\alpha$ going to $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
+An interpretation of $\alpha$ approaching $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
 One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq.
@@
  \end{equation}
- Note that for $N=1$, this becomes the standard logistic function, while higher order terms will move closer and closer to the Gompertz curve as the hypergerometric term goes to zero.
+ Note that for $N=1$, this relative growth rate becomes the standard logistic function, while higher order terms will move closer and closer to the Gompertz curve as the hypergerometric term approaches zero.
 This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig.
@@
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in an SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
-We further show that the observed mortality across countries can be explained by a model where the biological system is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which gradual return to pre-epidemic conditions are mediated.
+We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
 The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders.

### 2022-08-05 ScientificReports/main.tex
use githash for data table

@@
 
-This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
-These observations are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
+This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, a conclusion made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
 What is causing such a discrepancy between these observations and the current theory of communicable diseases \cite{castro2020turning}?
@@
 
+\subsection*{Data Availability}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore,
+USA \cite{dong2020interactive}, and are available at \url{https://github.com/CSSEGISandData/COVID-19}, specifically the table \newline\verb|csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv|, \newline downloaded Aug.
+$5^{th}$ 2022.
+
 \section*{Competing interests}

### 2022-09-19 NonlinearDynamics/sn-article.tex
add nonlinear dynamics draft

--- 2022-08-05 ScientificReports/main.tex
@@
-\documentclass[fleqn,10pt]{wlscirep}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing
+
+%%=======================================================%%
+%% to print line numbers in the margin use lineno option %%
+%%=======================================================%%
+
+%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+
+%%======================================================%%
+%% to compile with pdflatex/xelatex use pdflatex option %%
+%%======================================================%%
+
+%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+
+%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+\documentclass[sn-mathphys]{sn-jnl}% Math and Physical Sciences Reference Style
+%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
+%%\documentclass[sn-vancouver]{sn-jnl}% Vancouver Reference Style
+%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style
+%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
+%%\documentclass[sn-standardnature]{sn-jnl}% Standard Nature Portfolio Reference Style
+%%\documentclass[default]{sn-jnl}% Default
+%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout
+
+%%%% Standard Packages
+%%<additional latex packages if required can be included here>
+%%%%
+
+%%%%%=============================================================================%%%%
+%%%%  Remarks: This template is provided to aid authors with the preparation
+%%%%  of original research articles intended for submission to journals published 
+%%%%  by Springer Nature. The guidance has been prepared in partnership with 
+%%%%  production teams to conform to Springer Nature technical requirements. 
+%%%%  Editorial and presentation requirements differ among journal portfolios and 
+%%%%  research disciplines. You may find sections in this template are irrelevant 
+%%%%  to your work and are empowered to omit any such section if allowed by the 
+%%%%  journal you intend to submit to. The submission guidelines and policies 
+%%%%  of the journal take precedence. A detailed User Manual is available in the 
+%%%%  template package for technical guidance.
+%%%%%=============================================================================%%%%
+
 \newcommand{\insertSmallPdfFig}[3]{
@@
 }
-\graphicspath{{../figures/}}
+% \graphicspath{{../figures/}}
 
-\title{Unifying the communicable disease spreading paradigm with Gompertzian growth}
+\jyear{2021}%
 
-\author[1,*]{Matz Haugen}
-\author[2]{Dorothea Gilbert}
-\affil[1]{Abel X, Oslo, Norway}
-\affil[2]{Independent Researcher, Oslo, Norway}
+%% as per the requirement new theorem styles can be included as shown below
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}% 
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
 
-\affil[*]{matzhaugen@gmail.com}
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
 
-%\keywords{Keyword1, Keyword2, Keyword3}
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}%
 
-\begin{abstract}
-Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
+\raggedbottom
+%%\unnumbered% uncomment this for unnumbered level heads
+
+\begin{document}
+
+\title[Unifying the communicable disease spreading paradigm]{Unifying the communicable disease spreading paradigm with Gompertzian growth}
+
+%%=============================================================%%
+%% Prefix	-> \pfx{Dr}
+%% GivenName	-> \fnm{Joergen W.}
+%% Particle	-> \spfx{van der} -> surname prefix
+%% FamilyName	-> \sur{Ploeg}
+%% Suffix	-> \sfx{IV}
+%% NatureName	-> \tanm{Poet Laureate} -> Title after name
+%% Degrees	-> \dgr{MSc, PhD}
+%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
+%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
+%%=============================================================%%
+
+\author*[1]{\fnm{Matz} \sur{Haugen}}\email{matzhaugen@gmail.com}
+
+\author[2]{\fnm{Dorothea} \sur{Gilbert}}\email{D.Thea@protonmail.ch}
+% \equalcont{These authors contributed equally to this work.}
+
+\affil*[1]{\orgdiv{}, \orgname{Abel X}, \orgaddress{\city{Oslo}, \postcode{1360}, \state{Baerum}, \country{Norway}}}
+
+\affil[2]{\orgname{Independent Researcher}, \orgaddress{\city{Oslo}, \country{Norway}}}
+
+% \affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}
+
+%%==================================%%
+%% sample for unstructured abstract %%
+%%==================================%%
+
+\abstract{Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
 We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor.
@@
 We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
-\end{abstract}
-\begin{document}
+}
 
-\flushbottom
+%%================================%%
+%% Sample for structured abstract %%
+%%================================%%
+
+% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}
+
+\keywords{Gompertz, Communicable Diseases, Network, Stochastic Processes}
+
+%%\pacs[JEL Classification]{D8, H51}
+
+%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}
+
 \maketitle
-% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
-%
-%  Click the title above to edit the author information and abstract
-%
-\thispagestyle{empty}
 
-\section*{Introduction}
+\section{Introduction}\label{sec1}
 
@@
 
-\subsection*{Code Availability}
+\backmatter
+
+\bmhead{Supplementary information}
+
+A supplement is available for this article.
+% \bmhead{Acknowledgments}
+
+% \section*{Declarations}
+
+% Some journals require declarations to be submitted in a standardised format.
+Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section.
+If yes, your manuscript must contain the following sections under the heading `Declarations':
+
+\section*{Competing interests}
+The authors have no relevant financial or non-financial interests to disclose.”
+
+\section*{Author Contributions}
+M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
+
+\section*{Data Availability}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore,
+USA \cite{dong2020interactive}, and are available at \url{https://github.com/CSSEGISandData/COVID-19}, specifically the table \verb|time_series_covid19_deaths_global.csv|, downloaded Aug.
+$5^{th}$ 2022.
+
+\section*{Code Availability}
 
@@
 
-\subsection*{Data Availability}
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore,
-USA \cite{dong2020interactive}, and are available at \url{https://github.com/CSSEGISandData/COVID-19}, specifically the table \newline\verb|csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv|, \newline downloaded Aug.
-$5^{th}$ 2022.
-
-\section*{Competing interests}
-The authors declare no competing interests.
+\section*{Funding}
+The authors declare that no funds, grants, or other support were received during the preparation of this manuscript.
 
@@
 
+%%===================================================%%
+%% For presentation purpose, we have included        %%
+%% \bigskip command. please ignore this.             %%
+%%===================================================%%
+% \bigskip
+% \begin{flushleft}%
+% Editorial Policies for:
+
+% \bigskip\noindent
+% Springer journals and proceedings: \url{https://www.springer.com/gp/editorial-policies}
+
+% \bigskip\noindent
+% Nature Portfolio journals: \url{https://www.nature.com/nature-research/editorial-policies}
+
+% \bigskip\noindent
+% \textit{Scientific Reports}: \url{https://www.nature.com/srep/journal-policies/editorial-policies}
+
+% \bigskip\noindent
+% BMC journals: \url{https://www.biomedcentral.com/getpublished/editorial-policies}
+% \end{flushleft}
+
+% \begin{appendices}
+
+% \section{Section title of first appendix}\label{secA1}
+
+% An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.
+
+% %%=============================================%%
+% %% For submissions to Nature Portfolio Journals %%
+% %% please use the heading ``Extended Data''.   %%
+% %%=============================================%%
+
+% %%=============================================================%%
+% %% Sample for another appendix section			       %%
+% %%=============================================================%%
+
+% %% \section{Example of another appendix section}\label{secA2}%
+% %% Appendices may be used for helpful, supporting or essential material that would otherwise 
+% %% clutter, break up or be distracting to the text. Appendices can consist of sections, figures, 
+% %% tables and equations etc.
+
+% \end{appendices}
+
+%%===========================================================================================%%
+%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
+%% system, please include the references within the manuscript file itself. You may do this  %%
+%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
+%% file, and delete the associated \verb+\bibliography+ commands.                            %%
+%%===========================================================================================%%
+
 \bibliography{../references2}
+%% if required, the content of .bbl file can be included here once bbl is generated
+%%\input sn-article.bbl
+
+%% Default %%
+%%\input sn-sample-bib.tex%
 

### 2022-11-09 ScientificReports/main.tex
minor changes to sci per

@@
 \begin{abstract}
-Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
+A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
 We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor.
@@
 Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
-A large body of research reveals how these models fit the recent mortality seen due to the Coronavirus epidemic \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
+A large body of research reveals how these models fit the mortality seen due to the Coronavirus epidemic in 2020 \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
 
@@
 In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
-Many of these animal mortality spikes are not attributable to a Coronavirus infection, but the question still remains why these spikes all occurred within the same time frame.
+Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame?
 While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model.
@@
 
- At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
+ At this limit, the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
  The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning.
@@
 To show this, we follow Carletti et al.
-(2020) \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
+(2020) \cite{carletti2020covid} and consider the extended version of the SIR model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
+This extension requires an addition to our original set of equations \ref{eq:SIR}, with an equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 
-This perturbation model is called an Ornstein–Uhlenbeck process \cite{risken1996fokker}, where $Z(t)$ is Gaussian.
+This perturbation model is an Ornstein–Uhlenbeck process \cite{risken1996fokker}, where $Z(t)$ is Gaussian.
 It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \cite{akaike1970statistical}.
@@
 
-In contrast, as discussed above, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
+In contrast and as discussed above, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
 A log-normal distribution of abundance implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
@@
 
-Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \cite{zhang1994log}.
+With this phenomenological argument we see that the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor and due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \cite{zhang1994log}.
 We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$.
@@
  \end{equation}
- Note that for $N=1$, this relative growth rate becomes the standard logistic function, while higher order terms will move closer and closer to the Gompertz curve as the hypergerometric term approaches zero.
+ Note that for $N=1$, this relative growth rate becomes the standard logistic function, while higher order terms will move closer to the Gompertz curve as the hypergerometric term approaches zero.
 This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig.

### 2022-11-27 arxiv/main.tex
add arxiv article

--- 2022-05-18 main.tex
@@
 \usepackage{lineno}
-\linenumbers
+% \linenumbers
 
@@
 % \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
-\title{Unifying Gompertzian growth with the communicable disease spreading paradigm}
-\graphicspath{{figures/}}
+\title{Unifying the communicable disease spreading paradigm with Gompertzian growth}
+% \graphicspath{{../figures/}}
 %\date{September 9, 1985} % Here you can change the date presented in the paper title
@@
 
-% \author{Matz A. Haugen, \\
-% Independent Reseacher, Oslo\\
-% \texttt{matzhaugen@gmail.com} \\
-%   \And
-%   Dorothea Gilbert \\
-%   University of Oslo\\
-% }
+\author{Matz A. Haugen, \\
+Independent Reseacher, Oslo\\
+\texttt{matzhaugen@gmail.com} \\
+  \And
+  Dorothea Gilbert \\
+  Independent Reseacher, Oslo\\
+}
 
@@
 \begin{abstract}
-Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
-We show that the Gompertz curve is incompatible with the traditional communicable disease spreading hypothesis, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor.
+A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
+We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor.
 Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
-Third, we show that the population acts like a coherent organism under growth/depletion. 
-Finally, we connect the Susceptible-Infected-Recovered (SIR) model with our new theory and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+Crucially, this augmentation must be accompanied by a causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people.
+We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
 \end{abstract}
@@
 
-\section{Introduction}
+\section*{Introduction}
 Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
-Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
-A large body of research has emerged that explores how these models fit the recent mortality seen due to the Coronavirus epidemic \citep{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
+Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
+A large body of research reveals how these models fit the mortality seen due to the Coronavirus epidemic in 2020 \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
 
-However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825,bajzer1997mathematical} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
+However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \cite{Gompertz1825,bajzer1997mathematical} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
 \begin{equation}
@@
 
-This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves.
-What is causing such a discrepancy between reality and the current theory of communicable diseases \citep{castro2020turning}?
+This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, a conclusion made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
+What is causing such a discrepancy between these observations and the current theory of communicable diseases \cite{castro2020turning}?
 
-One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}.
-How would $\phi(t)$ have to behave?
-Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
+In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
+Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame?
+While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model.
+This alternative theory represents an alternative paradigm in disease spreading dynamics.
+We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model as an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.
+
+\subsection*{Simple communicable disease models}
+We start by reviewing the basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \cite{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
+First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
 \begin{align}
 \label{eq:SIR}
-\frac{dS}{dt}& = -\beta \frac{IS}{N}\\
+\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
 \frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
-\frac{dR}{dt}& = \alpha I,
+\frac{dR}{dt}& = \alpha I\nonumber,
 \end{align}
@@
 
-A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
-both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
+A line of reasoning employed by Rypdal and Rypdal (2020) \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
+both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, are much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
 \begin{align}
@@
 and where the number of recovered, $R(t)$, is under this linearization decoupled from the other variables.
+They further assume the number of deceased is proportional to the number of cumulative infected, offset by a time lag, which allows us to use the same set of linearized equations to model the number of cumulative people deceased without loss of generality.
 
-They further assume the number of diseased is proportional to the number of cumulative infected, offset by a time lag, which allows us to use the same set of linearized equations to model the number of cumulative people diseased without loss of generality.
-
- Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
-And since $I$ is a function of $Y$, we can combine (\ref{eq:linSIR}) via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
+ Due to this linearization, the infection/recovery rate ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
+And since $I$ is a function of $Y$, we can combine Eq.
+\ref{eq:linSIR} via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
 \begin{subequations}
@@
 where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$.
-This parametrization is the commonly used Richard's growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \citep{wu2020generalized}.
-Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}.
+This parametrization is the commonly used Richards growth curve \cite{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \cite{wu2020generalized}.
+Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \cite{petroni2020logistic}.
 Note that at $\theta=1$, the traditional logistic growth curve is obtained, while at $\theta\rightarrow \infty$ we recover the exponential (Malthusian) explosion.
@@
 
- At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
+ At this limit, the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
  The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning.
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
- \citet{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries.
+ Rypdal and Rypdal (2020) \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries, particularly with an exponentially decreasing growth rate.
+In fact, Wang et al.
+(2012) \cite{wang2012richards} showed that any values of $\theta<1$ are not physically realistic under a similar SIR model as the one given above.
 
  Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
- On the other hand, an infection of peripheral nodes should cause immediate exponential growth. 
- \citet{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. 
+ Herrmann and Schwartz (2020) \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve.
+Estrada and Bartesaghi (2022) \cite{estrada2022networked} found that the Gompertz curve can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
  Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
@@
 
-\subsection{The SIR model family is almost Gompertzian}
-Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain Gompertzian growth from the communicable disease models, i.e.
-a straight line under a double-log transform of (\ref{eq:gomp_solution}), viz.
+\subsection*{The extended SIR model family is almost Gompertzian}
+Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models.
+Under a double-log-transform of Eq.
+\ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
 \begin{equation}
@@
 with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$.
-To show this, we follow \citet{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
-This requires an addition to our original set of equations (\ref{eq:SIR}), with a extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
+To show this, we follow Carletti et al.
+(2020) \cite{carletti2020covid} and consider the extended version of the SIR model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
+This extension requires an addition to our original set of equations \ref{eq:SIR}, with an equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
 \begin{equation}
@@
 
-Some algebraic manipulations reveal that the diseased group is described by a single equation, viz.
+Algebraic manipulations reveal that the deceased group is described by a single equation, viz.
 \begin{equation}
@@
 where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is again synonymous to the initial susceptible pool of people.
-This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$.
-When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.
+This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$.
+However, when only modeling the deceased as we do here, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.
 
-Note here that small values of D follow logistic growth, viz.
+In the beginning of an epidemic one can assume small values of D and thus the SIRD model follows logistic growth with
 \begin{equation}
@@
 \end{equation}
-obtained with a second order expansion of the exponential term in (\ref{eq:SIRD_D}) and with $c=\eta\xi - \kappa$ and $K=2b/\eta\xi^2$.
-The SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.
+obtained with a second order expansion of the exponential term in Eq.
+\ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
+Both the SIR and SEIR model exhibit similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \cite{vattay2020forecasting}.
 
-With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig.
-\ref{fig:Gompertz_vs_SIRD_lnln_infty} and Supplemental Information).
-Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in (\ref{eq:GompertzODE}), it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
+Without this approximation, the full 3-parameters in Eq.
+\ref{eq:SIRD_D} of the deceased group can show a curve quite close to a straight line in the double-log-domain even in the initial observations, although there will always be a non-zero concavity (Fig.
+\ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
+Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq.
+\ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality \cite{Levitt2020},
 \begin{equation}
@@
 \end{equation}
-Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \citep{roda2020difficult}.
+Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \cite{roda2020difficult}.
 
-Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly.
-Rather, all entities are connected and communicating instantaneously as shown by \citet{mombach2002mean}.
-However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.
+The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly.
+Rather, all entities are connected and communicating instantaneously as shown by Mombach et al.
+(2002) \cite{mombach2002mean}.
+However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log-domain.
 We are thus prompted to look for another model which can explain the observed mortality patterns.
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GompertzODE2param}), it is here obtained using non-linear least squares to put both models on equal footing.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
+Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text.
+Although the Gompertz curve can also be obtained with a simple linear fit using Eq.
+\ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing.
 The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
@@
 Fitting is done with Python-Scipy and a 6-day moving average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}
 
-\section{An alternative theory for observed mortality}
+\section*{An alternative theory for observed mortality}
 \label{seq:alt}
-An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.
-This could explain the immediate dampening of mortality growth observed. 
-It could also explain the lack of correlation between population density and mortality (or infection) rates observed by some \citep{Hamidi2020,Hamidi2020a,Carozzi2020,Arpino2020,khavarian2021high,barak2021urban}.
+An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.
 
-Inspired by \citet{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
+Inspired by De Lauro et al. (2014) \cite{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
 As with all multiplicative processes, it is convenient to log-transform mortality and work in a dimensionless space such that $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
@@
 
-This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}.
-It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}.
-The diffusion coefficient, $\sigma$, represents the strength of the perturbation, which we directly see if we recast this equation in terms of mortality while introducing a new parameter $K=\exp{(-\frac{\sigma}{2\beta})}$, viz.
+This perturbation model is an Ornstein–Uhlenbeck process \cite{risken1996fokker}, where $Z(t)$ is Gaussian.
+It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \cite{akaike1970statistical}.
+As argued by De Lauro et al.
+(2014) \cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation.
+This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz.
 \begin{equation}
 \label{eq:microscopicOriginal}
-dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{K}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
+dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{C}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
 \end{equation}
 
-To obtain the deterministic observable, we first take the average in the log-domain (\ref{eq:microscopic}) and transform back to the original domain, 
+To obtain the deterministic observable, we first take the average in the log-domain (Eq. \ref{eq:microscopic}), 
 \begin{equation}
@@
 
-Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}.
-Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation in (\ref{eq:microscopicOriginal}) \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
-Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
+Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a perturbation of the biosphere, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \cite{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \cite{zhang1994log}.
+Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \cite{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \cite{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}.
+Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (Eq.
+\ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
 This implies that the central limit theorem applies to the log-domain.
 
-\section{Unifying the SIR model and the Gompertz model}
+\section*{Unifying the SIR model and the Gompertz model}
 The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
@@
 \end{equation}
-Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.
+Still, no Gompertz curve will emerge at the onset of the growth process. Estrada and Bartesaghi (2022) \cite{estrada2022networked} provide further analysis on this topic.
 
-In contrast, as discussed in Section \ref{seq:alt}, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
+In contrast and as discussed above, the Gompertz model is implied by a multiplicative stochastic process with a log-normal distribution in its abundance at any given point in time.
 A log-normal distribution of abundance implies that the log-domain is the natural domain in which the central limit theorem applies, thus implying correlated mortality growth through the geometric mean,
@@
 
-Under this model, correlation between entities is present at all orders in the original domain and all the nodes in the network communicate instantaneously\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - M(t))} = -\beta$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - M(t) = \exp{(-\beta t)}$ \citep{shklovskii2005simple}.
-In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.
+Under this model, correlation between entities is present at all orders in the original domain and all the nodes in the network communicate instantaneously.
 
-Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
-We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
-This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
-\citet{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
-This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
-In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields.
+With this phenomenological argument we see that the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor and due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \cite{zhang1994log}.
+We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$.
+This feature of $\theta$ was also obtained by Petroni et al.
+(2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
+\ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
+Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
+This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
+In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields.
 The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.
 
-\subsection{Unification through a modified SIR model}
-However, it is possible to reconcile the SIR model with the Gompertz curve. 
+\subsection*{Unification through a modified SIR model}
+It is also possible to reconcile the SIR model with the Gompertz curve. 
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
-Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
+Then, we reverse the causality where the population of infected is now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
 \begin{equation}
@@
 \end{equation}
-As we are interested in the aggregate macroscopic behavior, we take averages in the log domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
+As we are interested in the aggregate macroscopic behavior, we take averages in the log-domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
 \begin{equation}
-\frac{d}{{dt}} \ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} - \alpha.
+\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}\sum_i^N x_i.
 \end{equation}
 If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
-An interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
-One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain (\ref{eq:medianGomp}),
+An interpretation of $\alpha$ approaching $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above.
+One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq.
+\ref{eq:medianGomp},
 \begin{equation}
@@
 
-\section{Conclusion}
+Finally we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
+Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with
+
+\begin{equation}
+\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
+\end{equation}
+Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
+\begin{eqnarray*}
+  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
+  &=&\int_s \frac{1-s^N}{1-s} ds\\
+  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
+\end{eqnarray*}
+where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
+ \begin{equation}
+ \label{eq:modLogistic}
+\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)}.
+ \end{equation}
+ Note that for $N=1$, this relative growth rate becomes the standard logistic function, while higher order terms will move closer to the Gompertz curve as the hypergerometric term approaches zero.
+This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig.
+\ref{fig:hypergeometric}).
+
+\insertPdfFig{hypergeometric}{Comparison relative growth rate in the Richards $\theta$-logistic model (Eq.
+\ref{eq:rypdal}), Gompertz model (Eq.
+\ref{eq:medianGomp}), and the augmented causality-reversed SIR model (Eq.
+\ref{eq:modLogistic}), all as a function of the abundance variable.
+Both dependent and independent variables are normalized so that final abundance (or size) is set to unity.
+The Richards $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}
+
+\section*{Conclusion}
 In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in an SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
 
-We further show that the observed mortality across countries can be explained by a model where the biological system is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which gradual return to pre-epidemic conditions are mediated.
+We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
 The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders.
-From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \citep{molski2003coherent}.
+From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \cite{molski2003coherent}.
 
 Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles.
-The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology \citep{lambert2013quantum}.
+The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology \cite{lambert2013quantum}.
 One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
-
-\newpage
 
@@
 
-\subsection{Estimating model parameters}
-The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig.
-\ref{fig:Gompertz_vs_SIRD}).
-Fitting is done using Python-Scipy's non-linear least squares algorithm \verb|curve_fit| with specified Jacobians and a 6-day windowed average of observed deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.
+\subsection*{Estimating model parameters}
+The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text (see SI Appendix for details).
+Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths, similar to the technique by Carletti et al.
+\cite{carletti2020covid}.
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.
 
-The transmission rate is modeled as the product of $\eta \xi$ in (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
+% On statistical grounds, one could argue that the daily mortality counts follow a Poisson distribution.
+However, in a growth setting, one might expect heteroskedasticity inherent in the biological process where e.g.
+growth has less variance in the later stages, perhaps dependent on proximity to a stationary state.
+This latter hypothesis could be incorporated in the stochastic framework presented in the text, but to keep the scope of this paper confined we use a simple non-linear least squares minimization as our results are not meaningfully different compared with a more complex variance model.
+
+The transmission rate is modeled as the product of $\eta \xi$ in Eq.
+\ref{eq:SIRD_D}, and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \cite{nadarajah2016distribution},
 \begin{equation}
@@
 \newline
-\verb|https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb|
-% The computer code required to produce the results are given in the following Github repository,
-% \begin{verbatim}
-% https://github.com/matzhaugen/GompertzManuscript
-% \end{verbatim}
+\url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb}
 
-\section{Author contributions}
-X conceived the idea and wrote the manuscript while Y advised.
+\subsection*{Data Availability}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore,
+USA \cite{dong2020interactive}, and are available at \url{https://github.com/CSSEGISandData/COVID-19}, specifically the table \newline\verb|csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv|, \newline downloaded Aug.
+$5^{th}$ 2022.
 
-\section{Competing interests}
+\section*{Competing interests}
 The authors declare no competing interests.
+
+\section*{Author Contributions}
+M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
 
@@
 
-\begin{table}[H]
+\begin{table}
 \label{tab:sird}
 \centering
-\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq.
-(\ref{eq:SIRD_D}).
+\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure 1 using notation from eq.
+(12).
 Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
@@
 Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
-Austria        &     8(1) &    36(5) &    11(1) \\
+Austria        &   6.9(5) &    40(3) &   8.7(6) \\
 Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
@@
 
-\begin{table}[H]
+\begin{table}
 \label{tab:gomp}
 \centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq. (\ref{eq:GompertzODE}).}
+\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure 1 using notation from eq. (1).}
 \begin{tabular}{llll}
@@
 Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
-Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
+Austria        &  6.98(5) &   7.6(1) &   7.8(4) \\
 Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
@@
 \end{table}
+
 \bibliographystyle{bst/abbrvnat}

### 2022-11-28 arxiv/main.tex
reviewer number 3 response

@@
 
- At this limit, the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
- The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning.
+ The Gompertzian limit implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. 
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
@@
 The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly.
-Rather, all entities are connected and communicating instantaneously as shown by Mombach et al.
+Rather, all entities are connected and communicating through long-range interactions as shown by Mombach et al.
 (2002) \cite{mombach2002mean}.
@@
 An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.
+[PLACEHOLDER: This stressor could be in the form of anomalous geomagnetic storms, or man-made disturbances in the atmosphere.
+For example, the ionosphere, which is the reservoir of the earth's electromagnetic potential, could have been disturbed either by man-made or natural events, causing organisms to experience an immune response.
+The exact mechanism of such hypotheses would need further study.]
 
@@
 \ref{eq:GompertzODE} gives $\nu=\sigma/2$, suggesting that the stressor is indeed the source of growth, while $\beta$ is the growth-limiting factor.
+This comparison is clear after taking the median of Eq.
+\ref{eq:microscopicOriginal} wherein the cross-term of $F(t)dW(t)$ goes to zero due to the independence of $F(t)$ and $dW(t)$.
 We also see that by comparing equations \ref{eq:rypdalODE} and \ref{eq:rypdalLimit} with the stochastic counterpart in Eq.
@@
 \section*{Unifying the SIR model and the Gompertz model}
+\subsection*{Comparing microscopic logistic and Gompertz growth}
 The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
@@
 With this phenomenological argument we see that the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor and due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \cite{zhang1994log}.
-We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$.
-This feature of $\theta$ was also obtained by Petroni et al.
-(2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
-\ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''.
-Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
-This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
-In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields.
-The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.
 
 \subsection*{Unification through a modified SIR model}
-It is also possible to reconcile the SIR model with the Gompertz curve. 
+We can now reconcile the SIR model with the Gompertz curve. 
 Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
@@
 \section*{Conclusion}
-In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
-In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in an SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
+In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states.
+In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in an augmented SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$ and where the source of growth is driven by the number of susceptible and not the number of infected.
 
-We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
+We can now further appreciate the Richards parametrization as a transition from non-collaborative to maximally collaborative growth as $\theta \rightarrow 0$. 
+This feature of $\theta$ was also obtained by Petroni et al.
+(2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
+\ref{eq:rypdal} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$.
+Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
+This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
+In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields.
+The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.
+
+Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected, and be driven by the susceptible as opposed to the infected.
+We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.
+
+We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
 The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders.
@@
 
+%  Carletti (left) transformation to our notation (right)
+% gamma - eta
+% eta - xi
+% beta - kappa 
 \begin{table}
@@
 (12).
-Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
+Standard deviation estimates is given in the parenthesis corresponding to the last significant digit of the reported number.}
 \begin{tabular}{llll}
@@
 
-\bibliographystyle{bst/abbrvnat}
+\bibliographystyle{abbrvnat}
 \bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .

### 2022-11-28 ResponseToReviewers.txt
add response

Response to reviewer #3:

Reviewer #3: After reading the article reviewed, I came to the following conclusions: from a mathematical point of view, the model seems to be correct (unless I have overlooked a mistake), while the interpretation of the results obtained is, to put it mildly, controversial, and in some respects incorrect or requiring additional explanations. In particular:

p.4. (after Eq.7) …At this limit the relative growth rate approaches infinity as Y→0, which seems odd under the hypothesis that the pathogen has just started spreading.
Comment: The form of the growth functions used in the biological domain excludes the possibility of reaching the limiting value Y=0, which is Y(t→0)=Y0. Therefore, the growth rate will never reach an infinite value.

Answer: This is a good point and the sentence has been removed.

p.6. …all entities are connected and communicating instantaneously as shown by Mombach et al.
Comment: Mombach et al. derived the Gompertz function in the framework of the cellular mean field theory, in which the presence of long-range (slowly decaying) interactions is taken into account. In biological systems, interactions at the distance (nonlocal correlations) even when they seem instantaneous, they do not allow the superluminal communication (information transfer), which, according to the theory of relativity is impossible.

Answer: I see that the the word "instantaneous" is inaccurate in this context and have changed this paragraph to 
"The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating *through long-range interactions* as shown by Mombach et al. (2002) \cite{mombach2002mean}.
"

The idea was to quote the extra assumptions that are present in mean field theories which do not have a physical basis in the communicable disease paradigm, namely that pathogens have long-range interactions, which by construction would imply near-luminal effect-velocities, which is a clear violation of this paradigm.

p.6 …the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.
Comment: The authors should explain: what is the nature and origin of this globally affecting biosphere stressor, which plays a key role in the interpretation of the results obtained. The context of the work suggests that it plays a similar role to the cosmological constant in the general relativity, guaranteeing the stability (stationarity) of the solutions.

Answer: Here we have intentionally left out a discussion of such stressor mechanisms and causes due to the uncertainty of such speculations. We do not wish to derail an otherwise scientific investigation and in some ways falsification of the communicable disease paradigm into somewhat speculative alternative theories. However, for now we have tentatively added the following
[PLACEHOLDER: This stressor could be in the form of anomalous geomagnetic storms, or man-made disturbances in the atmosphere. For example, the ionosphere, which is the reservoir of the earth's electromagnetic potential, could have been disturbed either by man-made or natural events, causing organisms to experience an immune response. The exact mechanism of such hypotheses would need further study.]

More litterature review would need to be done to back up these claims.

p.7 …Comparing the stochastic stressor σ in Eq. 14 with the deterministic growth equation in Eq. 1 gives γ= σ/2…
Comment: Why in comparison σ1/2dW(t)/dt (delta-correlated Wiener component) is not included? Only the mean of this term is equal zero, whereas the compared Eqs.1 and 14 do not contain means.

By virtue of Eq. 1 being deterministic, it is implicitly interpreted as a mean (non-random) quantity. In fact, we make the argument that it is more like the median (not mean) of the miscroscopic equation in 14. I see that this was not clearly written and have appended an extra sentence to make this clear:
"This comparison is clear after taking the median of Eq. \ref{eq:microscopicOriginal} wherein the cross-term of $F(t)dW(t)$ goes to zero due to independence."

p.8 …We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as Θ→0.
Comment: In the limit Θ→0 the Richards (generalized logistic) model reduces to the Gompertz one. The latter function is solution of the non-local equation for the time-dependent More oscillator in the dissociation state (Phys. Rev. E 68(2), 021916, 2003), hence, it represents the coherent in space (collaborative) growth. In a similar manner one may prove that the Richards function is solution of the identical equation but for hyperbolic Rosen-Morse potential, therefore it also describes the coherent growth. Hence, both models are collaborative in the nature.

Answer: Thanks for pointing this out. I see there are nuances here that can easily be misinterpreted. My assumption is that when theta is 1,  we get the logistic equation and the growth is non-collaborative, while when theta is 0 we get a "maximally" coherent growth in line with the Morse oscillator. Based on the work of Petroni et al (https://arxiv.org/pdf/2004.10478.pdf) it seems logistic is not collaborative in the sense that "resources are not shared" to use their phrase. Based on this insight, I have modified the inaccurate statement above to:

"We can now further appreciate the Richards parametrization as a transition from non-collaborative to maximally collaborative growth as $\theta \rightarrow 0$. " By inserting the word "maximally", it is hopefully clear that we are on a continuoum of collaborative growth if we could allow such a concept.

p.11…In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
Comment: This conclusion is inconsistent with the Feinberg's extended relativity (Phys. Rev. 159, 1089, 1967). According to Feinberg's theory, pathogens (observable, i.e. subluminal entities) cannot move faster than the speed of light c in a vacuum, because subluminal objects reaching infinite speed would cross the light barrier, which is impossible from the relativistic point of view. Non-local objects (fields) can move (propagate) with the allowed group velocity v=(c, ∞), but this kinetical characteristic does not apply to pathogens moving at v=(0,c). This part of the work should be removed or its interpretation changed.

Your point is well taken, and we of course agree that pathogens cannot travel at superluminal speeds. This is again a flaw in our communication. The intention is to present the implausibility of the tasks of the "pathogen", thus showing that no such pathogen could exist (near-lunumal speed and driven by susceptible instead of infected). We failed to statethe latter part clearly, namely that due to the implications of the gompertz model, we have to reject the concept of a pathogen in this context. A new theory driven by a non-local field must is our suggested replacement, but we are open to ideas. The whole conclusion is now re-written with the relevant section stating:

"
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states. In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in an augmented SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$ and where the source of growth is driven by the number of susceptible and not the number of infected.
[..]
Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected, and be driven by the susceptible as opposed to the infected. We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth. 
"

Additional remarks: A comparison of the SIRD with Gompertz model, graphically presented in Fig.1 does not reflect the differences of the models considered in an acceptable way. In order to compare the models, I suggest to calculate the basic reproduction number R0, which indicates how many secondary infections are caused by an infected individual at the beginning of an epidemic when infection spreading is of exponential type. Indispensable in the calculations formulae and data one may find in the Kong et al. work (PLoS One 16, e0252373, 2021). References [12], [48], [51] are incomplete and should be supplemented with the missing information.

In conclusion: the work NODY-D-22-02727 submitted for publication in Nonlinear Dynamics requires the inclusion of the corrections suggested, explanations mentioned and supplementary calculations that will reveal the advantages of the proposed model over the models functioning in the area of current research. Only then it will be possible to decide whether the work is suitable for publication or not.

### 2022-11-29 arxiv/main.tex
resond to reviewer 2

@@
 The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
-Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates.
+Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \citep{carletti2020covid}.
+The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \citep{efron2012bayesian}.
 Fitting is done with Python-Scipy and a 6-day moving average of deaths.
@@
 
-\section*{Unifying the SIR model and the Gompertz model}
-\subsection*{Comparing microscopic logistic and Gompertz growth}
+\section*{Comparing logistic and Gompertzian growth in the microscopic domain}
 The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
@@
 
-\subsection*{Unification through a modified SIR model}
-We can now reconcile the SIR model with the Gompertz curve. 
-Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
-Then, we reverse the causality where the population of infected is now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
+\section*{Unifying the SIR model with Gompertzian growth}
+We now reconcile the SIR model with the Gompertz curve. 
+Start with the networked SIR equation for the infected group,
 \begin{equation}
-\label{eq:reverseNetworkSIR}
-\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(1-x_j) \quad \forall i.
+\label{eq:SIRnetwork}
+\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j - \alpha x_i\quad \forall i,
 \end{equation}
-This crucial change is based on the environmental stressor hypothesis rather than a communicable disease assumption.
+
+We then modify this equation in two ways: 1.
+Reverse the causality where the population of infected is now dependent on the population of the susceptible.
+This causality reversal is the crucial change based on the environmental stressor hypothesis rather than a communicable disease assumption.
+2.
+Augment the transmission term to higher orders, which as we shall see is necessary to achieve Gompertzian growth.
+
+Thus, Letting $s_i = 1 - x_i$ be the probability of being susceptible,
+
+\begin{equation}
+\label{eq:modSIR}
+\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
+\end{equation}
+
 Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
@@
 \end{equation}
-If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model's infection term becomes 
-\begin{equation}
-\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
-\end{equation}
+
+To build some intuition, we consider some simple scenarios. 
+\begin{enumerate}
+\item Let $x_j=1$, $\forall j \neq i$.
+In this scenario, Eq.
+\ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the environmental stressor is gone, and the value of $x_i$ goes to zero.
+In the communicable disease paradigm, Eq.
+\ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will growth as long as the right-hand side is positive.
+\item Let $x_j \rightarrow 0$, $\forall j \neq i$.
+In this scenario, Eq.
+\ref{eq:modSIR} $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases.
+This behavior is in line with Gompertzian growth and is expected as growth is driven by the susceptible.
+The implication is that the environmental stressor has just started.
+In the communicable disease paradigm Eq.
+\ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = - \alpha x_i$, and the value of $x_i$ goes to zero due to the absence of infectious people.
+\end{enumerate}
+It is important to note that both sets of equations only describe the paradigm they represent. Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm.
 
@@
 \begin{equation}
-\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}\sum_i^N x_i.
+\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}.
 \end{equation}
@@
 \end{equation}
+We see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and correlated to each other as under an environmental stressor.
 

### 2022-11-30 ResponseToReviewers.txt
resond to reviewer 2
finalize reviewer two response

@@
+Comments from Reviewer #2:
+--------------------------------
+Authors propose a new theory describing epidemic spreads, which is based on the hypothesis of the action of an environmental stress, with a proposal of unification of SIR model and Gompertz model.
+To this end they resort to a microscopic (networked) description in which the macroscopic variable derives from a geometric mean instead from an arithmetic one.
+In their approach the usual set of
+equations in a networked model is replaced by one in which the causality between the infected and susceptible populations is reversed.
+
+First of all, the usual networked equation for the node i is
+
+d x_i (t)/dt = \beta (1 - x_i (t)) \sum_{j} a_{ij) x_j (t) - \gamma x_i (t),    \beta, \gamma>0,
+
+while in the article the last term is omitted. Is this a motivated choice or a misprint?
+
+Anyway, authors replace these equations by the set
+
+d x_i (t)/dt = \beta x_i (t)) \sum_{j} a_{ij) (1 - x_j (t)).
+
+To understand meaning and plausibility of the equations one can consider a single node in very simplified situations.
+
+In the first usual form, if a_{ij) is not zero only for j different from i and the x_j (j different from i) are all zero (connected sites not infected), then x_i goes exponentially to zero also if it starts from an initial non zero value.
+In the same hypothesis on a_{ij}, for \gamma = 0 and x_j = 1 for all j different from i (all the connected site surely infected) x_i goes exponetially to 1 for large times.
+These behaviors appear reasonable.
+
+Let us look now to the form proposed by the authors.
+In the same hypothesis that a_{ij} is not zero only for j different from i, if the x_j (j different from i) are all zero (connected sites not infected) the probability x_i of the site i to be infected goes to infinity (unless \beta < 0), while if the x_j (j different from i) are all 1 (all connected sites surely infected) the probability x_i keeps its initial value at all times.
+This seams not reasonable.
+
+Therefore, the conceptual hypotheses of the autors could be interesting, but they have been traduced in not sound arguments.
+
+In conclusion, the article is not suitable for publication.
+
+Response to reviewer #2:
+--------------------------------
+The reviewer makes one specific point about the microscopic logistic and modified logistic model and one the basis of this comment deems our publiation not suitable for publication.
+Let us first address the specific critisicm, which related to the plausibility of the modified SIR model, where causality is reversed, meaning that the "susceptible" drives the "infections".
+Because we are operating under a new paradigm, we do not consider an infected person as infectious, but rather consider the susceptible as the drivers of the growth, as they represent the stage of the environmental stressor.
+For the case where all other entities are infected (x_j = 1), the initial value equals the final value.
+This is because there are no more susceptible entities which is interpreted as the environmental stressor having passed.
+For the case where all other entities are susceptible (x_j = 0), the growth rate will be at it's maximum, exactly as in the gompertz model, meaning that we are just at the start of the environmental stressor.
+So the equations do indeed reflect the desired outcome, but they must be taken under the correct paradigm.
+Consequently, they would appear nonsensical if applied under the opposite paradigm.
+The text has now included a more detailed discussion on both of these simple yet instructive scenarios, and we thank the reviewer for pointing out this confusion.
+
+Further, the reviewer points out that the SIR equation for infected was wrongly quoted.
+We see that the confusion arose because the section was titled "Unifying the SIR with the Gompertz" and we thus proceeded to start with the logistic equation *not* the SIR infection equation, hence the confusion.
+However, to make it more clear, the section has been renamed to "Comparing logistic and Gompertzian growth in the microscopic domain".
+We then proceed with the actual unification in the proceeding section now named "Unifying the SIR model with Gompertzian growth".
+Hopefully this will be more clear.
+
 Response to reviewer #3:
-
+--------------------------------
 Reviewer #3: After reading the article reviewed, I came to the following conclusions: from a mathematical point of view, the model seems to be correct (unless I have overlooked a mistake), while the interpretation of the results obtained is, to put it mildly, controversial, and in some respects incorrect or requiring additional explanations.
@@
 
+Answer:
+Thank you for these suggestions.
+We have now modified our calculations to quote the basic reproduction number as modeled by Carletti et al.
+(see references) instead of the transmission rate.
+We also use a bootstrap to get reasonable confidence intervals (unlike Carletti).
+
 In conclusion: the work NODY-D-22-02727 submitted for publication in Nonlinear Dynamics requires the inclusion of the corrections suggested, explanations mentioned and supplementary calculations that will reveal the advantages of the proposed model over the models functioning in the area of current research.

### 2022-11-30 arxiv/main.tex
finalize reviewer two response

@@
 
-In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
+In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,liu2020emerging,medica2020covid,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
 Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame?
@@
 \end{equation}
-Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \cite{roda2020difficult}.
+Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters, which plagues more complex models \cite{roda2020difficult}.
 
@@
 \section*{Unifying the SIR model with Gompertzian growth}
-We now reconcile the SIR model with the Gompertz curve. 
-Start with the networked SIR equation for the infected group,
+It is possible to reconcile the SIR model with the Gompertz curve under a common model different from the traditional Richards model.
+First, start with the networked SIR equation for the infected group,
 \begin{equation}
 \label{eq:SIRnetwork}
-\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j - \alpha x_i\quad \forall i,
+\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j - \alpha x_i\quad \forall i.
 \end{equation}
@@
 
-Thus, Letting $s_i = 1 - x_i$ be the probability of being susceptible,
+Thus, letting $s_i = 1 - x_i$ be the probability of being susceptible,
 
@@
 
-To build some intuition, we consider some simple scenarios. 
+To build some intuition about these equations, we consider some simple scenarios. 
 \begin{enumerate}
@@
 In the communicable disease paradigm, Eq.
-\ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will growth as long as the right-hand side is positive.
+\ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will grow as long as the right-hand side is positive.
+This is expected as all neighbors are able to transmit the disease.
 \item Let $x_j \rightarrow 0$, $\forall j \neq i$.
@@
 \ref{eq:modSIR} $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases.
-This behavior is in line with Gompertzian growth and is expected as growth is driven by the susceptible.
+This behavior is in line with Gompertzian growth and is expected as growth rates are indicated by the susceptible.
 The implication is that the environmental stressor has just started.
@@
 \end{equation}
-We see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and correlated to each other as under an environmental stressor.
+We thus see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and are correlated with each other as under an environmental stressor.
 
-Finally we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
+Finally, we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
 Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with
@@
  \label{eq:modLogistic}
-\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)}.
+\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)},
  \end{equation}
- Note that for $N=1$, this relative growth rate becomes the standard logistic function, while higher order terms will move closer to the Gompertz curve as the hypergerometric term approaches zero.
+ which represents the novel bridge between logistic and Gompertzian growth.
+Note that for $N=1$, this relative growth rate becomes the standard logistic function, while increasing the number of higher order terms will move the equation closer to the Gompertz curve as the Hypergeometric term approaches zero.
 This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig.
 \ref{fig:hypergeometric}).
+This novel growth equation can thus be included in the SIR equation for the infected to model collaborative growth.
 
@@
 
-Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected, and be driven by the susceptible as opposed to the infected.
+Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected.
 We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.
@@
 The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders.
-From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \cite{molski2003coherent}.
+From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems.
 
 Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles.
-The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology \cite{lambert2013quantum}.
-One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
+This observation adds to the list of emerging biological phenomena that exhibit coherent quantum behavior at the macroscopic level \cite{lambert2013quantum}.
+Although en environmental stressor can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
 
@@
 
-The transmission rate is modeled as the product of $\eta \xi$ in Eq.
-\ref{eq:SIRD_D}, and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \cite{nadarajah2016distribution},
-\begin{equation}
-\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\text{E}[\eta]^2 + \text{Var}[\eta] \text{E}[\xi]^2,
-\end{equation}
-where estimates are obtained by substituting the model parameters with those estimated through the technique outlined above.
+The basic reproduction number is modeled as the product of $\eta \xi / \kappa$ in Eq.
+\ref{eq:SIRD_D}, and its associated variance is obtained through a parametric bootstrap assuming a multivariate normal distribution with mean and covariance equal to the estimate from the non-linear least squares technique.
+The $95^{th}$ percent confidence interval is obtained after 10000 parameter samples drawn from this distribution, each computing a resampled estimate of $\mathcal{R}_0$ \citep{efron2012bayesian}.
 

### 2022-12-09 ResponseToReviewers.txt
expand on mechanistic and evidence discussion

@@
 We do not wish to derail an otherwise scientific investigation and in some ways falsification of the communicable disease paradigm into somewhat speculative alternative theories.
-However, for now we have tentatively added the following
-[PLACEHOLDER: This stressor could be in the form of anomalous geomagnetic storms, or man-made disturbances in the atmosphere.
-For example, the ionosphere, which is the reservoir of the earth's electromagnetic potential, could have been disturbed either by man-made or natural events, causing organisms to experience an immune response.
-The exact mechanism of such hypotheses would need further study.]
-
-More litterature review would need to be done to back up these claims.
+However, after some deliberation, we have added a few paragraphs on mechanitic and observational evidence for an environmental stressor theory.
+Although this mixing of biological mechanisms and strictly physical modelling might be spreading ourselves too thin, so to speak, so we have keept the former discussion to a minimum and relied mostly on the existing literature.
 

### 2022-12-09 arxiv/main.tex
expand on mechanistic and evidence discussion

@@
 While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model.
-This alternative theory represents an alternative paradigm in disease spreading dynamics.
-We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model as an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.
+
+This environmental stressor could be in the form of short-term solar and geomagnetic disturbances, which have been associated with physiological changes in cardiac system processes \cite{vieira2022geomagnetic,Alabdulgader2018,ghione1998geomagnetic,chernouss2001geophysical,stoupel1995ambulatory,stoupel2011twenty,malin1979correlation,stoupel1993sudden}, melatonin/serotonin balance \cite{burch1999geomagnetic,rapoport1997effects,bergiannaki1996seasonal}, endothelial activation and inflammation \cite{schiff2022role}, and calcium fluxes \cite{blackman1990elf}, all with the ability to cause neurological and pulmonary adverse effects.
+More general effects have also been studied in this context, e.g.
+overall mortality, suicide, metal disorders, homicides, and epileptic seizures \cite{zilli2019geomagnetic,gordon2003effect,kay1994geomagnetic,kay2004schizophrenia,halberg2005chronomics,berk2006ambient,doronin1998effect,persinger1995sudden}.
+Cherry (2002) put forward a mechanism whereby the ionosphere acts as a coupling agent between geomagnetic disturbances and physiological effects based on the disturbance of the background electromagnetic oscillations between the surface of the Earth and the lower ionosphere, called the Schumann Resonance signal \cite{cherry2002schumann,schumann1952strahlungslosen}.
+The Schumann Resonance would then couple with the brain to alter calcium fluxes and the melatonin/serotonin balance, which was a causal link observed by Wever (1974) through the circadian rhythm \cite{wever1974elf}.
+
+While the Schumann Resonance mechanism is postulated to be caused by natural geomagnetic disturbances, others have looked at possible causative links through man-made radiation stressors and have found a similar pattern in the symptomology of a Coronavirus infection and radiation injury \cite{rios2021commonalities,Rubik2021}.
+
+Whatever the cause, the environmental stressor theory represents an alternative paradigm in disease spreading dynamics.
+Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model.
+This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations.
+In doing this we shall show from a physical argument that the Gompertz model is incompatible with disease spreading through social interactions.
 
@@
 The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution.
-Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \citep{carletti2020covid}.
-The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \citep{efron2012bayesian}.
+Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \cite{carletti2020covid}.
+The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \cite{efron2012bayesian}.
 Fitting is done with Python-Scipy and a 6-day moving average of deaths.
@@
 An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.
-[PLACEHOLDER: This stressor could be in the form of anomalous geomagnetic storms, or man-made disturbances in the atmosphere.
-For example, the ionosphere, which is the reservoir of the earth's electromagnetic potential, could have been disturbed either by man-made or natural events, causing organisms to experience an immune response.
-The exact mechanism of such hypotheses would need further study.]
 
@@
 \ref{eq:rypdal} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$.
+The maximally coherent scenario implies that all units in the network are exposed to the stressor simultaneously.
 Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
@@
 
-Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected.
+Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected.
 We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.
@@
 \ref{eq:SIRD_D}, and its associated variance is obtained through a parametric bootstrap assuming a multivariate normal distribution with mean and covariance equal to the estimate from the non-linear least squares technique.
-The $95^{th}$ percent confidence interval is obtained after 10000 parameter samples drawn from this distribution, each computing a resampled estimate of $\mathcal{R}_0$ \citep{efron2012bayesian}.
+The $95^{th}$ percent confidence interval is obtained after 10000 parameter samples drawn from this distribution, each computing a resampled estimate of $\mathcal{R}_0$ \cite{efron2012bayesian}.
 
@@
 
-% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\citep{risken1996fokker},
+% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\cite{risken1996fokker},
 % \begin{equation}
@@
 Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}
+Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}
 

### 2022-12-16 arxiv/main.tex
first draft after Thea's comments

@@
 A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
-We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor.
+We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on a global biosphere disturbance.
 Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
@@
 Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame?
-While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model.
+While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on a global biosphere disturbance with fewer parameters than the most basic communicable disease model.
 
-This environmental stressor could be in the form of short-term solar and geomagnetic disturbances, which have been associated with physiological changes in cardiac system processes \cite{vieira2022geomagnetic,Alabdulgader2018,ghione1998geomagnetic,chernouss2001geophysical,stoupel1995ambulatory,stoupel2011twenty,malin1979correlation,stoupel1993sudden}, melatonin/serotonin balance \cite{burch1999geomagnetic,rapoport1997effects,bergiannaki1996seasonal}, endothelial activation and inflammation \cite{schiff2022role}, and calcium fluxes \cite{blackman1990elf}, all with the ability to cause neurological and pulmonary adverse effects.
+This global biosphere disturbance could be in the form of short-term solar and geomagnetic disturbances, which have been associated with physiological changes in cardiac system processes \cite{vieira2022geomagnetic,Alabdulgader2018,ghione1998geomagnetic,chernouss2001geophysical,stoupel1995ambulatory,stoupel2011twenty,malin1979correlation,stoupel1993sudden}, melatonin/serotonin balance \cite{burch1999geomagnetic,rapoport1997effects,bergiannaki1996seasonal}, endothelial activation and inflammation \cite{schiff2022role}, and calcium fluxes \cite{blackman1990elf}, all with the ability to cause neurological and pulmonary adverse effects.
 More general effects have also been studied in this context, e.g.
-overall mortality, suicide, metal disorders, homicides, and epileptic seizures \cite{zilli2019geomagnetic,gordon2003effect,kay1994geomagnetic,kay2004schizophrenia,halberg2005chronomics,berk2006ambient,doronin1998effect,persinger1995sudden}.
+overall mortality, suicide, mental disorders, homicides, and epileptic seizures \cite{zenchenko2021possible,zilli2019geomagnetic,gordon2003effect,kay1994geomagnetic,kay2004schizophrenia,halberg2005chronomics,berk2006ambient,doronin1998effect,persinger1995sudden}.
 Cherry (2002) put forward a mechanism whereby the ionosphere acts as a coupling agent between geomagnetic disturbances and physiological effects based on the disturbance of the background electromagnetic oscillations between the surface of the Earth and the lower ionosphere, called the Schumann Resonance signal \cite{cherry2002schumann,schumann1952strahlungslosen}.
@@
 While the Schumann Resonance mechanism is postulated to be caused by natural geomagnetic disturbances, others have looked at possible causative links through man-made radiation stressors and have found a similar pattern in the symptomology of a Coronavirus infection and radiation injury \cite{rios2021commonalities,Rubik2021}.
+It is further conceivable to have an anthropogenic disturbance of the ionosphere through the use of e.g.
+satellite communication technologies \cite{siingh2005review}.
+Zaporozhan and Ponomarenko (2010) proposed the cryptochrome protein-family to be the designated geomagentic receptors able to pick up minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}.
+The cryptochrome proteins have also been linked to circadian regulation \cite{rodgers2009chemical} and have been extensively studied in the context of avian navigation \cite{hiscock2016quantum}.
 
-Whatever the cause, the environmental stressor theory represents an alternative paradigm in disease spreading dynamics.
+Whatever the cause, global biosphere disturbance theory represents an alternative paradigm in disease spreading dynamics.
 Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model.
@@
 
- The Gompertzian limit implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. 
+ The Gompertzian limit implies a decreasing relative growth rate from the very first time point, which under the SIR model is unlikely given the large pool of susceptible people in the beginning. 
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
@@
 Under a double-log-transform of Eq.
-\ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
+\ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve \cite{Levitt2020},
 \begin{equation}
@@
 
-The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly.
+More generally, the SIR model family considers averaged macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly.
 Rather, all entities are connected and communicating through long-range interactions as shown by Mombach et al.
@@
 Reverse the causality where the population of infected is now dependent on the population of the susceptible.
-This causality reversal is the crucial change based on the environmental stressor hypothesis rather than a communicable disease assumption.
+This causality reversal is the crucial change based on the global biosphere disturbance hypothesis rather than a communicable disease assumption.
 2.
@@
 In this scenario, Eq.
-\ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the environmental stressor is gone, and the value of $x_i$ goes to zero.
+\ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the global biosphere disturbance is gone, and the value of $x_i$ goes to zero.
 In the communicable disease paradigm, Eq.
@@
 This behavior is in line with Gompertzian growth and is expected as growth rates are indicated by the susceptible.
-The implication is that the environmental stressor has just started.
+The implication is that the global biosphere disturbance has just started.
 In the communicable disease paradigm Eq.
@@
 \end{enumerate}
-It is important to note that both sets of equations only describe the paradigm they represent. Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm.
+It is important to note that both sets of equations only describe the paradigm they represent.
+Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm.
+In the new paradigm, we do not consider an infected person as infectious, but rather consider the susceptible as the drivers of the growth, as they represent the stage of the global biosphere disturbance.
 
@@
 \end{equation}
-We thus see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and are correlated with each other as under an environmental stressor.
+We thus see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and are correlated with each other as under a global biosphere disturbance.
 
@@
 Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected.
-We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.
+We are thus forced to reject the concept of a pathogen in the context of Gompertzian growth based on physical grounds.
 
 We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
-The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders.
+The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between individuals or microscopic entities is present at all orders.
 From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems.
@@
 This observation adds to the list of emerging biological phenomena that exhibit coherent quantum behavior at the macroscopic level \cite{lambert2013quantum}.
-Although en environmental stressor can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
+Although an global biosphere disturbance can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
 

### 2022-12-19 arxiv/main.tex
updating intro with thea's new refs

@@
 
-This global biosphere disturbance could be in the form of short-term solar and geomagnetic disturbances, which have been associated with physiological changes in cardiac system processes \cite{vieira2022geomagnetic,Alabdulgader2018,ghione1998geomagnetic,chernouss2001geophysical,stoupel1995ambulatory,stoupel2011twenty,malin1979correlation,stoupel1993sudden}, melatonin/serotonin balance \cite{burch1999geomagnetic,rapoport1997effects,bergiannaki1996seasonal}, endothelial activation and inflammation \cite{schiff2022role}, and calcium fluxes \cite{blackman1990elf}, all with the ability to cause neurological and pulmonary adverse effects.
+This global biosphere disturbance could be in the form of sudden ionospheric disturbances (SIDs) \cite{cannon1982schumann}, which in the context of short-term solar and geomagnetic disturbances have been associated with physiological changes in cardiac system processes \cite{vieira2022geomagnetic,Alabdulgader2018,ghione1998geomagnetic,chernouss2001geophysical,stoupel1995ambulatory,stoupel2011twenty,malin1979correlation,stoupel1993sudden}, melatonin/serotonin balance \cite{burch1999geomagnetic,rapoport1997effects,bergiannaki1996seasonal}, endothelial activation and inflammation \cite{schiff2022role}, and disturbed magnetoreception in migrating animals \cite{keeton1974normal,schreiber1976correlation,kowalski1988normal,bianco2019magnetic,ferrari2017cetacean,vanselow2018solar,granger2020gray}.
 More general effects have also been studied in this context, e.g.
@@
 Cherry (2002) put forward a mechanism whereby the ionosphere acts as a coupling agent between geomagnetic disturbances and physiological effects based on the disturbance of the background electromagnetic oscillations between the surface of the Earth and the lower ionosphere, called the Schumann Resonance signal \cite{cherry2002schumann,schumann1952strahlungslosen}.
-The Schumann Resonance would then couple with the brain to alter calcium fluxes and the melatonin/serotonin balance, which was a causal link observed by Wever (1974) through the circadian rhythm \cite{wever1974elf}.
+The Schumann Resonance would then couple with the brain to alter the melatonin/serotonin balance, which was a causal link observed by Wever (1974) through the circadian rhythm \cite{wever1974elf}.
+Indeed, more recent findings support this hypothesized synchronization between Schumann Resonances and the brain \cite{pobachenko2006contingency,saroka2014quantitative,persinger2015human,konig2012biologic}.
 
-While the Schumann Resonance mechanism is postulated to be caused by natural geomagnetic disturbances, others have looked at possible causative links through man-made radiation stressors and have found a similar pattern in the symptomology of a Coronavirus infection and radiation injury \cite{rios2021commonalities,Rubik2021}.
-It is further conceivable to have an anthropogenic disturbance of the ionosphere through the use of e.g.
-satellite communication technologies \cite{siingh2005review}.
-Zaporozhan and Ponomarenko (2010) proposed the cryptochrome protein-family to be the designated geomagentic receptors able to pick up minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}.
+In support of Cherry's theory, specific phenomena such as solar X-ray bursts \cite{dyrda2015novel,roldugin2004schumann,satori2005response,satori2016effects,shvets2017effect}, solar proton events \cite{roldugin2004schumann,schlegel1999schumann,singh2014x} and geomagnetic storms \cite{pazos2019analysis,salinas2016solar} can modify the shape of the ionosphere cavity and shift the Schumann resonance frequencies \cite{kudintseva2018modifications,satori2016effects,zhou2015studies}.
+Seismic activity has also been connected to variations in Schumann resonances \cite{christofilakis2019significant,galuk2019scattering,hayakawa2020scattering}, and conversely, ionospheric modifications have been connected to global seismicity \cite{Marchitelli2020}.
+
+While Schumann Resonance disturbances can be caused by natural geomagnetic events, others have looked at possible causative links through man-made influences on the ionosphere and magnetosphere through the use of electromagnetic infrastructure \cite{siingh2005review}.
+It is thus conceivable to have an anthropogenic disturbance of the ionosphere through the use of e.g.
+satellite communication technologies.
+Not only might this effect the
+global electric circuit \cite{rycroft2008overview} and the Schumann Resonance, but the usage of satellite communication technologies might also alter the magnetosphere itself leaving a signal similar to Power Line Harmonic Radiation \cite{bullough2017power,kositsky2001influence}.
+
+Zaporozhan and Ponomarenko (2010) proposed the cryptochrome protein-family to be the designated geomagnetic receptors able to pick up minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}.
 The cryptochrome proteins have also been linked to circadian regulation \cite{rodgers2009chemical} and have been extensively studied in the context of avian navigation \cite{hiscock2016quantum}.
+Besides the cryptochromes, Krylov (2017) presents evidence and references that another iron-containing protein called the magnetoreceptor protein (MagR) encoded by the gene CG8198 is of equal importance \cite{krylov2017biological}.
+He suggested that geomagnetic storms could be perceived by animals as a disruption of the precise synchronization of biological circadian rhythms with diurnal geomagnetic variation, an interplay involving the cryptochrome, MagR, and melatonin molecules.
 

### 2022-12-19 arxiv/main.tex
moved paragraph about macroscopic assumptions

@@
 Estrada and Bartesaghi (2022) \cite{estrada2022networked} found that the Gompertz curve can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
- Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
+ Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to establish this connection. 
  We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.
+
+ In general, the SIR models and their extensions which do not consider network interactions explicitly consider rather averaged macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond}.
+In such models, all entities are connected and communicating through long-range interactions as shown by Mombach et al.
+(2002) \cite{mombach2002mean}.
+However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIR model family predicts, but rather a constant negative slope in the double-log-domain, which we will now show.
 
@@
 
-More generally, the SIR model family considers averaged macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly.
-Rather, all entities are connected and communicating through long-range interactions as shown by Mombach et al.
-(2002) \cite{mombach2002mean}.
-However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log-domain.
+In summary, we see that the SIR model family both lacks the observed immediate decrease in relative growth rate, and makes strong assumptions about the disease spreading network structure that imply non-local interactions.
 We are thus prompted to look for another model which can explain the observed mortality patterns.
@@
 
+\subsection*{A closed-form expression for a unified model}
 Finally, we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.

### 2022-12-20 arxiv/main.tex
update for thea's comments 2

@@
 \begin{abstract}
-A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
+A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial covid pandemic period, March-April 2020. 
 We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on a global biosphere disturbance.
-Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
-Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+Second, we show that for the Gompertz curve to emerge, the disturbance has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
+Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model family is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
 Crucially, this augmentation must be accompanied by a causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people.
@@
 Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}.
-A large body of research reveals how these models fit the mortality seen due to the Coronavirus epidemic in 2020 \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
+A large body of research reveals how these models fit the mortality seen due to the Covid pandemic in 2020, a pandemic due to an influenza-like disease \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.
 
@@
 
-This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, a conclusion made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
+This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the predicted logistic curves, an observation made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics.
 What is causing such a discrepancy between these observations and the current theory of communicable diseases \cite{castro2020turning}?
@@
 Not only might this effect the
-global electric circuit \cite{rycroft2008overview} and the Schumann Resonance, but the usage of satellite communication technologies might also alter the magnetosphere itself leaving a signal similar to Power Line Harmonic Radiation \cite{bullough2017power,kositsky2001influence}.
+global electric circuit \cite{rycroft2008overview} and the Schumann Resonance, but the usage of satellite communication technologies might also alter the magnetosphere itself leaving a signal similar to Power Line Harmonic Radiation \cite{bullough2017power}.
 
@@
 This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations.
-In doing this we shall show from a physical argument that the Gompertz model is incompatible with disease spreading through social interactions.
+In doing this we shall show that the Gompertz model is incompatible with disease spreading through physical proximity interactions.
+Further, we show that the Gompertz curve emerge when individuals are connected through a common stressor affecting them simultaneously.
 
@@
  In general, the SIR models and their extensions which do not consider network interactions explicitly consider rather averaged macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond}.
-In such models, all entities are connected and communicating through long-range interactions as shown by Mombach et al.
+In such models, all entities are implicitly assumed connected and communicating through long-range interactions as shown by Mombach et al.
 (2002) \cite{mombach2002mean}.
@@
 \subsection*{The extended SIR model family is almost Gompertzian}
-Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models.
+Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease SIR-like models.
 Under a double-log-transform of Eq.
@@
 \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$.
-Both the SIR and SEIR model exhibit similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \cite{vattay2020forecasting}.
+Both the SIR and SEIR model exhibit similar properties and their discrepancy with initial observations related to the Covid pandemic has been noted by others \cite{vattay2020forecasting}.
 
@@
 
-\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
+\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Covid pandemic, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020.
 Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text.

### 2022-12-20 arxiv/main.tex
capitalize covid in abstract

@@
 \begin{abstract}
-A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial covid pandemic period, March-April 2020. 
+A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Covid pandemic period, March-April 2020. 
 We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on a global biosphere disturbance.

### 2022-12-21 arxiv/main.tex
fix to emerges
update small fixes in main body

@@
 Zaporozhan and Ponomarenko (2010) proposed the cryptochrome protein-family to be the designated geomagnetic receptors able to pick up minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}.
-The cryptochrome proteins have also been linked to circadian regulation \cite{rodgers2009chemical} and have been extensively studied in the context of avian navigation \cite{hiscock2016quantum}.
+The cryptochrome proteins have also been linked to circadian regulation \cite{rodgers2009chemical} and have been extensively studied in the context of avian navigation \cite{engels2014anthropogenic,
+hiscock2016quantum}.
 Besides the cryptochromes, Krylov (2017) presents evidence and references that another iron-containing protein called the magnetoreceptor protein (MagR) encoded by the gene CG8198 is of equal importance \cite{krylov2017biological}.
@@
 In doing this we shall show that the Gompertz model is incompatible with disease spreading through physical proximity interactions.
-Further, we show that the Gompertz curve emerge when individuals are connected through a common stressor affecting them simultaneously.
+Further, we show that the Gompertz curve emerges when individuals are connected through a common stressor affecting them simultaneously.
 
@@
 Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease SIR-like models.
-Under a double-log-transform of Eq.
-\ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve \cite{Levitt2020},
+Under a double-log-transform of Gompertzian growth Eq.
+\ref{eq:gomp_solution}, the growth curve is a simple straight line which is one of its characteristic traits \cite{Levitt2020},
 \begin{equation}
@@
 with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$.
-To show this, we follow Carletti et al.
+To show this discrepancy between the SIR-like models and the Gompertz model, we follow Carletti et al.
 (2020) \cite{carletti2020covid} and consider the extended version of the SIR model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model.
@@
 We are thus prompted to look for another model which can explain the observed mortality patterns.
+For this challenge, we shall first present a stochastic approach and then present a deterministic approach which unifies the SIR family models with the Gompetz curve.
 
@@
 
-\section*{An alternative theory for observed mortality}
+\section*{A stochastic theory for observed mortality}
 \label{seq:alt}
@@
 In this scenario, Eq.
-\ref{eq:modSIR} $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases.
+\ref{eq:modSIR} becomes $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases.
 This behavior is in line with Gompertzian growth and is expected as growth rates are indicated by the susceptible.
@@
 \ref{fig:hypergeometric}).
-This novel growth equation can thus be included in the SIR equation for the infected to model collaborative growth.
+This novel growth equation can thus be included in the SIR equation for the infected.
 
 \insertPdfFig{hypergeometric}{Comparison relative growth rate in the Richards $\theta$-logistic model (Eq.
-\ref{eq:rypdal}), Gompertz model (Eq.
+\ref{eq:rypdalGamma}), Gompertz model (Eq.
 \ref{eq:medianGomp}), and the augmented causality-reversed SIR model (Eq.
@@
 (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq.
-\ref{eq:rypdal} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$.
+\ref{eq:rypdalGamma} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$.
 The maximally coherent scenario implies that all units in the network are exposed to the stressor simultaneously.
@@
 
-We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
+We further show that the observed mortality across countries can be explained by a model where the biosphere is disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
 The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between individuals or microscopic entities is present at all orders.
@@
 This observation adds to the list of emerging biological phenomena that exhibit coherent quantum behavior at the macroscopic level \cite{lambert2013quantum}.
-Although an global biosphere disturbance can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
+Although a global biosphere disturbance can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.
 

### 2022-12-21 letterToEditor/letter.tex
update letter
minor changes to sci per
update small fixes in main body

@@
 \email{matzhaugen@gmail.com}
-\address{Enerhauggata 1, 0651, Oslo, Norway}
+\address{Valhallveien 13, 0196, Oslo, Norway}
 
@@
 \recipient{To}{Editor, \\
-               Chaos, Solitons, and Fractals,\\ 
+               Bulletin of Mathematical Biology,\\ 
                }
@@
 Due to the conclusions of our paper, I request that this paper receives more reviews than your average, and at least 3 reviews.
+Some reviewers will probably have their scientific ``beliefs'' challenged.
+I would ask them to discern between dogma and observation.
+Science is only at the mercy of the latter.
 
@@
 
-\item Nicola Cufaro Petroni, cufaro@ba.infn.it,
-Dipartimento di Matematica and TIRES, University of Bari (Ret).
-Has done extensive work on the Gompertz model in relation to the Richard's model, connecting it with stochastic growth models and the Logistic (Lotka-Volterra) model.
-We use their concept of growth rooted in stochastic phenomena.
-\item Salvatore De Martino,
-demartino247@gmail.com, 
-Dipartimento di Ingegneria dell’Informazione ed Elettrica.
-Same as above
+\item Vladimir N. Binhi, vnbin@mail.ru, Prokhorov General Physics Institute of the Russian Academy of Sciences, 38 Vavilov St., 119991 Moscow, Russia
+Is familiar with the biological effects due geomagnetic disturbances and fields.
+\item Andrei B. Rubin, rubin@biophysics.msu.ru, Faculty of Biology, Lomonosov Moscow State University, Leninskie Gory 1/12, 119234 Moscow, Russia
+Is familiar with the biological effects due geomagnetic disturbances and fields.
+\item Viacheslav V. Krylov, I.D. Papanin Institute for Biology of Inland Waters Russian Academy of Sciences, Borok, Russian Federation, krylovviacheslav@mail.ru
+Is intimately familiar with geomagnetic disturbances' effect on biology.
 \item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
+\item Takuya Yamano, yamano@amy.hi-ho.ne.jp, Department of Physics, Ochanomizu University, 2-1-1 Otsuka, Bunkyo-ku, 112-8610, Tokyo, Japan
+Is familiar with the gompertz model as it relates to statistical physics.
+\item Shigui Ruan, ruan@math.miami.edu, Professor of Mathematics, University of Miami, 
+An editor of Bulletin of Mathematical Biology
 \item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
 Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
-\item Ernesto Estrada, estrada@ifisc.uib-csic.es, Institute of Interdisciplinary Physics and Complex Systems , mathematics, graph theory, SIR models.
-Has written a recent paper on the Gompertz model in relation to communicable disease modelling including close connections with graph theoretical concepts used in our paper. 
-\item K. Rypdal, kristoffer.rypdal@uit.no, University of Tromsø, mathematics, SIR models.
-Has written a paper showing that SIR models become Gompertz in the extreme of the Richard's parameter. We also follow this paper's arguments in our paper.
-\item Michael Levitt (NAS Member), michael.levitt@stanford.edu, Stanford University, computational biology, Gompertz models.
-Has a preprint on the gompertz model in relation to covid where they leave the explanation of why mortality follows gompertz as an open question.
-\item Chris G. Antonopoulos, canton@essex.ac.uk , U. of Essex, physics, chaos and SIR models.
+\item Rober M May, robert.may@zoo.ox.ac.uk, Professor of Zoology, University of Oxford, 
+\item Chris G. Antonopoulos, canton@essex.ac.uk, U. of Essex, physics, chaos and SIR models.
 Has written on the SIR model and chaotic dynamics.
-\item Constantino Tsallis, tsallis@cbpf.br, Complexity Science Hub Vienna, Austria.
-Has done extensive work on long-range interaction and statistical mechanics.
-He might be able to see connections with his famous Tsallis entropy.
 

### 2022-12-21 arxiv/main.tex
make rypdal more clear

@@
 They further assume the number of deceased is proportional to the number of cumulative infected, offset by a time lag, which allows us to use the same set of linearized equations to model the number of cumulative people deceased without loss of generality.
-
- Due to this linearization, the infection/recovery rate ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions.
 And since $I$ is a function of $Y$, we can combine Eq.
-\ref{eq:linSIR} via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
-\begin{subequations}
-\label{eq:rypdal}
-\begin{align}
-\frac{dY}{dt} =& \gamma(t) Y(t) \label{eq:rypdalODE}\\
-\gamma(t) =& \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] \label{eq:rypdalGamma},
-\end{align}
-\end{subequations}
+\ref{eq:linSIR} via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$,
+\begin{equation}
+\frac{dY}{dt} = \gamma(t) Y(t) \label{eq:rypdalODE}.
+\end{equation}
+Due to this linearization, the infection/recovery rate ratio, $\phi(t)$, and the relative growth rate, $\gamma(t)$, will have to change as a function of time to accommodate for the boundary conditions.
+The latter is typically also parameterized by a scaling factor, $\theta$, representing the shape of the growth,
+\begin{equation}
+\label{eq:rypdalGamma}
+\gamma(t) = \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] ,
+\end{equation}
 where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$.

### 2022-12-21 BulletinOfMathBio/sn-article.tex
add BMB script

--- 2022-12-21 arxiv/main.tex
@@
-\documentclass{article}
-% \documentclass{sn-jnl}
-% \documentclass[review]{elsarticle}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\usepackage{arxiv}
-\usepackage{booktabs}
-\usepackage{siunitx}
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{lipsum}   % Can be removed after putting your text content
-\usepackage{graphicx}
-\usepackage[square,compress,numbers]{natbib}
-\usepackage{doi}
-\usepackage{amsmath}
-\usepackage{float}
-\usepackage{lineno}
-% \linenumbers
+%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing
 
-% \def\bibcommenthead{}%
+%%=======================================================%%
+%% to print line numbers in the margin use lineno option %%
+%%=======================================================%%
 
-\newcommand{\insertPdfFig}[3]{
-  \begin{figure}[H]
-  \centering
-  \includegraphics[width=8.4cm]{#1.pdf}
-  \caption{#2}
-  \label{fig:#1}
-  \end{figure}
-}
+%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+
+%%======================================================%%
+%% to compile with pdflatex/xelatex use pdflatex option %%
+%%======================================================%%
+
+%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+
+%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+\documentclass[sn-mathphys]{sn-jnl}% Math and Physical Sciences Reference Style
+%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
+%%\documentclass[sn-vancouver]{sn-jnl}% Vancouver Reference Style
+%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style
+%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
+%%\documentclass[sn-standardnature]{sn-jnl}% Standard Nature Portfolio Reference Style
+%%\documentclass[default]{sn-jnl}% Default
+%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout
+
+%%%% Standard Packages
+%%<additional latex packages if required can be included here>
+%%%%
+
+%%%%%=============================================================================%%%%
+%%%%  Remarks: This template is provided to aid authors with the preparation
+%%%%  of original research articles intended for submission to journals published 
+%%%%  by Springer Nature. The guidance has been prepared in partnership with 
+%%%%  production teams to conform to Springer Nature technical requirements. 
+%%%%  Editorial and presentation requirements differ among journal portfolios and 
+%%%%  research disciplines. You may find sections in this template are irrelevant 
+%%%%  to your work and are empowered to omit any such section if allowed by the 
+%%%%  journal you intend to submit to. The submission guidelines and policies 
+%%%%  of the journal take precedence. A detailed User Manual is available in the 
+%%%%  template package for technical guidance.
+%%%%%=============================================================================%%%%
 
@@
 }
+\newcommand{\insertPdfFig}[3]{
+  \begin{figure}[h]
+  \centering
+  \includegraphics[width=12.4cm]{#1.pdf}
+  \caption{#2}
+  \label{fig:#1}
+  \end{figure}
+}
+% \graphicspath{{../figures/}}
 
-% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
-\title{Unifying the communicable disease spreading paradigm with Gompertzian growth}
-% \graphicspath{{../figures/}}
-%\date{September 9, 1985} % Here you can change the date presented in the paper title
-%\date{}          % Or removing it
+\jyear{2021}%
 
-\author{Matz A. Haugen, \\
-Independent Reseacher, Oslo\\
-\texttt{matzhaugen@gmail.com} \\
-  \And
-  Dorothea Gilbert \\
-  Independent Reseacher, Oslo\\
-}
+%% as per the requirement new theorem styles can be included as shown below
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}% 
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
 
-% Uncomment to remove the date
-%\date{}
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
 
-% Uncomment to override  the `A preprint' in the header
-%\renewcommand{\headeright}{Technical Report}
-%\renewcommand{\undertitle}{Technical Report}
-\renewcommand{\shorttitle}{On the nature of the Gompertz}
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}%
 
-%%% Add PDF metadata to help others organize their library
-%%% Once the PDF is generated, you can check the metadata with
-%%% $ pdfinfo template.pdf
-\hypersetup{
-pdftitle={A template for the arxiv style},
-pdfsubject={q-bio.NC, q-bio.QM},
-pdfauthor={},
-pdfkeywords={},
-}
+\raggedbottom
+%%\unnumbered% uncomment this for unnumbered level heads
 
 \begin{document}
-\maketitle
 
-\begin{abstract}
-A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Covid pandemic period, March-April 2020. 
+\title[Unifying the communicable disease spreading paradigm]{Unifying the communicable disease spreading paradigm with Gompertzian growth}
+
+%%=============================================================%%
+%% Prefix	-> \pfx{Dr}
+%% GivenName	-> \fnm{Joergen W.}
+%% Particle	-> \spfx{van der} -> surname prefix
+%% FamilyName	-> \sur{Ploeg}
+%% Suffix	-> \sfx{IV}
+%% NatureName	-> \tanm{Poet Laureate} -> Title after name
+%% Degrees	-> \dgr{MSc, PhD}
+%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
+%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
+%%=============================================================%%
+
+\author*[1]{\fnm{Matz} \sur{Haugen}}\email{matzhaugen@gmail.com}
+
+\author[2]{\fnm{Dorothea} \sur{Gilbert}}\email{D.Thea@protonmail.ch}
+% \equalcont{These authors contributed equally to this work.}
+
+\affil*[1]{\orgname{Abel X}, \orgaddress{\city{Oslo}, \postcode{1360}, \state{Baerum}, \country{Norway}}}
+
+\affil[2]{\orgname{Independent Researcher}, \orgaddress{\city{Oslo}, \country{Norway}}}
+
+% \affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}
+
+%%==================================%%
+%% sample for unstructured abstract %%
+%%==================================%%
+
+\abstract{A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Covid pandemic period, March-April 2020. 
 We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on a global biosphere disturbance.
@@
 We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
-\end{abstract}
+}
 
-% keywords can be removed
-\keywords{Gompertz \and Coherence \and Covid \and Coronavirus \and Network Analysis \and Stochastic}
+%%================================%%
+%% Sample for structured abstract %%
+%%================================%%
 
-\section*{Introduction}
+% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}
+
+\keywords{Gompertz, Coherence, Covid, Communicable Diseases, Network, Stochastic Processes}
+
+%%\pacs[JEL Classification]{D8, H51}
+
+%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}
+
+\maketitle
+
+\section{Introduction}\label{sec1}
 Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
@@
 
-\section*{Unifying the SIR model with Gompertzian growth}
+\subsection*{Unifying the SIR model with Gompertzian growth}
 It is possible to reconcile the SIR model with the Gompertz curve under a common model different from the traditional Richards model.
@@
 
-\subsection*{Code Availability}
+\backmatter
 
-Scripts used to produce the plots are downloadable in the form of a Jupyter notebook using Python from 
-\newline
-\url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb}
+\bmhead{Supplementary information}
 
-\subsection*{Data Availability}
+A supplement is available for this article.
+% \bmhead{Acknowledgments}
+
+% \section*{Declarations}
+
+% Some journals require declarations to be submitted in a standardised format.
+Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section.
+If yes, your manuscript must contain the following sections under the heading `Declarations':
+
+\section*{Competing interests}
+The authors have no relevant financial or non-financial interests to disclose.”
+
+\section*{Author Contributions}
+M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
+
+\section*{Data Availability}
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore,
-USA \cite{dong2020interactive}, and are available at \url{https://github.com/CSSEGISandData/COVID-19}, specifically the table \newline\verb|csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv|, \newline downloaded Aug.
+USA \cite{dong2020interactive}, and are available at \url{https://github.com/CSSEGISandData/COVID-19}, specifically the table \newline\verb|time_series_covid19_deaths_global.csv|, \newline downloaded Aug.
 $5^{th}$ 2022.
@@
 
+\section*{Code Availability}
+Scripts used to produce the plots are downloadable in the form of a Jupyter notebook using Python from 
+\newline
+\url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb}
+and \newline
+\url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/hypergeometric.ipynb}
+
+\section*{Funding}
+The authors declare that no funds, grants, or other support were received during the preparation of this manuscript.
+
 \section*{Author Contributions}
@@
 
-\section*{Supplement}
+%%===================================================%%
+%% For presentation purpose, we have included        %%
+%% \bigskip command. please ignore this.             %%
+%%===================================================%%
+% \bigskip
+% \begin{flushleft}%
+% Editorial Policies for:
 
-% Estimating the parameters using real data requires some consideration around the variance of the growth variable as a function of time.
-This variance will in turn inform the estimation process by considering a weighted minimization of the residuals of the observations regressed onto the model.
-In the stochastic Gompertz process postulated in Section \ref{seq:alt}, the process under a log-transformation is the Ornstein-Uhlenbeck process which has a closed from expression for its transition probability, and thus its variance as a function of time.
-This variance can then be used in a weighted least squares estimation technique or any other weighted technique.
-In the original domain there is also a closed form expression available but it is somewhat more convoluted and given that observations in this domain are visually quite homoscedastic with equal variance across time (Fig \ref{fig:Gompertz_vs_SIRD} and Tables \ref{tab:gomp} and \ref{tab:sird}), this is expression and consideration is omitted.
-But for the log-domain, observations exibit changing variance with time.
+% \bigskip\noindent
+% Springer journals and proceedings: \url{https://www.springer.com/gp/editorial-policies}
 
-% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\cite{risken1996fokker},
-% \begin{equation}
-% P(z, t|z', t') = \sqrt{\frac{\beta}{2\pi \sigma (1 - e^{-2\beta (t-t')})}}\exp{\left[ -\frac{\beta (z - e^{-\beta(t-t')}z')^2}{2\sigma(1-e^{-2 \beta(t-t')})} \right]}.
-% \end{equation}
+% \bigskip\noindent
+% Nature Portfolio journals: \url{https://www.nature.com/nature-research/editorial-policies}
 
-% If we consider $t$ the time at which the stationary distribution is reached and that this stationary distribution is given as a delta function, variance of the growth will be given as
-% \begin{equation}
-% \text{Var[Z(t)|t']} = \sigma(1-e^{-2 \beta(t-t')}).
-% \end{equation}
+% \bigskip\noindent
+% \textit{Scientific Reports}: \url{https://www.nature.com/srep/journal-policies/editorial-policies}
 
-% Because weights on observations in a regression are invariant to multiplication we can let $\sigma=1$.
-The slope coefficient would have to be estimated either by a standard least squares or through the 3-parameter model in (\ref{eq:GompertzODE}).
+% \bigskip\noindent
+% BMC journals: \url{https://www.biomedcentral.com/getpublished/editorial-policies}
+% \end{flushleft}
 
-\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model.
-Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text.
-Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
-Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}
+% \begin{appendices}
 
-% \insertPdfFig{Gompertz_vs_SIRD_gompfit_log_domain}{Log transformed cumulative diseased plotted against its time derivative after $Y(t) / Y_{max}>0.005$, with a straight line fit regressing onto the $50^{th}$ quantile.}
+% \section{Section title of first appendix}\label{secA1}
 
-%  Carletti (left) transformation to our notation (right)
-% gamma - eta
-% eta - xi
-% beta - kappa 
-\begin{table}
-\label{tab:sird}
-\centering
-\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure 1 using notation from eq.
-(12).
-Standard deviation estimates is given in the parenthesis corresponding to the last significant digit of the reported number.}
-\begin{tabular}{llll}
-\toprule
-{} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
-\midrule
-Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
-Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
-Austria        &   6.9(5) &    40(3) &   8.7(6) \\
-Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
-Sweden         &    19(5) &     8(2) &     3(1) \\
-Italy          &   137(3) &  1.77(5) &  3.71(1) \\
-France         &  240(10) &  1.24(7) &   8.1(4) \\
-Spain          &   145(2) &  2.27(5) &  4.75(8) \\
-Germany        &    61(4) &   3.9(3) &   6.5(5) \\
-Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
-United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
-Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
-Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
-Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
-US             &   312(5) &  0.76(2) &  1.83(7) \\
-China          &  130(30) &   3.5(5) &    26(4) \\
-\bottomrule
-\end{tabular}
+% An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.
 
-\end{table}
+% %%=============================================%%
+% %% For submissions to Nature Portfolio Journals %%
+% %% please use the heading ``Extended Data''.   %%
+% %%=============================================%%
 
-\begin{table}
-\label{tab:gomp}
-\centering
-\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure 1 using notation from eq. (1).}
-\begin{tabular}{llll}
-\toprule
-{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
-\midrule
-Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
-Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
-Austria        &  6.98(5) &   7.6(1) &   7.8(4) \\
-Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
-Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
-Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
-France         &  10.2(1) &   8.2(1) &  11.0(8) \\
-Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
-Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
-Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
-United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
-Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
-Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
-Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
-US             &  7.53(7) &  5.32(7) &   8.4(5) \\
-China          &   9.8(1) &   9.4(2) &    10(1) \\
-\bottomrule
-\end{tabular}
+% %%=============================================================%%
+% %% Sample for another appendix section			       %%
+% %%=============================================================%%
 
-\end{table}
+% %% \section{Example of another appendix section}\label{secA2}%
+% %% Appendices may be used for helpful, supporting or essential material that would otherwise 
+% %% clutter, break up or be distracting to the text. Appendices can consist of sections, figures, 
+% %% tables and equations etc.
 
-\bibliographystyle{abbrvnat}
-\bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .
+% \end{appendices}
 
-%%% Uncomment this section and comment out the \bibliography{references} line above to use inline references.
-% \begin{thebibliography}{1}
+%%===========================================================================================%%
+%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
+%% system, please include the references within the manuscript file itself. You may do this  %%
+%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
+%% file, and delete the associated \verb+\bibliography+ commands.                            %%
+%%===========================================================================================%%
 
-%   \bibitem{kour2014real}
-%   George Kour and Raid Saabne.
-%   \newblock Real-time segmentation of on-line handwritten arabic script.
-%   \newblock In {\em Frontiers in Handwriting Recognition (ICFHR), 2014 14th
-%       International Conference on}, pages 417--422. IEEE, 2014.
+\bibliography{references2}
+%% if required, the content of .bbl file can be included here once bbl is generated
+%%\input sn-article.bbl
 
-%   \bibitem{kour2014fast}
-%   George Kour and Raid Saabne.
-%   \newblock Fast classification of handwritten on-line arabic characters.
-%   \newblock In {\em Soft Computing and Pattern Recognition (SoCPaR), 2014 6th
-%       International Conference of}, pages 312--318. IEEE, 2014.
-
-%   \bibitem{hadash2018estimate}
-%   Guy Hadash, Einat Kermany, Boaz Carmeli, Ofer Lavi, George Kour, and Alon
-%   Jacovi.
-%   \newblock Estimate and replace: A novel approach to integrating deep neural
-%   networks with existing applications.
-%   \newblock {\em arXiv preprint arXiv:1804.09028}, 2018.
-
-% \end{thebibliography}
+%% Default %%
+%%\input sn-sample-bib.tex%
 

### 2022-12-21 arxiv/main.tex
add BMB script

@@
 \url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb}
+and 
+\url{https://nbviewer.org/urls/emf-research.fra1.digitaloceanspaces.com/gompertz/hypergeometric.ipynb}
 
@@
 Fitting is done with Python-Scipy and a 6-day windowed average of deaths.
+The basic reproduction number, $R_0$ is computed as specified in the Methods.
 Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}

### 2023-01-03 letterToEditor/letter.tex
finalize

@@
 \recipient{To}{Editor, \\
-               Bulletin of Mathematical Biology,\\ 
+               Journal of Mathematical Biology,\\ 
                }
@@
 Due to the conclusions of our paper, I request that this paper receives more reviews than your average, and at least 3 reviews.
-Some reviewers will probably have their scientific ``beliefs'' challenged.
-I would ask them to discern between dogma and observation.
-Science is only at the mercy of the latter.
 
@@
 \begin{enumerate}
-
+\item Tom Britton, Dean of Mathematics and Physics, Stockholm University, tom.britton@math.su.se
+\item Mark Lewis, marklewis@uvic.ca, Department of Mathematics and Statistics, University of Victoria
+\item Alexandru Hening, Department of Mathematics at Texas A and M University,  ahening@tamu.edu
+\item K-Y. Lam, Mathematics,  the Ohio State University, Columbus, lam.184@osu.edu 
+\item X. Zou, Western University, London, Ontario, xzou@uwo.ca
+\item S. Petrovskii, Mathematical Sciences, University of Leicester, sp237@le.ac.uk
+\item Pierre Magal, Institut de Mathématiques de Bordeaux, Université de Bordeaux, pierre.magal@u-bordeaux.fr 
 \item Vladimir N. Binhi, vnbin@mail.ru, Prokhorov General Physics Institute of the Russian Academy of Sciences, 38 Vavilov St., 119991 Moscow, Russia
@@
 \item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
-\item Takuya Yamano, yamano@amy.hi-ho.ne.jp, Department of Physics, Ochanomizu University, 2-1-1 Otsuka, Bunkyo-ku, 112-8610, Tokyo, Japan
-Is familiar with the gompertz model as it relates to statistical physics.
-\item Shigui Ruan, ruan@math.miami.edu, Professor of Mathematics, University of Miami, 
-An editor of Bulletin of Mathematical Biology
 \item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
 Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
-\item Rober M May, robert.may@zoo.ox.ac.uk, Professor of Zoology, University of Oxford, 
-\item Chris G. Antonopoulos, canton@essex.ac.uk, U. of Essex, physics, chaos and SIR models.
-Has written on the SIR model and chaotic dynamics.
+\item Robert M May, robert.may@zoo.ox.ac.uk, Professor of Zoology, University of Oxford.
+\end{enumerate}
+
+Other possible reviewers:
+\begin{enumerate}
+\item Joel Miller
+\item Mark Chaplain
+\item Vishwesha Guttal
+\item Natalia Komorova
+\item Shigui Ruan
+\item Simon Levin
+
+\item Helen Wearing
+\item James Watmough
+\item Jane Heffernan
+\item Jing-an Cui
+\item Julien Arino
 

### 2023-03-30 arxiv/main.tex
finalize
update

@@
 \begin{equation}
-\label{eq:GompertzODE}
-\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
+\label{eq:GompertzODEsimple}
+\frac{d}{dt}\ln{F(t)} = -\beta\ln{F(t)},
 \end{equation} 
-% Redundant definitions for full script
-with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
-\begin{equation}
-\label{eq:gomp_solution}
-Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
-\end{equation}
-where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.
-
+using normalized mortality such that $F(t\rightarrow\infty) = 1$.
 % \insertSmallPdfFig{Gompertz_concept}{A conceptual display of the discrepancy seen between the logistic growth and the observations, superimposed by a Gompertz curve.
@@
 Seismic activity has also been connected to variations in Schumann resonances \cite{christofilakis2019significant,galuk2019scattering,hayakawa2020scattering}, and conversely, ionospheric modifications have been connected to global seismicity \cite{Marchitelli2020}.
-
-While Schumann Resonance disturbances can be caused by natural geomagnetic events, others have looked at possible causative links through man-made influences on the ionosphere and magnetosphere through the use of electromagnetic infrastructure \cite{siingh2005review}.
-It is thus conceivable to have an anthropogenic disturbance of the ionosphere through the use of e.g.
-satellite communication technologies.
-Not only might this effect the
-global electric circuit \cite{rycroft2008overview} and the Schumann Resonance, but the usage of satellite communication technologies might also alter the magnetosphere itself leaving a signal similar to Power Line Harmonic Radiation \cite{bullough2017power}.
-
-Zaporozhan and Ponomarenko (2010) proposed the cryptochrome protein-family to be the designated geomagnetic receptors able to pick up minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}.
-The cryptochrome proteins have also been linked to circadian regulation \cite{rodgers2009chemical} and have been extensively studied in the context of avian navigation \cite{engels2014anthropogenic,
-hiscock2016quantum}.
-Besides the cryptochromes, Krylov (2017) presents evidence and references that another iron-containing protein called the magnetoreceptor protein (MagR) encoded by the gene CG8198 is of equal importance \cite{krylov2017biological}.
-He suggested that geomagnetic storms could be perceived by animals as a disruption of the precise synchronization of biological circadian rhythms with diurnal geomagnetic variation, an interplay involving the cryptochrome, MagR, and melatonin molecules.
+Zaporozhan and Ponomarenko (2010) postulated that minute disturbances in magnetic fields which could translate into influenza symptoms \cite{zaporozhan2010mechanisms}, mediated by the cryptochrome protein-family which has been shown to be geomagnetic receptors \cite{rodgers2009chemical,engels2014anthropogenic,hiscock2016quantum}.
 
 Whatever the cause, global biosphere disturbance theory represents an alternative paradigm in disease spreading dynamics.
-Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model.
-This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations.
-In doing this we shall show that the Gompertz model is incompatible with disease spreading through physical proximity interactions.
-Further, we show that the Gompertz curve emerges when individuals are connected through a common stressor affecting them simultaneously.
+Our goal with this study is to present a novel mathematical bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model.
+This connection represents an alternative to the well-known parametrization given by Richard (1959) \cite{richards1959flexible}, where he augmented standard logistical growth by one parameter.
+This is why it is also sometimes called $\theta$-logistic growth \cite{petroni2020logistic}.
+% \begin{equation}
+% \frac{d}{dt}\ln{Y(t)} = -\beta\left[1 - \left(\frac{Y(t)}{{Y_{\infty}}}\right)^\theta\right],
+% \end{equation}
+% where traditional logistic growth is augmented by a parameter $\theta \in \mathbb R^+$.
 
-\subsection*{Simple communicable disease models}
+ One issue with Richard's parametrization is that the traditional communicable disease models cannot be realistically extrapolated to the Gompertz model, as shown by Wang et \textit{al.} (2012) \cite{wang2012richards}.
+The reason for why Richard's model cannot realistically represent a communicable disease with a Gompertz model is because it would imply that there is ``long-range interaction'' between the susceptible and infected (see \cite{petroni2020logistic} for details), which is counter to the basic assumption that communicable diseases spread through proximity between the susceptible and infected.
+Based on these findings, one would then be forced to conclude that Gompertz growth cannot arise in the context of a communicable disease.
+
+In this paper we will show this incompatibility in another way by creating a model built on higher order interaction terms between the susceptible and the infected groups asymptotically leading to Gompertz growth, but only when the susceptible are the drivers of the growth.
+In other words, Gompertz growth is a causality-reversed communicable disease.
+While this causality-reversal of growth cannot be explained by communicable disease theory, it can however be explained by a biosphere disturbance theory, where a stressor is simultaneously applied to the population.
+This theory does not preclude the possibility of multiple sources of disturbance, some potentially local and interacting with the global source.
+The main idea is that over some possibly geographically constrained area, everyone is exposed to the same source of disturbance simultaneously.
+
+This paper is organized as follows: We first review the basic communicable disease models, before presenting a new stochastic model from which Gompertz growth emerges and thus explains the observed mortality data.
+We then proceed to juxtapose of the logistic model and the Gompertz model in the microscopic domain where multiple entities comprise the macroscopic behavior.
+The final section is dedicated to unifying the communicable disease paradigm with a biosphere disturbance though a new generalized closed form of the Logistic-Gompertzian growth equations.
+
+\section*{Simple communicable disease models: The SIR model family}
 We start by reviewing the basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \cite{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge.
@@
 
- The Gompertzian limit implies a decreasing relative growth rate from the very first time point, which under the SIR model is unlikely given the large pool of susceptible people in the beginning. 
+ The Gompertzian limit implies an exponentially decreasing relative growth rate from the very first time point, which under the SIR model is unlikely given the large pool of susceptible people in the beginning.
  One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
@@
 \begin{equation}
-d\langle\ln{F(t)}\rangle = -\beta \langle\ln{F(t)}\rangle dt,
+d \;\mathbb E[\ln{F(t)}] = -\beta \mathbb E[\ln{F(t)}] dt,
 \end{equation}
-where the bra-ket notation signifies the averaging operation.
+where $\mathbb E$ signifies the averaging operation.
 Then we use the property that for the log-normal quantity $F(t)$, the average of its logarithm is the logarithm of the \emph{median} quantity \cite{petroni2020gompertz}, where we denote the median of $F(t)$ as $M(t)$,
@@
 \ref{eq:microscopicOriginal} wherein the cross-term of $F(t)dW(t)$ goes to zero due to the independence of $F(t)$ and $dW(t)$.
-We also see that by comparing equations \ref{eq:rypdalODE} and \ref{eq:rypdalLimit} with the stochastic counterpart in Eq.
+We also see this by comparing equations \ref{eq:rypdalODE} and \ref{eq:rypdalLimit} with the stochastic counterpart in Eq.
 \ref{eq:microscopicOriginal} that the final growth level is governed by the stressor magnitude,
@@
 \ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$.
-This implies that the central limit theorem applies to the log-domain.
 
 \section*{Comparing logistic and Gompertzian growth in the microscopic domain}
-The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,
+The remarkable observation that the log-transformed domain is the natural one merits closer study.
+In this section, we will present a phenomenological argument that shows how Gompertzian growth implies long-range interactions of entities.
+This argument will be made more rigorous in the proceeding section.
+First, juxtapose the logistic model with the Gompertz model,
 
@@
 
+\section*{Appendix}
+A elaboration of the Gompertz equation in \ref{eq:GompertzODEsimple} is given below, which includes an estimate of the abundance as time goes to infinity,
+\begin{equation}
+\label{eq:GompertzODE}
+\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
+\end{equation} 
+% Redundant definitions for full script
+with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
+\begin{equation}
+\label{eq:gomp_solution}
+Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
+\end{equation}
+where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.
+
 \section*{Supplement}

### 2023-04-03 arxiv/main.tex
respond to comments

@@
 
-Whatever the cause, global biosphere disturbance theory represents an alternative paradigm in disease spreading dynamics.
-Our goal with this study is to present a novel mathematical bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model.
-This connection represents an alternative to the well-known parametrization given by Richard (1959) \cite{richards1959flexible}, where he augmented standard logistical growth by one parameter.
+Whatever the cause, global biosphere disturbance theory represents an alternative paradigm compared to the traditional contact-transmissible contagion theory in the context of disease spreading dynamics.
+Our goal with this study is to present a novel mathematical bridge between these two paradigms which at its core is a connection between the Gompertzian growth model and the logistic growth model.
+Our proposed bridge represents an alternative to the well-known parametrization given by Richard (1959) \cite{richards1959flexible}, where he augmented standard logistical growth by one parameter.
 This is why it is also sometimes called $\theta$-logistic growth \cite{petroni2020logistic}.
@@
  One issue with Richard's parametrization is that the traditional communicable disease models cannot be realistically extrapolated to the Gompertz model, as shown by Wang et \textit{al.} (2012) \cite{wang2012richards}.
-The reason for why Richard's model cannot realistically represent a communicable disease with a Gompertz model is because it would imply that there is ``long-range interaction'' between the susceptible and infected (see \cite{petroni2020logistic} for details), which is counter to the basic assumption that communicable diseases spread through proximity between the susceptible and infected.
-Based on these findings, one would then be forced to conclude that Gompertz growth cannot arise in the context of a communicable disease.
+The reason why Richard's model cannot realistically represent a communicable disease with a Gompertz model is because it would imply that there is ``long-range interaction'' between the susceptible and infected (see \cite{petroni2020logistic} for details), which is counter to the basic assumption that communicable diseases spread through proximity between the susceptible and infected.
+Based on these findings, one would then be forced to conclude that Gompertz growth cannot arise in the context of spread through proximity between the susceptible and infected.
 

### 2023-04-04 arxiv/main.tex
respond to comments2

@@
 
-\author{Matz A. Haugen, \\
-Independent Reseacher, Oslo\\
-\texttt{matzhaugen@gmail.com} \\
-  \And
-  Dorothea Gilbert \\
-  Independent Reseacher, Oslo\\
-}
+% \author{Matz A. Haugen, \\
+% Independent Reseacher, Oslo\\
+% \texttt{matzhaugen@gmail.com} \\
+%   \And
+%   Dorothea Gilbert \\
+%   Independent Reseacher, Oslo\\
+% }
 
@@
 \label{eq:rypdalGamma}
-\gamma(t) = \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] ,
+\gamma(t) = \frac{\beta}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right].
 \end{equation}
-where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$.
 This parametrization is the commonly used Richards growth curve \cite{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \cite{wu2020generalized}.
@@
 \label{eq:rypdalLimit}
-\gamma(t) = \lim_{\theta \rightarrow 0}\frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right]
-= \gamma_{\infty}\ln{\frac{Y_{\infty}}{Y}}
+\gamma(t) = \lim_{\theta \rightarrow 0}\frac{\beta}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right]
+= \beta\ln{\frac{Y_{\infty}}{Y}},
+\end{equation}
+such that 
+\begin{equation}
+\frac{d \ln Y}{dt} = -\beta \ln \frac{Y}{Y_\infty} \label{eq:gompertzODEfirst}.
 \end{equation}
@@
 Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease SIR-like models.
-Under a double-log-transform of Gompertzian growth Eq.
-\ref{eq:gomp_solution}, the growth curve is a simple straight line which is one of its characteristic traits \cite{Levitt2020},
+Under a double-log-transform of Gompertzian growth, the growth curve is a simple straight line which is one of its characteristic traits \cite{Levitt2020},
 \begin{equation}
@@
 \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
-Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq.
-\ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality \cite{Levitt2020},
+
+Meanwhile, even though the Gompertz model can be fit with a 3-parameter model, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality \cite{Levitt2020},
 \begin{equation}
@@
 \end{equation}
-Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters, which plagues more complex models \cite{roda2020difficult}.
+
+The 3-parameter model has one extra degree of freedom which can stabilize the asymptotic abundance estimate, 
+\begin{equation}
+\label{eq:GompertzODE}
+\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
+\end{equation} 
+% Redundant definitions for full script
+with constants $\tilde{Y}$, $\beta$, and $\nu$. The Gompertz growth solution is given as
+\begin{equation}
+\label{eq:gomp_solution}
+Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
+\end{equation}
+where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=\tilde{Y}e^{\nu/\beta}$.
+
+Whether though the 3-parameter model or the 2-parameter linear model, the Gompertz model is relatively parsimonious in its parameters and thus mitigates the possibility of non-identifiability issues, which plagues more complex models \cite{roda2020difficult}.
 
@@
 We are thus prompted to look for another model which can explain the observed mortality patterns.
-For this challenge, we shall first present a stochastic approach and then present a deterministic approach which unifies the SIR family models with the Gompetz curve.
+For this challenge, we shall first present a stochastic approach and then present a deterministic approach which unifies the SIR family models with the Gompertz curve.
 
@@
 \ref{eq:microscopicOriginal} wherein the cross-term of $F(t)dW(t)$ goes to zero due to the independence of $F(t)$ and $dW(t)$.
-We also see this by comparing equations \ref{eq:rypdalODE} and \ref{eq:rypdalLimit} with the stochastic counterpart in Eq.
-\ref{eq:microscopicOriginal} that the final growth level is governed by the stressor magnitude,
-\begin{equation}
-\sigma = 2\gamma_{\infty}\ln{Y_{\infty}}.
-\end{equation}
 
@@
 \begin{equation}
-\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
+\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha x_i = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
 \end{equation}
@@
 
-\section*{Author Contributions}
-M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
-
-\section*{Appendix}
-A elaboration of the Gompertz equation in \ref{eq:GompertzODEsimple} is given below, which includes an estimate of the abundance as time goes to infinity,
-\begin{equation}
-\label{eq:GompertzODE}
-\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
-\end{equation} 
-% Redundant definitions for full script
-with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
-\begin{equation}
-\label{eq:gomp_solution}
-Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
-\end{equation}
-where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.
+% \section*{Author Contributions}
+% M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
 

### 2023-04-04 letterToEditor/letter.tex
respond to comments2

@@
 \recipient{To}{Editor, \\
-               Journal of Mathematical Biology,\\ 
+               Journal of Physics A: Mathematical and Theoretical,\\ 
                }
@@
 
-If you do decide to review this paper you may find the enclosed list of possible referees helpful.
-Some of them are listed in the references of the manuscript.
 Due to the conclusions of our paper, I request that this paper receives more reviews than your average, and at least 3 reviews.
+Even though we have recommended a few reviewers, perhaps you know someone who is better qualified to review, especially since we are submitting to a special section of your journal.
 
@@
 % \end{enumerate}
-\textbf{List of possible referees}
-\begin{enumerate}
-\item Tom Britton, Dean of Mathematics and Physics, Stockholm University, tom.britton@math.su.se
-\item Mark Lewis, marklewis@uvic.ca, Department of Mathematics and Statistics, University of Victoria
-\item Alexandru Hening, Department of Mathematics at Texas A and M University,  ahening@tamu.edu
-\item K-Y. Lam, Mathematics,  the Ohio State University, Columbus, lam.184@osu.edu 
-\item X. Zou, Western University, London, Ontario, xzou@uwo.ca
-\item S. Petrovskii, Mathematical Sciences, University of Leicester, sp237@le.ac.uk
-\item Pierre Magal, Institut de Mathématiques de Bordeaux, Université de Bordeaux, pierre.magal@u-bordeaux.fr 
-\item Vladimir N. Binhi, vnbin@mail.ru, Prokhorov General Physics Institute of the Russian Academy of Sciences, 38 Vavilov St., 119991 Moscow, Russia
-Is familiar with the biological effects due geomagnetic disturbances and fields.
-\item Andrei B. Rubin, rubin@biophysics.msu.ru, Faculty of Biology, Lomonosov Moscow State University, Leninskie Gory 1/12, 119234 Moscow, Russia
-Is familiar with the biological effects due geomagnetic disturbances and fields.
-\item Viacheslav V. Krylov, I.D. Papanin Institute for Biology of Inland Waters Russian Academy of Sciences, Borok, Russian Federation, krylovviacheslav@mail.ru
-Is intimately familiar with geomagnetic disturbances' effect on biology.
-\item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
-\item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
-Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
-\item Robert M May, robert.may@zoo.ox.ac.uk, Professor of Zoology, University of Oxford.
+% \textbf{List of possible referees}
+% \begin{enumerate}
+% % IOP REVIEWERS
+% \item Martin Evans,Professor of Statistical Physics,  https://www2.ph.ed.ac.uk/~mevans/, m.evans@ed.ac.uk
+% \item Prof Ralf Metzler, Institute of Physics & Astronomy
+% University of Potsdam
+% Karl-Liebknecht-Str 24/25, Haus 28
+% D-14476 Potsdam-Golm 
+% \item Michael Victor Berry, U of Bristol, Prof. of Physics, asymptotico(add at bristol.ac.uk) 
+% \item Ronald Dickman, Professor of Physics, UFMG, dickman@fisica.ufmg.br
+% \item Ewa Gudowska-Nowak, ewa.gudowska-nowak[at]uj.edu.pl , 
+% Marian Smoluchowski
+% Institute of Physics,
+% Jagiellonian University
+% \item Shamik Gupta Tata Institute of Fundamental Research, Mumbai, India, shamik.gupta-At-theory.tifr.res.in (Official)
+% \item Alain Barrat, CNRS & Aix-Marseille Université, France , Alain.Barrat@cpt.univ-mrs.fr
+% \item Daniel ben-Avraham, dbenavra@clarkson.edu, Professor of Physics, Clarkson University
+% \item Tom Britton, Dean of Mathematics and Physics, Stockholm University, tom.britton@math.su.se
+% \item Mark Lewis, marklewis@uvic.ca, Department of Mathematics and Statistics, University of Victoria
+% \item Alexandru Hening, Department of Mathematics at Texas A and M University,  ahening@tamu.edu
+% \item K-Y. Lam, Mathematics,  the Ohio State University, Columbus, lam.184@osu.edu 
+% \item X. Zou, Western University, London, Ontario, xzou@uwo.ca
+% \item S. Petrovskii, Mathematical Sciences, University of Leicester, sp237@le.ac.uk
+% \item Pierre Magal, Institut de Mathématiques de Bordeaux, Université de Bordeaux, pierre.magal@u-bordeaux.fr 
+% \item Vladimir N. Binhi, vnbin@mail.ru, Prokhorov General Physics Institute of the Russian Academy of Sciences, 38 Vavilov St., 119991 Moscow, Russia
+% Is familiar with the biological effects due geomagnetic disturbances and fields.
+% \item Andrei B. Rubin, rubin@biophysics.msu.ru, Faculty of Biology, Lomonosov Moscow State University, Leninskie Gory 1/12, 119234 Moscow, Russia
+% Is familiar with the biological effects due geomagnetic disturbances and fields.
+% \item Viacheslav V. Krylov, I.D. Papanin Institute for Biology of Inland Waters Russian Academy of Sciences, Borok, Russian Federation, krylovviacheslav@mail.ru
+% Is intimately familiar with geomagnetic disturbances' effect on biology.
+% \item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
+% \item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
+% Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
+% \item Robert M May, robert.may@zoo.ox.ac.uk, Professor of Zoology, University of Oxford.
 \end{enumerate}

### 2023-04-05 arxiv/main.tex
subtitle change

@@
 
-\subsection*{A closed-form expression for a unified model}
+\subsection*{A closed-form expression for a unified Gompertz-Logistic model}
 Finally, we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.

### 2023-04-05 arxiv/main.tex

@@
 Finally, we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
-Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with
+Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with the well-known identity
 
@@
 \end{equation}
-Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
+Using standard integrals we see that this sum amounts to a Gompertz growth term adjusted by a Hypergeometric function,
 \begin{eqnarray*}
@@
 \ref{eq:rypdalGamma} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$.
-The maximally coherent scenario implies that all units in the network are exposed to the stressor simultaneously.
+Used in a mortality model, the maximally coherent scenario implies that all units in the network are exposed to the stressor simultaneously.
 Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.

### 2023-08-03 arxiv/main.tex
add new draft

@@
 Second, we show that for the Gompertz curve to emerge, the disturbance has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
-Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model family is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction.
+Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model family is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite interaction.
 Crucially, this augmentation must be accompanied by a causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people.

### 2023-08-03 arxiv2/new_draft.tex
add new draft

\documentclass{article}
% \documentclass{sn-jnl}
% \documentclass[review]{elsarticle}

\usepackage{arxiv}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}   % Can be removed after putting your text content
\usepackage{graphicx}
\usepackage[square,compress,numbers]{natbib}
\usepackage{doi}
\usepackage{amsmath}
\usepackage{float}
\usepackage{lineno}
% \linenumbers

% \def\bibcommenthead{}%

\newcommand{\insertPdfFig}[3]{
  \begin{figure}[H]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}

\newcommand{\insertSmallPdfFig}[3]{
  \begin{figure}[h]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}

% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
\title{Gompertzian growth as infinitely correlated growth}

% \graphicspath{{../figures/}}
%\date{September 9, 1985} % Here you can change the date presented in the paper title
%\date{}          % Or removing it

\author{Matz A. Haugen, \\
Independent Reseacher, Oslo\\
\texttt{matzhaugen@gmail.com} \\
  \And
  Dorothea Gilbert \\
  Independent Reseacher, Oslo\\
}

% Uncomment to remove the date
%\date{}

% Uncomment to override  the `A preprint' in the header
%\renewcommand{\headeright}{Technical Report}
%\renewcommand{\undertitle}{Technical Report}
\renewcommand{\shorttitle}{On the nature of the Gompertz}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={A template for the arxiv style},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={},
pdfkeywords={},
}

\begin{document}
\maketitle

\begin{abstract}
Logistic and Gompertzian growth has traditionally been connected by exponentiating the logistic abundance variable with an extra parameter, known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertzian growth is still little understood. We show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth, where the latter involves a network of particles or entities. The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself to a log transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth. We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range communication. This conclusion adds to the interpretation of biological systems that undergo Gompertzian growth where long-range communication could be interpreted as the source of growth being external to the system of particles, like an external field that stimulates growth simultaneously across all particles. 
\end{abstract}

% keywords can be removed
\keywords{Gompertz \and Coherence \and Network Analysis \and Stochastic}

\section{Introduction}

In 1825, Gompertz published a paper on the nature of mortality where he discovered that from adulthood and onward ``the power to avoid destruction'' decreases exponentially with age. Equivalently, mortality rates increase in geometric progression for arithmetic increases in age, an observation now commonly referred to as Gompertz' law of mortality \cite{Gompertz1825}. Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time \cite{makeham1889further}.

This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exemplifying the Gompertz growth law as ubiquitous in nature. The idea that the Gompertz model can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}. Gompertz law has also been shown to apply to societal phenomena like growth in railway traffic, and the demand for goods and services, and sales of tobacco \cite{prescott1922law,peabody1924growth}.

Alongside the many contexts wherein Gompertz' law has been observed, an almost equal number of efforts have been made to interpret and derive this law from various phenomenological principals \cite{bajzer1997mathematical}. This effort has often been accompanied by a phenomenological comparison of another popular growth model called logistic growth, or Verhulst growth, where ``the power to avoid destruction'' decreases linearly with time, as opposed to decreasing exponentially with time as seen with the Gompertz model. In a study comparing the Gompertz and the logistic grwoth models, Petroni et al. (2020) \cite{petroni2020gompertz} show that Gompertz growth can be interpreted as maximally coherent growth based on long-range communication or correlation, while logistic growth is minimally coherent without long-range behavior. With this interpretation, one is lead to postulate that the Gompertz function could be a result of a global or external stimuli while the logistic function is a result of internal or local interaction. As such, the logistic function has been popular both for communicable disease models \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately} and and for predator-prey theory \cite{berryman1992orgins}, where disease or injury spreads through pairwise interactions.

A number of studies in the biological context have argued that the Gompertzian model could be sourced or activated from an external factor which is ubiquitously and instantaneously introduced to the system, e.g. as a toxic agent or an environmental stressor to which the system gradually adapts \cite{neafsey1988gompertz,neafsey1989gompertz,thompson1990biphasic}. In the original case of mortality rates as a function of age, this interpretation would imply that time itself is the environmental ubiquitous stressor. Other environmental stressors like chronic radiation exposure have been observed to shift the Gompertz curve, an observation also theoretically developed by Sacher (1956) (\cite{sacher1956statistical}). Sacher modeled the toxic stressor as a stochastic perturbation to physiologic state of the population wherein beyond a certein point of perturbation in a given entity would result in its death. The interpretation of the Gompertz curve as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al. (2014) \cite{de2014stochastic} who arrived at the same result without alluding to a threshold value beyond which death was inevitable, but rather by letting the stochastic perturbation be the source of growth or decay.

Another framework used to show how the Gompertz can result from an external or ubiquitous stressor is chemical reaction theory. In reaction theory, a set of reactants combine to produce products, possibly in the presence of a catalyst. Morkov (2019) \cite{markov2019reaction} showed that the Gompertz can be interpreted as growth under a catalyst whose effect diminishes at a rate independent of the growth rate of the reactant. In other words, the catalyst can be considered as an external and ubiquitous stressor or catalyst to the reactants, perfectly in line with the ubiquitous stochastic stressor model from Sacher and De Lauro et al. On the other hand, if the logistic growth equation is recast into the reaction network equivalent, the catalyst is diminished by the growth of the reactant, which suggests that the source of growth is internal or at least dependent on the internal state. So even in reaction theory the gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions.

In this paper, we will elaborate on this idea that the Logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will use a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by Petroni et al. (2020), \citet{tjorve2017use}, and \cite{wang2012richards}). In particular, instead of considering a real number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $n$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the references, this is in line both with experimental and theoretical results. In presenting this argument, we will present a novel bridge between the Logistic and the Gompertz model which offers both a mathematical and biological foundation to source of growth.

It is worth noting that the notion of Gompertz growth as maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}. Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schroedinger equation. From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory. This adds to our exhibits that the Gomepertz model can be interpreted as non-local in origin. Coherent quantum states was originally formulated in the context of electromagnetic fields. Perhaps some of the biological macroscopic growth phenomena references are also due to such fields, for example in the context of cancer cell progression.

Finally, the Gompertz model has been shown by Wang et al. \cite{wang2012richards} to not originate from a communicable disease paradigm where only local disease spread is possible. They used the common Susceptible-Infected-Recovered model by Kermack and McKendrick \cite{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter value range with the communicable disease assumption does not cover the parameter value taken for Gompertz to emerge. This would imply that any mortality that shows a Gompertzian characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.

Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which will offer an alternative to the traditional Richard's model. Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena. This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor. We proceed with a discussion section and end with a conclusion.

% \begin{enumerate}

%   \item Introduce gompertz and logistic growth

%   \item Discuss usages in litterature, cancer, mortality curves
%   \item Mention Bajzer 2020's effort to derive macroscopic foundation for Gompertz growth, but leaving out a microscopic foundation.
%   \item Reaction theory formulates gompertz as the interaction between growth variable $x$ and environmental ``catalyst'' $s$, where s is exponentially decreasing in size exactly analogous with an external source of growth. See https://biomath.math.bas.bg/biomath/index.php/biomath/article/view/j.biomath.2019.04.167/pdf
%   \item Reaction theory could also be self contained if we let s to s1, s2, .. going to infinity. Since this would imply long range correlation, the only explenation within our current knowledge of nature is to posit a field.
%   \item Molski interprets spatial gompertz behavior as a result of an EM field in space for molecular dynamics. The temporal equivalent would be to interpret the source of growth to be stressor field in time, equal across space.
%   \item De Lauro discusses Gompertz growth as a stochastic phenomena where the source of growth is a random and ubiquitous perturbation which aligns with the source of growth as external.
%   \item Patroni et al. (2020) \cite{petroni2020gompertz} show that gompertz growth can be interpreted as maximally coherent growth based on long-range cooperation or sharing of resources. In the extreme case when the long-range behavior is operating between all nodes, growth can be described as due to an external source inducing identical behavior across all nodes, which manifests itself as coherence.
%   \item Gompertz himself considered time as the ubiquitous external stressor in the original formulation of the Gompertz model.
%   \item Wang et al. (2012) \cite{wang2012richards} show that gompertz growth cannot emerge from intrinsic interactions only in a generalized logistic growth using the richards parametrization and a communicable disease paradigm.
%   \item The crux of our argument lies in showing that gompertz can be interpreted as an infinitely interacting growth equation, implying that growth is either communicating instantaneously, or triggered by an external source. And since the former is as odds with our current physical understanding but also perhaps overlapping with the second interpretation, we claim that the latter is the ostensible root of Gompertz growth. 
% \end{enumerate}

\section{Macroscopic growth}
\label{sec:macro}

First consider a system that grows according to logistic growth. If we let the variable $X(t)$ denote the abundance of a quantity in a normalized system where its maximum size is 1, the equation that determines its size as a function of time is
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)).
\end{equation}
This is commonly interpreted as the right hand side representing ``resources available for growth'' which decreases linearly with respect to the relative abundance's growth rate on the left hand side. One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
\begin{equation}
\label{eq:Rich}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)^{\theta}),
\end{equation}
also called $\theta$-logistic growth or Richards' model. Petroni et al. (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of co-operation in the microscopic domain. In the limit where the system becomes \textit{maximaly cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
 \begin{equation}
 \label{eq:Gomp}
  \frac{1}{X}\frac{d}{dt} X(t) = - \beta \ln X(t).
\end{equation}
In this limit, the relative growth goes to infinity as the system approaches $t=0$, which could be seen as a verification of the maximally coherent characteristic of the system. This definition of coherence in the context of the gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.

Another way to generalize the logistic equation is to augment the resource terms to higher orders as follows,
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta \left[(1 - X(t)) + \frac{1}{2}(1 - X(t))^2 + ... \frac{1}{K}(1 - X(t))^K\right],
\end{equation}
where fractional prefactors to each term will be justified in the sequel.
After some algebra, this augmented logistic equation reduces to a closed form expression involving Gauss' Hypergeometric function, ${}_1F_{2}$, viz.
\begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{X(t)} = - \frac{{}_{1}F_{2}({1,K+1,K+2;1-X(t)})}{K+1}(1-X(t))^{K+1} - \ln{X(t)}.
 \end{equation}

 In the context of higher order terms, the equivalent of $\theta\rightarrow 0$ in $\theta$-logistic growth is that the higher order terms go to infinity, i.e. $K \rightarrow \infty$, which again reduces to the Gompertz equation by virtue of the Hypergeometric term going to zero.

 The emergence of Gompertz growth from higher order logistic growth can be seen as another way to encode non-linear effects that were seen as coherence in accordance with Petroni and Molski. In practice, the augmented logistic model has a faster growth rate decay than the familiar $\theta$-logistic growth \ref{fig:hypergeometric_plain}. If we interpret speed of relative growth decay as coherence, the augmented logistic model would be more coherent than the $\theta$-logistic model for a given starting value.

\insertPdfFig{hypergeometric_plain}{Comparison relative growth rate in the Richards $\theta$-logistic model (Eq. \ref{eq:Rich}), Gompertz model (Eq. \ref{eq:Gomp}), and the augmented causality-reversed SIR model (Eq. \ref{eq:modLogistic}), all as a function of the abundance variable. Both dependent and independent variables are normalized so that final abundance (or size) is set to unity. The Richards $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}

In spite of the connection of the Gompertz model with higher order terms in a logistic model, it is still not clear whether the source of growth is external or internal. The right hand side ``resource'' term is not specific enough so as to specify the location of the source. However, one could argue that the since the Gompertz model has asymptoticaly infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal that initiates the growth and subsequently decays exponentially just like a memoryless process. To further explore the source of growth we turn to the microscopic domain.

\section{Microscopic growth}
\label{sec:micro}

\subsection{Microscopic Logistic growth}
Consider $N$ entities each represented by a set of independent and identically distributed random abundance variables, $x_i(t), \quad i=1, ..., N$, which are unitless and take values between 0 and 1, where 0 represents no realized growth while 1 represents the maximum potential growth having been realized and $\mu$ representing the mean of the distribution. With these variables it will also be convenient to also define an \textit{abundance potential} variable, $s_i = 1 - x_i$, where we omit the temporal variable for ease of exposition.

Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables. Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ which a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = x_i\sum_j{a_{ij}}s_{j} \quad \forall i.
\end{equation}
First note that for a single pair this would amount to traditional logistic growth. We choose to work in a unitless domain without parameters for the interaction strength and carrying capacity for simplicity of exposition. These can be added without changing the conclusions of these arguments.

Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and give growth potential to a single possibly different entity}. Using this rule and rearranging and summing across all entities,
\begin{equation}
\label{eq:modSIR2}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
\end{equation}

With our assumption of the variables being independent and identically distributed, we obtain the simplest form of logistic growth of the geometric mean of the ensemble driven by the sample mean of the inverse abundance, or the \textit{potential abundance}, 
\begin{equation}
\label{eq:modSIR3}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} = \frac{1}{N}\sum_i^N s_i 
\end{equation}

As N grows, both of geometric and the arithmetic means converge in expectation to their respective distibutial parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles. Specifically, letting $\tilde{x}$ represent the sample geometric mean of $x_i$ and $\bar{x}$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have 
\begin{eqnarray*}
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \tilde{x} &=& \lim_{N\rightarrow \infty}(1 - \bar{x}) \\
  \frac{d}{dt} \ln \gamma &=&(1 - \mu)
  % \frac{d}{dt} \ln m &=&- \ln \bar{s},
\end{eqnarray*}

We further make special note that logistic growth emerges from pairwise interactions only. This does not mean that the logistic growth cannot emerge from other microscopic settings but it nevertheless provides one possible scenario under which the logistic curve can emerge. We often see logistic growth in economic growth indicators where pairwise interactions are the dominant form of interaction (see e.g. ). Now, as we shall allow higher order interactions we will see the Gompertz curve emerge.

\subsection{Microscopic Gompertz growth}

Instead of limiting the growth equation to pairwise interactions, we allow higher order interactions in the form of triplets, quadruplets, and so on. Analogous to the pairwise exclusitivity rule, each entity can again only participate in one interaction group at any given order and any given direction. For example, any triplet can only interact with one other triplet and have exclusive posession of its entities at the triplet level. This means that we consider the geometric mean of each $n-$group as follows.
\begin{equation}
\label{eq:modSIR4}
\frac{d}{dt} \ln x_{i_1} x_{i_2} \dots x_{i_n} = s_{j_1}s_{j_2}\dots s_{j_n}, 
\end{equation}
where the indeces $i$ and $j$ represent the higher order interaction between the mutually exclusive groups of entities for each $n$ from 1 and up to $N/2$. The higher orders are capped at $N/2$ since that would be the largest number of entities in a group that could interact with an identically sized group, viz.
\begin{equation}
\label{eq:MicroGomp0}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} s_{j_1}s_{j_2}\dots s_{j_{N/2}},
\end{equation}
where $S_k$ is the set of all mutually exclusive $k-$sized groups. In each of the sums the total number of terms is equal to the total number of groups which at each order is $N/k$ where $k$ is the interaction order. Thus, if we divide the right hand side by $N$, we can identify a convergence property as both $k$ and $N$ grows:
\begin{enumerate}
\item As the number of entities, N, grows each term will converge in expectation to higher powers of the mean of $s_i$, multiplied by the prefactor $1$, $1/2$, $1/3$, etc., 
\begin{equation}
\label{eq:MicroGomp11}
\lim_{N\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = \frac{1}{k} \mu^{k}.
\end{equation}
\item As the higher order interaction terms, k, grow
\begin{equation}
\label{eq:MicroGomp12}
\lim_{k\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = 0
\end{equation}
\end{enumerate}
Thus, as each higher order term is composed of a decreasing number of summations and an increasing number of multiplications, the product of $s_i$ goes to zero since each random variable is $iid$ and bounded between 0 and 1. 
Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic and the geometric means of the microscopic entities
\begin{eqnarray*}
\label{eq:MicroGomp13}
\lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\\
\frac{d}{dt} \ln \gamma &=&  \mu + \frac{1}{2} \mu^2 + ...\\
\frac{d}{dt} \ln \gamma &=& - \ln (1-\mu).
\end{eqnarray*}

We finally note that for the right hand side to exhibit asymptotically infinite growth at the onset of growth, it is not only sufficient but also necessary to have higher order terms. This is seen by virtue of the infinite polynomial representation of the logarithm. This implies that Gompertz growth cannot result from pairwise interactions only under the assumptions presented in this model.

\section{Discussion}
\label{sec:discussion}

We thus see that Gompertz emerges from a group of entities interacting at all orders of interaction. An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be external in nature or the internal source communicates instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}. With an external source, entities need not be interacting with each other but rather respond to an external stimuli, like an external field. Examples of where models are already given by De Lauro (2014) and Markov (2019).

De Lauro (2014)\cite{de2014stochastic} consider the source of growth to be stochastic in nature, and while they do not discuss the external nature of such a source, their model does not preclude it. More generally, with growth as stochastic it is also ubiquitously present uniformly across all entities from the very onset of growth.

Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Laura's model.

With an internal source of growth a potential mediating field could be the electromagnetic field. While en external source also could be resultant from the electromagnetic field, it could also be comming from a ubiquitous stressor like air pollution, oxygen deprivation, or even instant psychologically induced sources.

Another way to see growth as external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case. To see this, swap the role of $x_i$ and $s_i$ on the right hand side Equation \ref{eq:modSIR} to see that the same result can be obtained under the exclusitivity condition.

This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes. But becuase higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading. In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.

It is worth noting that the original formilation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age. He noticed that mortality rates increased at a geometric rate with age. In this context the ubiquitous stressor postulated above is of course time itself, clearly an external source. In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field which in this case is time.

\section{Conclusion}

We have shown how Gompertz growth emerges both at the microscopic and the macroscopic level, where a novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function. This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance.

Macroscopic, hypergeometric, novel bridge.

Microscopic, mutually exclusive higher order interactions lead to gompertz.

This adds to the understanding of gompertz growth as growth initiated externally, and has close ties with Molski's quantum mechanical interpretation of gompertz growth as a coherent system which exhibits long range cooperation.

\section*{Competing interests}
The authors declare no competing interests.

% \section*{Author Contributions}
% M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.

\section*{Supplement}

% Estimating the parameters using real data requires some consideration around the variance of the growth variable as a function of time. This variance will in turn inform the estimation process by considering a weighted minimization of the residuals of the observations regressed onto the model. In the stochastic Gompertz process postulated in Section \ref{seq:alt}, the process under a log-transformation is the Ornstein-Uhlenbeck process which has a closed from expression for its transition probability, and thus its variance as a function of time. This variance can then be used in a weighted least squares estimation technique or any other weighted technique. In the original domain there is also a closed form expression available but it is somewhat more convoluted and given that observations in this domain are visually quite homoscedastic with equal variance across time (Fig \ref{fig:Gompertz_vs_SIRD} and Tables \ref{tab:gomp} and \ref{tab:sird}), this is expression and consideration is omitted. But for the log-domain, observations exibit changing variance with time.

% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\cite{risken1996fokker},
% \begin{equation}
% P(z, t|z', t') = \sqrt{\frac{\beta}{2\pi \sigma (1 - e^{-2\beta (t-t')})}}\exp{\left[ -\frac{\beta (z - e^{-\beta(t-t')}z')^2}{2\sigma(1-e^{-2 \beta(t-t')})} \right]}.
% \end{equation}

% If we consider $t$ the time at which the stationary distribution is reached and that this stationary distribution is given as a delta function, variance of the growth will be given as
% \begin{equation}
% \text{Var[Z(t)|t']} = \sigma(1-e^{-2 \beta(t-t')}).
% \end{equation}

% Because weights on observations in a regression are invariant to multiplication we can let $\sigma=1$. The slope coefficient would have to be estimated either by a standard least squares or through the 3-parameter model in (\ref{eq:GompertzODE}).

% \insertPdfFig{Gompertz_vs_SIRD_gompfit_log_domain}{Log transformed cumulative diseased plotted against its time derivative after $Y(t) / Y_{max}>0.005$, with a straight line fit regressing onto the $50^{th}$ quantile.}

%  Carletti (left) transformation to our notation (right)
% gamma - eta
% eta - xi
% beta - kappa

\bibliographystyle{abbrvnat}
\bibliography{references2}  %%% Uncomment this line and comment out the ``thebibliography'' section below to use the external .bib file (using bibtex) .

%%% Uncomment this section and comment out the \bibliography{references} line above to use inline references.
% \begin{thebibliography}{1}

%   \bibitem{kour2014real}
%   George Nour and Raid Saabne.
%   \newblock Real-time segmentation of on-line handwritten arabic script.
%   \newblock In {\em Frontiers in Handwriting Recognition (ICFHR), 2014 14th
%       International Conference on}, pages 417--422. IEEE, 2014.

%   \bibitem{kour2014fast}
%   George Nour and Raid Saabne.
%   \newblock Fast classification of handwritten on-line arabic characters.
%   \newblock In {\em Soft Computing and Pattern Recognition (SoCPaR), 2014 6th
%       International Conference of}, pages 312--318. IEEE, 2014.

%   \bibitem{hadash2018estimate}
%   Guy Hadash, Einat Nermany, Boaz Carmeli, Ofer Lavi, George Nour, and Alon
%   Jacovi.
%   \newblock Estimate and replace: A novel approach to integrating deep neural
%   networks with existing applications.
%   \newblock {\em arXiv preprint arXiv:1804.09028}, 2018.

% \end{thebibliography}

\end{document}

### 2023-08-05 arxiv2/new_draft.tex
finalize first new draft

@@
 % \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
-\title{Gompertzian growth as infinitely correlated growth}
+\title{Gompertzian growth as infinitely correlated non-local growth}
 
@@
 \begin{abstract}
-Logistic and Gompertzian growth has traditionally been connected by exponentiating the logistic abundance variable with an extra parameter, known as $\theta$-logistic growth or the Richards model for growth.
+Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic abundance variable with an extra parameter, known as $\theta$-logistic growth or the Richards model for growth.
 In spite of this bridge, the biological foundation of Gompertzian growth is still little understood.
-We show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth, where the latter involves a network of particles or entities.
+We show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth, where the latter involves a network of entities.
 The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented.
-For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself to a log transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth.
-We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range communication.
-This conclusion adds to the interpretation of biological systems that undergo Gompertzian growth where long-range communication could be interpreted as the source of growth being external to the system of particles, like an external field that stimulates growth simultaneously across all particles.
+For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth.
+We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range or non-local communication.
+This conclusion adds to the interpretation of biological systems that undergo Gompertzian growth where long-range communication could be interpreted as the source of growth being external to the individual particles, like a field that stimulates growth simultaneously across all entities.
+In situations of Gompertzian decay this field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, while in situations of Gompertzian growth the field could be an internal or external electromagnetic field.
 \end{abstract}
@@
 The idea that the Gompertz model can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}.
-Gompertz law has also been shown to apply to societal phenomena like growth in railway traffic, and the demand for goods and services, and sales of tobacco \cite{prescott1922law,peabody1924growth}.
+Gompertz law has also been shown to apply to societal phenomena like growth in railway traffic, and the demand for goods and services, and sales of tobacco \cite{olshansky1997ever,prescott1922law,peabody1924growth}.
 
@@
 This effort has often been accompanied by a phenomenological comparison of another popular growth model called logistic growth, or Verhulst growth, where ``the power to avoid destruction'' decreases linearly with time, as opposed to decreasing exponentially with time as seen with the Gompertz model.
-In a study comparing the Gompertz and the logistic grwoth models, Petroni et al.
-(2020) \cite{petroni2020gompertz} show that Gompertz growth can be interpreted as maximally coherent growth based on long-range communication or correlation, while logistic growth is minimally coherent without long-range behavior.
+In a study comparing the Gompertz and the logistic growth models, Petroni et al.
+(2020) \cite{petroni2020logistic} show that Gompertz growth can be interpreted as maximally coherent growth based on long-range communication or correlation, while logistic growth is minimally coherent without long-range behavior.
 With this interpretation, one is lead to postulate that the Gompertz function could be a result of a global or external stimuli while the logistic function is a result of internal or local interaction.
@@
 Other environmental stressors like chronic radiation exposure have been observed to shift the Gompertz curve, an observation also theoretically developed by Sacher (1956) (\cite{sacher1956statistical}).
-Sacher modeled the toxic stressor as a stochastic perturbation to physiologic state of the population wherein beyond a certein point of perturbation in a given entity would result in its death.
+Sacher modeled the toxic stressor as a stochastic perturbation to physiologic state of the population wherein beyond a certain point of perturbation in a given entity would result in its death.
 The interpretation of the Gompertz curve as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al.
@@
 
-In this paper, we will elaborate on this idea that the Logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
+In this paper, we will elaborate on this idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
 We will use a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
-by Petroni et al.
-(2020), \citet{tjorve2017use}, and \cite{wang2012richards}).
+by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}).
 In particular, instead of considering a real number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model.
@@
 As seen in the references, this is in line both with experimental and theoretical results.
-In presenting this argument, we will present a novel bridge between the Logistic and the Gompertz model which offers both a mathematical and biological foundation to source of growth.
+In presenting this argument, we will present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to source of growth.
 
 It is worth noting that the notion of Gompertz growth as maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}.
-Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schroedinger equation.
+Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation.
 From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory.
-This adds to our exhibits that the Gomepertz model can be interpreted as non-local in origin.
-Coherent quantum states was originally formulated in the context of electromagnetic fields.
-Perhaps some of the biological macroscopic growth phenomena references are also due to such fields, for example in the context of cancer cell progression.
+This adds to our exhibits that the Gompertz model can be interpreted as non-local in origin.
+Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize).
+Perhaps some of the biological macroscopic growth phenomena references are also due to such fields.
+For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.
+The limb-regeneration process of the salamander is an example of such a phenomena.
 
-Finally, the Gompertz model has been shown by Wang et al.
+In fact, it has been shown that cancer malignancies are mostly negative in polarity with respect to the rest of the body, while nonmalignant tumor pathology is mostly positive \cite{langman1949technique}.
+One is then led to the interpretation that certain biological growth processes are coupled with an electromagnetic field of specific character.
+Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic or cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.
+The disruptive effect also extends to other biological processes, as shown by Wever and Brown's experiments on biological endogenous cycles lie the circadian cycle through manipulation of the environmental electromagnetic field \cite{brown1976biological,wever1974elf}.
+
+Finally, on the topic of disease modelling, the Gompertz model has been shown by Wang et al.
 \cite{wang2012richards} to not originate from a communicable disease paradigm where only local disease spread is possible.
-They used the common Susceptible-Infected-Recovered model by Kermack and McKendrick \cite{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter value range with the communicable disease assumption does not cover the parameter value taken for Gompertz to emerge.
-This would imply that any mortality that shows a Gompertzian characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.
+They used the common Susceptible-Infected-Recovered model by Kermack and McKendrick \cite{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter value range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
+This would imply that any mortality patterns that shows a Gompertzian characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.
 
@@
 
-\section{Microscopic growth}
+\section{Microscopic Growth}
 \label{sec:micro}
 
-\subsection{Microscopic Logistic growth}
+\subsection{Microscopic Logistic Growth}
 Consider $N$ entities each represented by a set of independent and identically distributed random abundance variables, $x_i(t), \quad i=1, ..., N$, which are unitless and take values between 0 and 1, where 0 represents no realized growth while 1 represents the maximum potential growth having been realized and $\mu$ representing the mean of the distribution.
@@
 
-As N grows, both of geometric and the arithmetic means converge in expectation to their respective distibutial parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles.
+As N grows, both of geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles.
 Specifically, letting $\tilde{x}$ represent the sample geometric mean of $x_i$ and $\bar{x}$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have
@@
 
-We further make special note that logistic growth emerges from pairwise interactions only.
+We make special note that logistic growth emerges from pairwise interactions only.
 This does not mean that the logistic growth cannot emerge from other microscopic settings but it nevertheless provides one possible scenario under which the logistic curve can emerge.
-We often see logistic growth in economic growth indicators where pairwise interactions are the dominant form of interaction (see e.g.
-).
 Now, as we shall allow higher order interactions we will see the Gompertz curve emerge.
 
-\subsection{Microscopic Gompertz growth}
+\subsection{Microscopic Gompertz Growth}
 
 Instead of limiting the growth equation to pairwise interactions, we allow higher order interactions in the form of triplets, quadruplets, and so on.
-Analogous to the pairwise exclusitivity rule, each entity can again only participate in one interaction group at any given order and any given direction.
-For example, any triplet can only interact with one other triplet and have exclusive posession of its entities at the triplet level.
+Analogous to the pairwise exclusivity rule, each entity can again only participate in one interaction group at any given order and any given direction.
+For example, any triplet can only interact with one other triplet and have exclusive possession of its entities at the triplet level.
 This means that we consider the geometric mean of each $n-$group as follows.
@@
 \end{equation}
-where the indeces $i$ and $j$ represent the higher order interaction between the mutually exclusive groups of entities for each $n$ from 1 and up to $N/2$.
+where the indices $i$ and $j$ represent the higher order interaction between the mutually exclusive groups of entities for each $n$ from 1 and up to $N/2$.
 The higher orders are capped at $N/2$ since that would be the largest number of entities in a group that could interact with an identically sized group, viz.
@@
 
-\section{Discussion}
+\section{Conclusion}
 \label{sec:discussion}
 
-We thus see that Gompertz emerges from a group of entities interacting at all orders of interaction.
-An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be external in nature or the internal source communicates instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}.
-With an external source, entities need not be interacting with each other but rather respond to an external stimuli, like an external field.
-Examples of where models are already given by De Lauro (2014) and Markov (2019).
+We have shown how Gompertz growth emerges both at the microscopic and the macroscopic level as an infinitely correlated non-local effect.
+A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model derived from first principals.
+This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance.
+On the microscopic level, the Gompertz model emerges through an exclusivity rule at every order of interaction.
+As a corollary, we add another exhibit that show how the Logistic function is the result of local growth processes with pairwise-interactions only.
 
-De Lauro (2014)\cite{de2014stochastic} consider the source of growth to be stochastic in nature, and while they do not discuss the external nature of such a source, their model does not preclude it.
+An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}.
+With a such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli.
+Examples of such models are already given by De Lauro (2014), Sacher (1956) and Markov (2019).
+\citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic in nature, and while the former do not discuss the external nature of such a source, their model does not preclude it.
 More generally, with growth as stochastic it is also ubiquitously present uniformly across all entities from the very onset of growth.
+Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Lauro and Sacher's models.
 
-Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Laura's model.
+For growth within biological organisms, it has been well established that the mediating field is the electromagnetic field as discussed by R.O.
+Becker (see \citet{becker1984electromagnetic} and references within), which suggests that the accompanying growth or decay pattern in these processes follows the Gompertz model as the result of a ubiquitous activating field.
+For growth or decay in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts.
 
-With an internal source of growth a potential mediating field could be the electromagnetic field.
-While en external source also could be resultant from the electromagnetic field, it could also be comming from a ubiquitous stressor like air pollution, oxygen deprivation, or even instant psychologically induced sources.
-
-Another way to see growth as external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case.
-To see this, swap the role of $x_i$ and $s_i$ on the right hand side Equation \ref{eq:modSIR} to see that the same result can be obtained under the exclusitivity condition.
+Another way to see the source growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case.
+To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same result can be obtained under the exclusivity condition.
+Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model.
 
 This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes.
-But becuase higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading.
+But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading.
 In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
 
-It is worth noting that the original formilation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age.
+It is worth noting that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age.
 He noticed that mortality rates increased at a geometric rate with age.
 In this context the ubiquitous stressor postulated above is of course time itself, clearly an external source.
-In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field which in this case is time.
-
-\section{Conclusion}
-
-We have shown how Gompertz growth emerges both at the microscopic and the macroscopic level, where a novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function.
-This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance.
-
-Macroscopic, hypergeometric, novel bridge.
-
-Microscopic, mutually exclusive higher order interactions lead to gompertz.
-
-This adds to the understanding of gompertz growth as growth initiated externally, and has close ties with Molski's quantum mechanical interpretation of gompertz growth as a coherent system which exhibits long range cooperation.
+In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field.
 

### 2023-08-05 arxiv2/new_draft.tex
fix

@@
  The emergence of Gompertz growth from higher order logistic growth can be seen as another way to encode non-linear effects that were seen as coherence in accordance with Petroni and Molski.
-In practice, the augmented logistic model has a faster growth rate decay than the familiar $\theta$-logistic growth \ref{fig:hypergeometric_plain}.
+In practice, the augmented logistic model has a faster growth rate decay than the familiar $\theta$-logistic growth (Figure \ref{fig:hypergeometric_plain}).
 If we interpret speed of relative growth decay as coherence, the augmented logistic model would be more coherent than the $\theta$-logistic model for a given starting value.

### 2023-08-05 arxiv2/new_draft.tex
fix

@@
 The right hand side ``resource'' term is not specific enough so as to specify the location of the source.
-However, one could argue that the since the Gompertz model has asymptoticaly infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal that initiates the growth and subsequently decays exponentially just like a memoryless process.
+However, one could argue that the since the Gompertz model has asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal that initiates the growth and subsequently decays exponentially just like a memoryless process.
 To further explore the source of growth we turn to the microscopic domain.

### 2023-08-07 arxiv2/new_draft.tex
add gauss and exponential

@@
 We finally note that for the right hand side to exhibit asymptotically infinite growth at the onset of growth, it is not only sufficient but also necessary to have higher order terms.
-This is seen by virtue of the infinite polynomial representation of the logarithm.
-This implies that Gompertz growth cannot result from pairwise interactions only under the assumptions presented in this model.
+This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions only under the assumptions presented in this model.
+
+\section{Unifying the Gaussian and Exponential curves with the Logistic and the Gompertz curves}
+
+Up until now we have seen how logistic growth emerges from pairwise correlated growth while Gompertz growth emerges from infinitely correlated growth.
+These are both growth modes where there is some degree of connectivity between the entities.
+But what if there is no connectivity between the entities but each entity is independent?
+In such a case we can formulate a growth equation where the relative growth is only linearly dependent on time, where we make a slight modification and model the \textit{potential abundance}, $s(t)$, instead of the actual abundance,
+\begin{equation}
+  \frac{d}{dt}s(t) = - k t s(t),
+\end{equation}
+or 
+\begin{equation}
+  \frac{d}{dt}\ln s(t) = - k t,
+\end{equation}
+for some constant $k>0$.
+
+Solving this equation yields the familiar Gaussian function, 
+\begin{equation}
+  s(t) = A e^{ - k t^2 / 2},
+\end{equation}
+where $k$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$.
+Thus, not surprisingly Gaussian growth emerges for mutually independent events that have a linear time dependence.
+The same result was originally formulated with space as the independent variable instead of time and with error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details).
+In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.
+
+If we remove the dependency on time altogether, we observe the exponential growth equation
+\begin{equation}
+  \frac{d}{dt}\ln s(t) = - k,
+\end{equation}
+which yields 
+\begin{equation}
+  s(t) = A e^{ - k t}.
+\end{equation}
+This exponential equation implies a complete independence on both time and of the mutual entities involved in the growth.
+In the context of measurement error this function would be the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}.
+Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable but then physical interpretation would be a decay process.
 
@@
 
-We have shown how Gompertz growth emerges both at the microscopic and the macroscopic level as an infinitely correlated non-local effect.
+To summarize, we have seen that both time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and infinitely correlated growth yields the Gompertz function.
+
+We have further shown how Gompertz growth emerges both at the microscopic and the macroscopic level as a non-local effect.
 A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model derived from first principals.

### 2023-08-07 arxiv2/new_draft.tex
rephrase

@@
 He noticed that mortality rates increased at a geometric rate with age.
-In this context the ubiquitous stressor postulated above is of course time itself, clearly an external source.
+In this context the ubiquitous stressor postulated above is of course time itself.
 In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field.
+One could then see the entities as communicating indirectly though this field, in a perceived one-way fashion from the field to each entity.
 

### 2023-08-07 arxiv2/new_draft.tex
complete zhang and popp section

@@
 In the context of measurement error this function would be the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}.
-Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable but then physical interpretation would be a decay process.
+Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable but then the physical interpretation would be a decay process.
+
+As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities.
+By letting the growth rate be the rate at which a zero-count is observed, one recovers the Gompertz model where the Poisson mean parameter equals $\beta$ in Equation \ref{eq:Gomp}.
+This argument was put forth by \citet{shklovskii2005simple} although \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.
+One the other hand, Glauber also showed that incoherent systems follow the Gaussian distribution.
+This mapping between coherence and incoherence was postulated as a medical diagnostic technique by \citet{zhang1994log}.
 

### 2023-08-07 arxiv2/new_draft.tex
small change

@@
 In the context of measurement error this function would be the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}.
-Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable but then the physical interpretation would be a decay process.
+Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable where the physical interpretation would be a decay process.
 
 As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities.
-By letting the growth rate be the rate at which a zero-count is observed, one recovers the Gompertz model where the Poisson mean parameter equals $\beta$ in Equation \ref{eq:Gomp}.
+By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}.
 This argument was put forth by \citet{shklovskii2005simple} although \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.

### 2023-08-07 arxiv2/new_draft.tex
update capitals

@@
 On the other hand, if the logistic growth equation is recast into the reaction network equivalent, the catalyst is diminished by the growth of the reactant, which suggests that the source of growth is internal or at least dependent on the internal state.
-So even in reaction theory the gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions.
+So even in reaction theory the Gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions.
 
@@
 In this limit, the relative growth goes to infinity as the system approaches $t=0$, which could be seen as a verification of the maximally coherent characteristic of the system.
-This definition of coherence in the context of the gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.
+This definition of coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.
 

### 2023-08-09 arxiv2/new_draft.tex
slight fix ups

@@
 \begin{abstract}
-Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic abundance variable with an extra parameter, known as $\theta$-logistic growth or the Richards model for growth.
-In spite of this bridge, the biological foundation of Gompertzian growth is still little understood.
+Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth.
+In spite of this bridge, the biological foundation of Gompertzian growth is still only vaguely understood.
 We show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth, where the latter involves a network of entities.
 The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented.
-For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth.
+For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth.
 We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range or non-local communication.
@@
 
-This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exemplifying the Gompertz growth law as ubiquitous in nature.
+This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exposing Gompertz growth as ubiquitous in nature.
 The idea that the Gompertz model can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}.
@@
 
-Alongside the many contexts wherein Gompertz' law has been observed, an almost equal number of efforts have been made to interpret and derive this law from various phenomenological principals \cite{bajzer1997mathematical}.
+Alongside the many contexts wherein Gompertz' law has been observed, an almost equal number of efforts have been made to interpret and derive this law from various phenomenological principles \cite{bajzer1997mathematical}.
 This effort has often been accompanied by a phenomenological comparison of another popular growth model called logistic growth, or Verhulst growth, where ``the power to avoid destruction'' decreases linearly with time, as opposed to decreasing exponentially with time as seen with the Gompertz model.
@@
 Other environmental stressors like chronic radiation exposure have been observed to shift the Gompertz curve, an observation also theoretically developed by Sacher (1956) (\cite{sacher1956statistical}).
-Sacher modeled the toxic stressor as a stochastic perturbation to physiologic state of the population wherein beyond a certain point of perturbation in a given entity would result in its death.
+Sacher modeled the toxic stressor as a stochastic perturbation to the physiologic state of a population wherein beyond a certain point of perturbation in a given entity would result in its death.
 The interpretation of the Gompertz curve as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al.
@@
 
-Another framework used to show how the Gompertz can result from an external or ubiquitous stressor is chemical reaction theory.
+Another framework used to show how the Gompertz model can result from an external or ubiquitous stressor is chemical reaction theory.
 In reaction theory, a set of reactants combine to produce products, possibly in the presence of a catalyst.
 Morkov (2019) \cite{markov2019reaction} showed that the Gompertz can be interpreted as growth under a catalyst whose effect diminishes at a rate independent of the growth rate of the reactant.
-In other words, the catalyst can be considered as an external and ubiquitous stressor or catalyst to the reactants, perfectly in line with the ubiquitous stochastic stressor model from Sacher and De Lauro et al.
+In other words, the catalyst can be considered as an external and ubiquitous catalyst to the reactants, perfectly in line with the ubiquitous stochastic stressor model from Sacher and De Lauro et al.
 On the other hand, if the logistic growth equation is recast into the reaction network equivalent, the catalyst is diminished by the growth of the reactant, which suggests that the source of growth is internal or at least dependent on the internal state.
@@
 
-In this paper, we will elaborate on this idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
-We will use a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
+In this paper, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
+We will devise a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
 by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}).
@@
 If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model.
-As seen in the references, this is in line both with experimental and theoretical results.
-In presenting this argument, we will present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to source of growth.
+As seen in the aforementioned references, this is in line both with experimental and theoretical results.
+In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth.
 
-It is worth noting that the notion of Gompertz growth as maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}.
+It is worth noting that the notion that Gompertz growth is maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}.
 Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation.
@@
 
-In fact, it has been shown that cancer malignancies are mostly negative in polarity with respect to the rest of the body, while nonmalignant tumor pathology is mostly positive \cite{langman1949technique}.
+In fact, it has been shown that cancer malignancies are mostly negative in electromagnetic polarity with respect to the rest of the body, while nonmalignant tumor pathology is mostly positive \cite{langman1949technique}.
 One is then led to the interpretation that certain biological growth processes are coupled with an electromagnetic field of specific character.
-Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic or cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.
-The disruptive effect also extends to other biological processes, as shown by Wever and Brown's experiments on biological endogenous cycles lie the circadian cycle through manipulation of the environmental electromagnetic field \cite{brown1976biological,wever1974elf}.
+Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cycle or cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.
+This disruptive effect also extends to other biological processes, as shown by Wever and Brown's experiments on biological endogenous cycles lie the circadian cycle through manipulation of the environmental electromagnetic field \cite{brown1976biological,wever1974elf}.
 
@@
 \cite{wang2012richards} to not originate from a communicable disease paradigm where only local disease spread is possible.
-They used the common Susceptible-Infected-Recovered model by Kermack and McKendrick \cite{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter value range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
+They used the common Susceptible-Infected-Recovered model by Kermack and McKendrick \cite{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
 This would imply that any mortality patterns that shows a Gompertzian characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.
@@
 This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor.
-We proceed with a discussion section and end with a conclusion.
+We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and finally end with a conclusion.
 
@@
 \label{eq:Rich}
-  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)^{\theta}),
+  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X^{\theta}(t)),
 \end{equation}
@@
 (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of co-operation in the microscopic domain.
-In the limit where the system becomes \textit{maximaly cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
+They showed that in the limit where the system becomes \textit{maximaly cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
  \begin{equation}
@@
 In this limit, the relative growth goes to infinity as the system approaches $t=0$, which could be seen as a verification of the maximally coherent characteristic of the system.
-This definition of coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.
+One is reminded of a system that is perturbed and left to reach a new equilibrium.
+Petroni's definition of maximal coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.
 
@@
 The right hand side ``resource'' term is not specific enough so as to specify the location of the source.
-However, one could argue that the since the Gompertz model has asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal that initiates the growth and subsequently decays exponentially just like a memoryless process.
+However, one could argue that the since the Gompertz model has asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal (or catalyst in Markov's language) that initiates the growth and subsequently decays exponentially just like a memoryless process.
 To further explore the source of growth we turn to the microscopic domain.
@@
 Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables.
-Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ which a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
+Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ with a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
 \begin{equation}
 \label{eq:modSIR}
-\frac{d x_i}{dt} = x_i\sum_j{a_{ij}}s_{j} \quad \forall i.
+\frac{d x_i}{dt} = x_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
 \end{equation}
+where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$.
 First note that for a single pair this would amount to traditional logistic growth.
@@
 
-Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and give growth potential to a single possibly different entity}.
+Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}.
 Using this rule and rearranging and summing across all entities,
@@
 
-With our assumption of the variables being independent and identically distributed, we obtain the simplest form of logistic growth of the geometric mean of the ensemble driven by the sample mean of the inverse abundance, or the \textit{potential abundance},
+Dividing by the total number of entities and with our assumption of the variables being independent and identically distributed, we obtain the simplest form of logistic growth of the geometric mean of the ensemble driven by the sample mean of the inverse abundance, or the \textit{potential abundance},
 \begin{equation}
 \label{eq:modSIR3}
-\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} = \frac{1}{N}\sum_i^N s_i 
+\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} = \frac{1}{N}\sum_j^N s_j 
 \end{equation}
@@
 As N grows, both of geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles.
-Specifically, letting $\tilde{x}$ represent the sample geometric mean of $x_i$ and $\bar{x}$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have
+Specifically, letting $\tilde{x}(N)$ represent the sample geometric mean of $x_i$ and $\bar{x}(N)$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have
 \begin{eqnarray*}
   \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
-  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \tilde{x} &=& \lim_{N\rightarrow \infty}(1 - \bar{x}) \\
-  \frac{d}{dt} \ln \gamma &=&(1 - \mu)
+  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \tilde{x}(N) &=& \lim_{N\rightarrow \infty}(1 - \bar{x}(N)) \\
+  \frac{d}{dt} \ln \gamma &=&(1 - \mu),
   % \frac{d}{dt} \ln m &=&- \ln \bar{s},
 \end{eqnarray*}
+where the last equality is from a convergence in expectation.
 
@@
 Instead of limiting the growth equation to pairwise interactions, we allow higher order interactions in the form of triplets, quadruplets, and so on.
-Analogous to the pairwise exclusivity rule, each entity can again only participate in one interaction group at any given order and any given direction.
+Analogous to the pairwise exclusivity rule, each entity can again only participate in \textit{one} interaction group at any given order and any given direction.
 For example, any triplet can only interact with one other triplet and have exclusive possession of its entities at the triplet level.
-This means that we consider the geometric mean of each $n-$group as follows.
+Thus, let $S_k$ be the set of all mutually exclusive $k$-sized groups where $k=1,..., N/2$, where the highest order are capped at $N/2$ since that would be the largest number of entities in a single group that could interact with an identically sized group.
+This means that we consider the geometric mean of each $k$-group as follows,
 \begin{equation}
 \label{eq:modSIR4}
-\frac{d}{dt} \ln x_{i_1} x_{i_2} \dots x_{i_n} = s_{j_1}s_{j_2}\dots s_{j_n}, 
+\frac{d}{dt} \ln x_{i_1} x_{i_2} \dots x_{i_k} = s_{j_1}s_{j_2}\dots s_{j_k}, 
 \end{equation}
-where the indices $i$ and $j$ represent the higher order interaction between the mutually exclusive groups of entities for each $n$ from 1 and up to $N/2$.
-The higher orders are capped at $N/2$ since that would be the largest number of entities in a group that could interact with an identically sized group, viz.
+where the indices $i$ and $j$ represent the higher order interaction indices between the mutually exclusive groups of entities for each $k$ from 1 and up to $N/2$.
+Thus, the aggregated growth is represented by the sum of each interaction order,
 \begin{equation}
@@
 \label{eq:MicroGomp11}
-\lim_{N\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = \frac{1}{k} \mu^{k}.
+\lim_{N\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = \frac{1}{N}\frac{N}{k} \mu^{k} = \frac{1}{k} \mu^{k}.
 \end{equation}
@@
 But what if there is no connectivity between the entities but each entity is independent?
-In such a case we can formulate a growth equation where the relative growth is only linearly dependent on time, where we make a slight modification and model the \textit{potential abundance}, $s(t)$, instead of the actual abundance,
+In such a case we can formulate a growth equation where the relative growth is only linearly dependent on time, where we make a slight modification and model the \textit{potential abundance}, $s(t)$, instead of the realized abundance,
 \begin{equation}
-  \frac{d}{dt}s(t) = - k t s(t),
+  \frac{d}{dt}s(t) = - \beta t s(t),
 \end{equation}
@@
 \begin{equation}
-  \frac{d}{dt}\ln s(t) = - k t,
+  \frac{d}{dt}\ln s(t) = - \beta t,
 \end{equation}
-for some constant $k>0$.
+for some constant $\beta>0$.
 
@@
 \begin{equation}
-  s(t) = A e^{ - k t^2 / 2},
+  s(t) = A e^{ - \beta t^2 / 2},
 \end{equation}
-where $k$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$.
+where $\beta$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$.
 Thus, not surprisingly Gaussian growth emerges for mutually independent events that have a linear time dependence.
-The same result was originally formulated with space as the independent variable instead of time and with error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details).
+The same result was originally formulated by Gauss with space as the independent variable instead of time and with error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details).
 In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.
@@
 \begin{equation}
-  \frac{d}{dt}\ln s(t) = - k,
+  \frac{d}{dt}\ln s(t) = - \beta,
 \end{equation}
@@
 \begin{equation}
-  s(t) = A e^{ - k t}.
+  s(t) = A e^{ - \beta t}.
 \end{equation}
@@
 As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities.
-By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}.
-This argument was put forth by \citet{shklovskii2005simple} although \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.
-One the other hand, Glauber also showed that incoherent systems follow the Gaussian distribution.
+By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}.
+This argument was put forth by \citet{shklovskii2005simple} and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.
+One the other hand, Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution.
 This mapping between coherence and incoherence was postulated as a medical diagnostic technique by \citet{zhang1994log}.
@@
 We have further shown how Gompertz growth emerges both at the microscopic and the macroscopic level as a non-local effect.
-A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model derived from first principals.
+A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model derived from first principles.
 This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance.
 On the microscopic level, the Gompertz model emerges through an exclusivity rule at every order of interaction.
-As a corollary, we add another exhibit that show how the Logistic function is the result of local growth processes with pairwise-interactions only.
+As a corollary, this thesis shows how the logistic function is the result of local growth processes with pairwise-interactions only, in line with its observation in the fields of predator-prey models and in communicable diseases.
 
@@
 Examples of such models are already given by De Lauro (2014), Sacher (1956) and Markov (2019).
-\citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic in nature, and while the former do not discuss the external nature of such a source, their model does not preclude it.
-More generally, with growth as stochastic it is also ubiquitously present uniformly across all entities from the very onset of growth.
+\citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic in nature, and while Lauro et al.
+do not discuss the external nature of such a source, their model does not preclude it.
+More generally, with growth as stochastically rooted it is also ubiquitously present uniformly across all entities from the very onset of growth.
 Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Lauro and Sacher's models.
@@
 
-This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes.
-But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading.
+This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g.
+\citet{estrada2022networked}).
+But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head.
 In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.

### 2023-08-09 arxiv2/new_draft.tex
chage abstract a bit

@@
 Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth.
-In spite of this bridge, the biological foundation of Gompertzian growth is still only vaguely understood.
-We show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth, where the latter involves a network of entities.
+In spite of this bridge, the biological foundation of Gompertzian growth is only vaguely understood.
+To add to the our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities.
 The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented.

### 2023-08-13 arxiv2/new_draft.tex
add figure

@@
 % \def\bibcommenthead{}%
+
+\newcommand{\insertPngFig}[3]{
+  \begin{figure}[H]
+  \centering
+  \includegraphics[width=8.4cm]{#1.png}
+  \caption{#2}
+  \label{fig:#1}
+  \end{figure}
+}
 
@@
 In spite of this bridge, the biological foundation of Gompertzian growth is only vaguely understood.
-To add to the our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities.
+To add to our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities.
 The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented.
@@
 Analogous to the pairwise exclusivity rule, each entity can again only participate in \textit{one} interaction group at any given order and any given direction.
-For example, any triplet can only interact with one other triplet and have exclusive possession of its entities at the triplet level.
+For example, any triplet can only interact with one other triplet and have exclusive possession of its entities at the triplet level (Figure \ref{fig:order}).
+
+\insertPngFig{order}{Exclusive interaction groups at each order: A) First order interaction, each entity is paired with a single other entity.
+B) Second order interaction, mutually exclusive groups of 2 entities interact with each other.
+B) Third order interaction, same as B) but for groups of three entities.}
+
 Thus, let $S_k$ be the set of all mutually exclusive $k$-sized groups where $k=1,..., N/2$, where the highest order are capped at $N/2$ since that would be the largest number of entities in a single group that could interact with an identically sized group.

### 2023-09-04 arxiv2/new_draft.tex
2 -> two
update abstract

@@
 In situations of Gompertzian decay this field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, while in situations of Gompertzian growth the field could be an internal or external electromagnetic field.
+Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and infinitely correlated growth yields the Gompertz function.
+
 \end{abstract}
@@
 \insertPngFig{order}{Exclusive interaction groups at each order: A) First order interaction, each entity is paired with a single other entity.
-B) Second order interaction, mutually exclusive groups of 2 entities interact with each other.
+B) Second order interaction, mutually exclusive groups of two entities interact with each other.
 B) Third order interaction, same as B) but for groups of three entities.}

### 2023-09-05 arxiv2/new_draft.tex
small edits

--- 2023-08-13 arxiv2/new_draft.tex
@@
 In situations of Gompertzian decay this field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, while in situations of Gompertzian growth the field could be an internal or external electromagnetic field.
+Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
+
 \end{abstract}
@@
 where fractional prefactors to each term will be justified in the sequel.
-After some algebra, this augmented logistic equation reduces to a closed form expression involving Gauss' Hypergeometric function, ${}_1F_{2}$, viz.
+After some algebra and letting $\beta=1$ for simplicity, this augmented logistic equation reduces to a closed form expression involving Gauss' Hypergeometric function, ${}_1F_{2}$, viz.
 \begin{equation}
@@
 \end{eqnarray*}
-where the last equality is from a convergence in expectation.
+where the last equality follows from a convergence in expectation.
 
@@
 \begin{enumerate}
-\item As the number of entities, N, grows each term will converge in expectation to higher powers of the mean of $s_i$, multiplied by the prefactor $1$, $1/2$, $1/3$, etc., 
+\item As the number of entities grow each term will converge in expectation to higher powers of the mean of $s_i$, multiplied by the prefactor $1$, $1/2$, $1/3$, etc., 
 \begin{equation}
@@
 \end{equation}
-\item As the higher order interaction terms, k, grow
+\item As the higher order interaction terms grow
 \begin{equation}
@@
 \label{eq:MicroGomp13}
-\lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\\
+\lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\left[\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\right]\\
 \frac{d}{dt} \ln \gamma &=&  \mu + \frac{1}{2} \mu^2 + ...\\
@@
 We finally note that for the right hand side to exhibit asymptotically infinite growth at the onset of growth, it is not only sufficient but also necessary to have higher order terms.
-This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions only under the assumptions presented in this model.
+This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.
 
@@
 
+In general, our choice of time as the dependent variable can be substituted for any suitable metric of choice.
+Another example of space as the dependent variable in the Gompertz setting is given by \citet{molski2003coherent}.
+
 As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities.
 By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}.
-This argument was put forth by \citet{shklovskii2005simple} and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.
-One the other hand, Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution.
-This mapping between coherence and incoherence was postulated as a medical diagnostic technique by \citet{zhang1994log}.
+This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.
+One the other hand, Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here.
+The mapping between coherence and incoherence has also been postulated as a medical diagnostic technique by \citet{zhang1994log}, where incoherence is a sign of disease while coherence is a sign of well-being.
 
@@
 
-To summarize, we have seen that both time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and infinitely correlated growth yields the Gompertz function.
+To summarize, we have seen that both time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
 
 We have further shown how Gompertz growth emerges both at the microscopic and the macroscopic level as a non-local effect.
-A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model derived from first principles.
+A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model and is derived from first principles.
 This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance.
@@
 He noticed that mortality rates increased at a geometric rate with age.
-In this context the ubiquitous stressor postulated above is of course time itself.
+In this context the ubiquitous stressor postulated above is of course time itself and thus Gompertz growth in this context will be implicitly \textit{time dependent}.
 In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field.

### 2023-09-18 arxiv2/new_draft.tex
minor fixes

@@
 % \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
-\title{Gompertzian growth as infinitely correlated non-local growth}
+\title{Gompertz growth as infinitely correlated non-local growth}
 
@@
 \begin{abstract}
-Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth.
-In spite of this bridge, the biological foundation of Gompertzian growth is only vaguely understood.
+Logistic and Gompertz growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth.
+In spite of this bridge, the biological foundation of Gompertz growth is only vaguely understood.
 To add to our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities.
 The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented.
-For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth.
-We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range or non-local communication.
-This conclusion adds to the interpretation of biological systems that undergo Gompertzian growth where long-range communication could be interpreted as the source of growth being external to the individual particles, like a field that stimulates growth simultaneously across all entities.
-In situations of Gompertzian decay this field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, while in situations of Gompertzian growth the field could be an internal or external electromagnetic field.
+For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertz growth.
+We thus equate Gompertz growth with correlated microscopic entities that exhibit long-range or non-local interaction.
+This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities.
+This field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, or an environmental stimulant like an electromagnetic field.
 Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
@@
 Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time \cite{makeham1889further}.
+Gompertz' law has later been known as the Gompertz model or Gompertz growth or decay.
 
-This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exposing Gompertz growth as ubiquitous in nature.
-The idea that the Gompertz model can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}.
-Gompertz law has also been shown to apply to societal phenomena like growth in railway traffic, and the demand for goods and services, and sales of tobacco \cite{olshansky1997ever,prescott1922law,peabody1924growth}.
+\subsection{Gompertz' law in existing work}
+This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exposing Gompertz growth as nearly ubiquitous in biological growth and decay processes.
+The idea that the Gompertz' law can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}.
+Gompertz' law has also been shown to apply to societal phenomena like growth in railway traffic, and the demand for goods and services, and sales of tobacco \cite{olshansky1997ever,prescott1922law,peabody1924growth}.
 
@@
 In a study comparing the Gompertz and the logistic growth models, Petroni et al.
-(2020) \cite{petroni2020logistic} show that Gompertz growth can be interpreted as maximally coherent growth based on long-range communication or correlation, while logistic growth is minimally coherent without long-range behavior.
-With this interpretation, one is lead to postulate that the Gompertz function could be a result of a global or external stimuli while the logistic function is a result of internal or local interaction.
-As such, the logistic function has been popular both for communicable disease models \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately} and and for predator-prey theory \cite{berryman1992orgins}, where disease or injury spreads through pairwise interactions.
+(2020) \cite{petroni2020logistic} show that Gompertz growth can be interpreted as \textit{maximally coherent} growth based on long-range interaction or correlation, while logistic growth is minimally coherent without long-range behavior.
+With this interpretation, one is lead to postulate that the Gompertz model is a result of an external stimulus with perceived long-range interactions while the logistic model is a result of an internal stimulus with local interaction.
+As such, the logistic model has been popular both for communicable disease models \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately} and and for predator-prey theory \cite{berryman1992orgins}, where disease or injury spreads through pairwise interactions.
 
-A number of studies in the biological context have argued that the Gompertzian model could be sourced or activated from an external factor which is ubiquitously and instantaneously introduced to the system, e.g.
+A number of studies in the biological context have argued that Gompertz growth could be activated from an external factor which is ubiquitously and instantaneously introduced to the system, e.g.
 as a toxic agent or an environmental stressor to which the system gradually adapts \cite{neafsey1988gompertz,neafsey1989gompertz,thompson1990biphasic}.
-In the original case of mortality rates as a function of age, this interpretation would imply that time itself is the environmental ubiquitous stressor.
+In the original case of mortality rates as a function of age, this interpretation would imply that simply living on earth is the ubiquitous stressor.
 Other environmental stressors like chronic radiation exposure have been observed to shift the Gompertz curve, an observation also theoretically developed by Sacher (1956) (\cite{sacher1956statistical}).
-Sacher modeled the toxic stressor as a stochastic perturbation to the physiologic state of a population wherein beyond a certain point of perturbation in a given entity would result in its death.
-The interpretation of the Gompertz curve as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al.
+Sacher modeled the toxic stressor as a stochastic perturbation to the physiologic state of a population wherein death of an entity would occur beyond some perturbation threshold.
+The interpretation of Gompertz growth as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al.
 (2014) \cite{de2014stochastic} who arrived at the same result without alluding to a threshold value beyond which death was inevitable, but rather by letting the stochastic perturbation be the source of growth or decay.
@@
 In reaction theory, a set of reactants combine to produce products, possibly in the presence of a catalyst.
-Morkov (2019) \cite{markov2019reaction} showed that the Gompertz can be interpreted as growth under a catalyst whose effect diminishes at a rate independent of the growth rate of the reactant.
-In other words, the catalyst can be considered as an external and ubiquitous catalyst to the reactants, perfectly in line with the ubiquitous stochastic stressor model from Sacher and De Lauro et al.
+Morkov (2019) \cite{markov2019reaction} showed that the Gompertz model can be interpreted as growth under a catalyst whose effect diminishes at a rate independent of the growth rate of the reactant.
+In other words, the catalyst can be considered as an external and ubiquitous stimulant or stressor to the reactants, perfectly in line with the ubiquitous stochastic perturbation model from Sacher and De Lauro et al.
 On the other hand, if the logistic growth equation is recast into the reaction network equivalent, the catalyst is diminished by the growth of the reactant, which suggests that the source of growth is internal or at least dependent on the internal state.
-So even in reaction theory the Gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions.
-
-In this paper, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
-We will devise a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
-by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}).
-In particular, instead of considering a real number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model.
-As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $n$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities.
-If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model.
-As seen in the aforementioned references, this is in line both with experimental and theoretical results.
-In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth.
+So even in reaction theory the Gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions of the reactants.
 
@@
 For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.
-The limb-regeneration process of the salamander is an example of such a phenomena.
 
-In fact, it has been shown that cancer malignancies are mostly negative in electromagnetic polarity with respect to the rest of the body, while nonmalignant tumor pathology is mostly positive \cite{langman1949technique}.
-One is then led to the interpretation that certain biological growth processes are coupled with an electromagnetic field of specific character.
-Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cycle or cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.
-This disruptive effect also extends to other biological processes, as shown by Wever and Brown's experiments on biological endogenous cycles lie the circadian cycle through manipulation of the environmental electromagnetic field \cite{brown1976biological,wever1974elf}.
+In fact, it has been shown that cancer malignancies are mostly \textit{negative} in electric polarity with respect to the rest of the body, while nonmalignant (or benign) tumor pathology is mostly \textit{positive} in polarity \cite{langman1949technique}.
+Gompertz growth has also been widely observed in cancerous cells \cite{kozusko2003combining,laird1964dynamics}.
+One is then led to the interpretation that certain biological growth processes that exhibit Gompertz growth are coupled with an electromagnetic field of specific character.
+Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.
 
 Finally, on the topic of disease modelling, the Gompertz model has been shown by Wang et al.
-\cite{wang2012richards} to not originate from a communicable disease paradigm where only local disease spread is possible.
-They used the common Susceptible-Infected-Recovered model by Kermack and McKendrick \cite{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
-This would imply that any mortality patterns that shows a Gompertzian characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.
+\cite{wang2012richards} to be excluded from a communicable disease paradigm where only local disease spread is possible.
+They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
+Given this observation by Wang et al., any mortality pattern that shows a Gompertz characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.
+While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.
+
+\subsection{Main Results and Roadmap}
+In this paper, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
+We will devise a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
+by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}).
+In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model.
+As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $n$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities.
+If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model.
+As seen in the aforementioned references, this is in line both with experimental and theoretical results.
+In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth.
 
@@
 This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor.
-We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and finally end with a conclusion.
+We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and then conclude.
 
@@
 Petroni et al.
-(2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of co-operation in the microscopic domain.
+(2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of cooperation in the microscopic domain.
 They showed that in the limit where the system becomes \textit{maximaly cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
@@
 
-Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables.
+Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables, where we define interaction in this context as the transfer of one entity's potential growth to another entity which converts potential to realized growth.
 Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ with a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
@@
 \subsection{Microscopic Gompertz Growth}
+\label{sec:microGomp}
 
@@
 
-We finally note that for the right hand side to exhibit asymptotically infinite growth at the onset of growth, it is not only sufficient but also necessary to have higher order terms.
+We finally note that for the right hand side to exhibit asymptotically infinite growth rate at the onset of growth, it is not only sufficient but also necessary to have higher order terms.
 This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.
@@
 This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.
-One the other hand, Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here.
+Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here.
 The mapping between coherence and incoherence has also been postulated as a medical diagnostic technique by \citet{zhang1994log}, where incoherence is a sign of disease while coherence is a sign of well-being.
 
-\section{Conclusion}
+\section{Discussion}
 \label{sec:discussion}
@@
 
-An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}.
+An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be a ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}.
 With a such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli.
@@
 He noticed that mortality rates increased at a geometric rate with age.
-In this context the ubiquitous stressor postulated above is of course time itself and thus Gompertz growth in this context will be implicitly \textit{time dependent}.
-In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field.
-One could then see the entities as communicating indirectly though this field, in a perceived one-way fashion from the field to each entity.
+In this context the ubiquitous stressor postulated above would be the living conditions on our planet, for which time acts as a proxy.
+One could further postulate this stressor as a field surrounding the earth, e.g.
+in its electromagnetic footprint often related with the Schumann Resonance \cite{cherry2002schumann,schumann1952strahlungslosen}.
+It is indeed curious that we do not see mortality rates behave as linearly dependent on time, for which a Gaussian curve would be appropriate.
+Nor do we see any finite-ordered polynomial time dependence on mortality rates.
+Instead we see growth of mortality rates as sourced from other entities at all interaction orders, which in light of our discussion above is equivalent to mortality coming from an environmental stressor.
 
@@
 
-\section*{Supplement}
+\section*{Appendix A}
+Here we arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.
+Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with the logistic growth equation augmented by $K$ terms and using the nomenclature defined in Section \ref{sec:micro}, viz.
+
+\begin{equation}
+\frac{d \ln{x}}{dt} = \sum_{i=1}^K s^i/i.
+\end{equation}
+Using standard integrals we see that this sum amounts to a Gompertz growth term adjusted by a Hypergeometric function,
+\begin{eqnarray*}
+  \sum_{i=1}^K s^i/i &=& \int_s \sum_{i=0}^{K-1} s^i ds\\
+  &=&\int_s \frac{1-s^K}{1-s} ds\\
+  &=&- \frac{{}_{1}F_{2}({1,K+1,K+2;s})}{K+1}s^{K+1} - \ln(1-s),
+\end{eqnarray*}
+where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
+ \begin{equation}
+ \label{eq:modLogistic2}
+\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,K+1,K+2;1-x(t)})}{K+1}(1-x(t))^{K+1} - \ln{x(t)},
+ \end{equation}
+ which represents the novel bridge between logistic and Gompertzian growth.
+Note that for $K=1$, this relative growth rate becomes the standard logistic function, while increasing the number of higher order terms will move the equation closer to the Gompertz curve as the Hypergeometric term approaches zero.
+This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig.
+\ref{fig:hypergeometric_plain}).
 

### 2023-09-20 arxiv2/new_draft.tex
add levitt and zondas paper

@@
 
-Finally, on the topic of disease modelling, the Gompertz model has been shown by Wang et al.
-\cite{wang2012richards} to be excluded from a communicable disease paradigm where only local disease spread is possible.
+On the topic of disease modelling, the Gompertz model has been shown by Wang et al.
+\cite{wang2012richards} to be excluded from a subset of models in the communicable disease paradigm.
 They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
-Given this observation by Wang et al., any mortality pattern that shows a Gompertz characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.
+If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}.
+Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor.
 While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.
@@
 
-For growth within biological organisms, it has been well established that the mediating field is the electromagnetic field as discussed by R.O.
-Becker (see \citet{becker1984electromagnetic} and references within), which suggests that the accompanying growth or decay pattern in these processes follows the Gompertz model as the result of a ubiquitous activating field.
+For many growth processes within biological organisms, it has been well established that the mediator is the electromagnetic field \cite{becker1984electromagnetic}, which suggests that the accompanying growth or decay pattern in these processes follows the Gompertz model as the result of a ubiquitous activating field.
 For growth or decay in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts.
@@
 
-It is worth noting that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age.
+We finally note that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age.
 He noticed that mortality rates increased at a geometric rate with age.

### 2023-09-29 arxiv2/new_draft.tex
move technical part of discussion to results

@@
 This field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, or an environmental stimulant like an electromagnetic field. 
-Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
+Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
 
@@
 Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time \cite{makeham1889further}.
-Gompertz' law has later been known as the Gompertz model or Gompertz growth or decay.
+Gompertz' law has later been known as the Gompertz model or Gompertz growth.
 
 \subsection{Gompertz' law in existing work}
-This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exposing Gompertz growth as nearly ubiquitous in biological growth and decay processes.
+This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exposing Gompertz growth as nearly ubiquitous in biological growth processes.
 The idea that the Gompertz' law can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}.
@@
 as a toxic agent or an environmental stressor to which the system gradually adapts \cite{neafsey1988gompertz,neafsey1989gompertz,thompson1990biphasic}.
-In the original case of mortality rates as a function of age, this interpretation would imply that simply living on earth is the ubiquitous stressor.
+In the original case of mortality rates as a function of age, this interpretation would imply that time itself is the environmental ubiquitous stressor.
 Other environmental stressors like chronic radiation exposure have been observed to shift the Gompertz curve, an observation also theoretically developed by Sacher (1956) (\cite{sacher1956statistical}).
@@
 The interpretation of Gompertz growth as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al.
-(2014) \cite{de2014stochastic} who arrived at the same result without alluding to a threshold value beyond which death was inevitable, but rather by letting the stochastic perturbation be the source of growth or decay.
+(2014) \cite{de2014stochastic} who arrived at the same result without alluding to a threshold value beyond which death was inevitable, but rather by letting the stochastic perturbation be the source of growth.
 
@@
 If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}.
-Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor.
+We will show that one can also obtain a Gompertz curve from a ubiquitous stressor scenario based on microscopic interactions.
+% Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor. 
 While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.
@@
 \subsection{Main Results and Roadmap}
-In this paper, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
-We will devise a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
+In this paper, our primary aim is to investigate the source of growth in self-organized processes.
+Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
+We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
 by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}).
@@
 In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth.
+Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz and Poisson.
 
-Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which will offer an alternative to the traditional Richard's model.
+Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which will offer an alternative to the traditional Richards model.
 Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena.
@@
 \end{equation}
-This is commonly interpreted as the right hand side representing ``resources available for growth'' which decreases linearly with respect to the relative abundance's growth rate on the left hand side.
+This is commonly interpreted as the right hand side representing ``resources available for growth'' which decrease linearly with respect to the relative abundance's growth rate on the left hand side.
 One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
@@
 \end{equation}
-also called $\theta$-logistic growth or Richards' model.
+also called $\theta$-logistic growth or Richards model \citep{richards1959flexible}.
 Petroni et al.
 (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of cooperation in the microscopic domain.
-They showed that in the limit where the system becomes \textit{maximaly cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
+They showed that in the limit where the system becomes \textit{maximally cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
  \begin{equation}
@@
 \end{equation}
-In this limit, the relative growth goes to infinity as the system approaches $t=0$, which could be seen as a verification of the maximally coherent characteristic of the system.
+In this limit, the relative growth goes to infinity as $\lim_{t\to 0}$, which could be seen as a verification of the maximally coherent characteristic of the system.
 One is reminded of a system that is perturbed and left to reach a new equilibrium.
@@
 In practice, the augmented logistic model has a faster growth rate decay than the familiar $\theta$-logistic growth (Figure \ref{fig:hypergeometric_plain}).
-If we interpret speed of relative growth decay as coherence, the augmented logistic model would be more coherent than the $\theta$-logistic model for a given starting value.
+If we interpret deceleration of relative growth as degree of coherence, the augmented logistic growth would be more coherent than the $\theta$-logistic model for a given starting value.
 
@@
 \ref{eq:Rich}), Gompertz model (Eq.
-\ref{eq:Gomp}), and the augmented causality-reversed SIR model (Eq.
+\ref{eq:Gomp}), and the augmented logistic model (Eq.
 \ref{eq:modLogistic}), all as a function of the abundance variable.
 Both dependent and independent variables are normalized so that final abundance (or size) is set to unity.
-The Richards $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the modified SIR model.}
+The Richards $\theta$-parameter is annotated in brackets in the legend, in addition to the number of terms used in the augmented logistic model.}
 
-In spite of the connection of the Gompertz model with higher order terms in a logistic model, it is still not clear whether the source of growth is external or internal.
+In spite of the connection of the Gompertz model with higher order terms in an augmented logistic model, it is not clear whether the source of growth is external or internal to the system.
 The right hand side ``resource'' term is not specific enough so as to specify the location of the source.
-However, one could argue that the since the Gompertz model has asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal (or catalyst in Markov's language) that initiates the growth and subsequently decays exponentially just like a memoryless process.
+One could argue that the since the Gompertz model has asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal (or catalyst in Markov's language) that initiates the growth and subsequently decays exponentially like a memoryless process.
 To further explore the source of growth we turn to the microscopic domain.
@@
 
-As N grows, both of geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles.
+As N grows, both of the geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles.
 Specifically, letting $\tilde{x}(N)$ represent the sample geometric mean of $x_i$ and $\bar{x}(N)$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have
@@
 \label{eq:MicroGomp0}
-\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} s_{j_1}s_{j_2}\dots s_{j_{N/2}},
+\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_{j\in S_1}^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} s_{j_1}s_{j_2}\dots s_{j_{N/2}},
 \end{equation}
@@
 Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic and the geometric means of the microscopic entities
-\begin{eqnarray*}
+\begin{eqnarray}
 \label{eq:MicroGomp13}
@@
 \frac{d}{dt} \ln \gamma &=& - \ln (1-\mu).
-\end{eqnarray*}
+\end{eqnarray}
 
-We finally note that for the right hand side to exhibit asymptotically infinite growth rate at the onset of growth, it is not only sufficient but also necessary to have higher order terms.
+Note that for the right hand side to exhibit asymptotically infinite growth rate at the onset of growth, it is not only sufficient but also necessary to have higher order terms.
 This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.
+
+Since all higher order interaction would necessarily imply non-local interaction, the natural interpretation would be that the growth is activated by a ubiquitous field accessible by all entities simultaneously.
+In this way one could think of the activating field as an effectively external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism.
+
+Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case.
+To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same result can be obtained under the exclusivity condition. 
+Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model.
+
+This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g.
+\citet{estrada2022networked}).
+But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head.
+In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
 
@@
 In the context of measurement error this function would be the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}.
-Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable where the physical interpretation would be a decay process.
+% Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable where the physical interpretation would be a decay process.
 
@@
 Another example of space as the dependent variable in the Gompertz setting is given by \citet{molski2003coherent}.
+
+\subsection{Unification with the Poisson distribution}
 
@@
 
-\section{Discussion}
+\section{Discussion and Conclusion}
 \label{sec:discussion}
 
-To summarize, we have seen that both time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
+To summarize, we have seen that both time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
 
 We have further shown how Gompertz growth emerges both at the microscopic and the macroscopic level as a non-local effect.
-A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model and is derived from first principles.
+A novel bridge between logistic and Gompertz growth is established involving Gauss' Hypergeometric function that serves as a replacement for the Richards model and is derived from first principles.
 This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance.
 On the microscopic level, the Gompertz model emerges through an exclusivity rule at every order of interaction.
-As a corollary, this thesis shows how the logistic function is the result of local growth processes with pairwise-interactions only, in line with its observation in the fields of predator-prey models and in communicable diseases.
+As a corollary, this thesis shows how the logistic function is the result of local growth processes with pairwise-interactions only, in line with its observation in the fields of predator-prey models and in communicable diseases \cite{may2007theoretical}.
 
@@
 
-For many growth processes within biological organisms, it has been well established that the mediator is the electromagnetic field \cite{becker1984electromagnetic}, which suggests that the accompanying growth or decay pattern in these processes follows the Gompertz model as the result of a ubiquitous activating field.
-For growth or decay in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts.
-
-Another way to see the source growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case.
-To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same result can be obtained under the exclusivity condition.
-Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model.
-
-This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g.
-\citet{estrada2022networked}).
-But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head.
-In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
+For many growth processes within biological organisms, it has been well established that the mediator is the electromagnetic field \cite{becker1984electromagnetic}, which suggests that the accompanying growth pattern in these processes follows the Gompertz model as the result of a ubiquitous growth-inducing field.
+For growth in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts.
 

### 2023-10-02 arxiv2/new_draft.tex
final thouches

@@
 The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. 
-For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertz growth.
+For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth rate decreases linearly, the trademark of Gompertz growth.
 We thus equate Gompertz growth with correlated microscopic entities that exhibit long-range or non-local interaction.
@@
 In a study comparing the Gompertz and the logistic growth models, Petroni et al.
-(2020) \cite{petroni2020logistic} show that Gompertz growth can be interpreted as \textit{maximally coherent} growth based on long-range interaction or correlation, while logistic growth is minimally coherent without long-range behavior.
+(2020) \cite{petroni2020logistic} show that Gompertz growth can be interpreted as \textit{maximally coherent} growth based on long-range interaction or correlation, while logistic growth is minimally coherent with non-collaborating elementary entities.
 With this interpretation, one is lead to postulate that the Gompertz model is a result of an external stimulus with perceived long-range interactions while the logistic model is a result of an internal stimulus with local interaction.
@@
 On the other hand, if the logistic growth equation is recast into the reaction network equivalent, the catalyst is diminished by the growth of the reactant, which suggests that the source of growth is internal or at least dependent on the internal state.
-So even in reaction theory the Gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions of the reactants.
+So also in reaction theory the Gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions of the reactants.
 
@@
 Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize).
-Perhaps some of the biological macroscopic growth phenomena references are also due to such fields.
+Perhaps some biological macroscopic growth phenomena are also due to such fields.
 For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.
@@
 \begin{equation}
-  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)).
+  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)),
 \end{equation}
-This is commonly interpreted as the right hand side representing ``resources available for growth'' which decrease linearly with respect to the relative abundance's growth rate on the left hand side.
+where $\beta>0$ is a growth parameter.
+Here, the right hand side is commonly interpreted as representing ``resources available for growth'' which decrease linearly with respect to the relative abundance's growth rate on the left hand side.
 One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
@@
 where fractional prefactors to each term will be justified in the sequel.
-After some algebra and letting $\beta=1$ for simplicity, this augmented logistic equation reduces to a closed form expression involving Gauss' Hypergeometric function, ${}_1F_{2}$, viz.
+After some algebra (Appendix \ref{appendix:a}) and letting $\beta=1$ for simplicity, this augmented logistic equation reduces to a closed form expression involving Gauss' Hypergeometric function, ${}_1F_{2}$, viz.
 \begin{equation}
@@
 Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables, where we define interaction in this context as the transfer of one entity's potential growth to another entity which converts potential to realized growth.
-Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ with a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
+Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ as a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
 \begin{equation}
@@
 Since all higher order interaction would necessarily imply non-local interaction, the natural interpretation would be that the growth is activated by a ubiquitous field accessible by all entities simultaneously.
-In this way one could think of the activating field as an effectively external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism.
+In this way one could think of the activating field as an \textit{effectively} external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism.
 
@@
 
-\section*{Appendix A}
+\section{Appendix A}
+\label{appendix:a}
 Here we arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model.

### 2023-10-09 arxiv2/new_draft.tex
update

@@
 \usepackage{lineno}
-% \linenumbers
+\linenumbers
 
@@
 
-\author{Matz A. Haugen, \\
+\author{Matz A. Haugen\footnote{hi}, \\
 Independent Reseacher, Oslo\\
@@
 In spite of this bridge, the biological foundation of Gompertz growth is only vaguely understood.
-To add to our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities.
+To add to this biological foundation, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities.
 The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. 
@@
 % keywords can be removed
-\keywords{Gompertz \and Coherence \and Network Analysis \and Stochastic}
+\keywords{Gompertz \and Coherence \and Network Analysis \and Self-organization}
+
+\section*{Highlights}
+\begin{itemize}
+\item Logistic and Gompertz growth is connected with a novel framework based on microscopic principles.
+\item Gompertz growth is connected with ubiquitous growth-activating field justified by physical and mathematical principles.
+\item Growth processes is put in a larger contextual framework based on varying dependence in spacetime, including the Exponential, Gaussian and Poisson distributions. 
+\end{itemize}
 
@@
  \end{equation}
- which represents the novel bridge between logistic and Gompertzian growth.
+ which represents the novel bridge between logistic and Gompertz growth.
 Note that for $K=1$, this relative growth rate becomes the standard logistic function, while increasing the number of higher order terms will move the equation closer to the Gompertz curve as the Hypergeometric term approaches zero.

### 2024-03-08 arxiv2/new_draft.tex
finalize second draft

@@
 Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize).
-Perhaps some biological macroscopic growth phenomena are also due to such fields.
+Some biological macroscopic growth phenomena are also due to such fields.
 For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.
@@
 They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
-If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}.
-We will show that one can also obtain a Gompertz curve from a ubiquitous stressor scenario based on microscopic interactions.
+If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near-Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}.
+We will show that one can obtain a true Gompertz curve from a ubiquitous stressor scenario based on microscopic interactions.
 % Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor. 
@@
 In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model.
-As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $n$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities.
+As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $N$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities.
 If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model.
@@
 
-Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which will offer an alternative to the traditional Richards model.
+Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which offers an alternative to the traditional Richards model.
 Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena.
@@
 
+Consider $N$ entities represented by a set of random abundance variables, $x_i(t), \quad i=1, ..., N$, which are unitless and take values between 0 and 1, where 0 represents no realized growth while 1 represents the maximum potential growth having been realized.
+In other words, these variables represent the \textit{realized potential}, either in the form of individual volume, mass, likelihood of infection or mortality by being preyed on, etc.
+Let these variables have a shared expected value as a function of time, $\mu(t)$, and corresponding higher moments, i.e. each variable follows the same statistical distribution.
+With these variables it will also be convenient to also define an \textit{abundance potential} variable, $s_i = 1 - x_i$, where we omit the temporal variable for ease of exposition.
+An abundance potential could represent the future capacity to grow, e.g.
+encoded in an organism's DNA combined with its environmental factors.
+Our goal now is to investigate under which conditions we can achieve logistic or Gompertz growth.
+
+Preliminarily, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables, where we define interaction in this context as the transfer of one entity's potential growth to another entity which converts potential to realized growth.
+Examples of such networks could be cells that communicate either through electron or protein transfer, predator-prey systems or even human interactions through mutual trade, spreading of knowledge, or disease.
+
 \subsection{Microscopic Logistic Growth}
-Consider $N$ entities each represented by a set of independent and identically distributed random abundance variables, $x_i(t), \quad i=1, ..., N$, which are unitless and take values between 0 and 1, where 0 represents no realized growth while 1 represents the maximum potential growth having been realized and $\mu$ representing the mean of the distribution.
-With these variables it will also be convenient to also define an \textit{abundance potential} variable, $s_i = 1 - x_i$, where we omit the temporal variable for ease of exposition.
 
-Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables, where we define interaction in this context as the transfer of one entity's potential growth to another entity which converts potential to realized growth.
-Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ as a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
+Now consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ as a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
 \begin{equation}
@@
 \end{equation}
+or
+\begin{equation}
+\label{eq:modSIR2}
+\frac{d}{dt}\ln x_i = \sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
+\end{equation}
 where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$.
-First note that for a single pair this would amount to traditional logistic growth.
+For example, if the adjacency matrix has only $1$s and no $0$s, this means that every node is connected to every other node.
+The diagonal of the adjacency matrix is zero as a node is not connected to itself by definition.
+
+Let us look at the simplest case of two connected entities only, viz.
+\begin{eqnarray}
+\frac{d x_1}{dt} = x_1 s_2, \\
+\frac{d x_2}{dt} = x_2 s_1.
+\end{eqnarray}
+Rearranging and summing the two equations, we can express the variables in terms of their averages,
+\begin{equation}
+\frac{1}{2} \left\{\frac{1}{x_1}\frac{d x_1}{dt} + \frac{1}{x_2}\frac{d x_2}{dt}\right\} = \frac{1}{2}(s_1 + s_2) = 1 - \frac{1}{2}(x_1 + x_2),
+\end{equation}
+where the left hand side represents the average of the relative growth, while the right hand side represents the average abundance potential.
+Thus, even for a simple two-variable system we see that the system can be described in terms of summary statistics, which we shall see that we will be able to generalize.
+
 We choose to work in a unitless domain without parameters for the interaction strength and carrying capacity for simplicity of exposition.
@@
 Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}.
-Using this rule and rearranging and summing across all entities,
+This condition translates into an adjacency matrix with each column containing a single non-zero entry.
+Using this rule and rearranging and summing across all entities we obtain the following relationship,
 \begin{equation}
@@
 
-Dividing by the total number of entities and with our assumption of the variables being independent and identically distributed, we obtain the simplest form of logistic growth of the geometric mean of the ensemble driven by the sample mean of the inverse abundance, or the \textit{potential abundance},
+Dividing by the total number of entities, we obtain the simplest form of logistic growth of the geometric mean of the ensemble driven by the arithmetic mean of the inverse abundance, or the \textit{potential abundance},
 \begin{equation}
 \label{eq:modSIR3}
-\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} = \frac{1}{N}\sum_j^N s_j 
+\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} = \frac{1}{N}\sum_j^N s_j.
 \end{equation}
 
-As N grows, both of the geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles.
-Specifically, letting $\tilde{x}(N)$ represent the sample geometric mean of $x_i$ and $\bar{x}(N)$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have
+Now note that the time derivative of the geometric mean translates into the \textit{average relative growth rate}, where if we let $v_i=\frac{1}{x_i}\frac{d}{dt}x_i$, we see that
+
+\begin{equation}
+\label{eq:modSIR4}
+\frac{1}{N}\sum_j^N v_i = \frac{1}{N}\sum_j^N s_j.
+\end{equation}
+
+At this point we have shown how macroscopic variables can be obtained from their microscopic counterparts.
+We have not needed to evoke the randomness of our variables, but have obtained a macroscopic understanding only using sample quantities starting from pairwise microscopic principles.
+
+However, if we assume our set of random variables, $\{x_i\}$, converge to a distributional mean as $N$ grows, both of the relative growth rate average and the abundance average will converge to their distributional counterparts.
+Specifically and with a slight abuse of notation in the arguments, letting $\bar{x}(N)$ represent the sample abundance mean for N nodes which converges to its distributional counterpart $\mu(t)$, it follows from the continuous mapping theorem that $\bar{v}(N)$ will converge to $\frac{1}{\mu(t)}\frac{1}{dt}\mu(t)$ so long as the time derivative is smooth.
+We thus have
 \begin{eqnarray*}
-  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
-  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \tilde{x}(N) &=& \lim_{N\rightarrow \infty}(1 - \bar{x}(N)) \\
-  \frac{d}{dt} \ln \gamma &=&(1 - \mu),
-  % \frac{d}{dt} \ln m &=&- \ln \bar{s},
+  \lim_{N\rightarrow \infty} \frac{1}{N}\sum_j^N v_i &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
+  \lim_{N\rightarrow \infty} \bar{v}(N) &=& \lim_{N\rightarrow \infty}(1 - \bar{x}(N)) \\
+   \frac{1}{\mu(t)}\frac{d}{dt}\mu(t) &=&1 - \mu(t),
 \end{eqnarray*}
-where the last equality follows from a convergence in expectation.
+where we make the time dependency explicit for clarity of exposition. Now we see that we have obtained the logistic model in terms of a macroscopic ensemble average.
 
-We make special note that logistic growth emerges from pairwise interactions only.
+We make special note that logistic growth emerges from pairwise interactions only and without any independence assumption across the variables.
 This does not mean that the logistic growth cannot emerge from other microscopic settings but it nevertheless provides one possible scenario under which the logistic curve can emerge.
@@
 Thus, let $S_k$ be the set of all mutually exclusive $k$-sized groups where $k=1,..., N/2$, where the highest order are capped at $N/2$ since that would be the largest number of entities in a single group that could interact with an identically sized group.
-This means that we consider the geometric mean of each $k$-group as follows,
-\begin{equation}
-\label{eq:modSIR4}
-\frac{d}{dt} \ln x_{i_1} x_{i_2} \dots x_{i_k} = s_{j_1}s_{j_2}\dots s_{j_k}, 
-\end{equation}
-where the indices $i$ and $j$ represent the higher order interaction indices between the mutually exclusive groups of entities for each $k$ from 1 and up to $N/2$.
 Thus, the aggregated growth is represented by the sum of each interaction order,
@@
 where $S_k$ is the set of all mutually exclusive $k-$sized groups.
-In each of the sums the total number of terms is equal to the total number of groups which at each order is $N/k$ where $k$ is the interaction order.
-Thus, if we divide the right hand side by $N$, we can identify a convergence property as both $k$ and $N$ grows:
+In each of the sums, the total number of terms is equal to the total number of groups which at each order is $N/k$ where $k$ is the interaction order.
+
+Now, we can use the identical distribution assumption and divide the right hand side by $N$ to identify convergence properties as both $k$ and $N$ grow:
 \begin{enumerate}
@@
 \end{enumerate}
-Thus, as each higher order term is composed of a decreasing number of summations and an increasing number of multiplications, the product of $s_i$ goes to zero since each random variable is $iid$ and bounded between 0 and 1.
-Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic and the geometric means of the microscopic entities
+Thus, as each higher order term is composed of a decreasing number of summations and an increasing number of multiplications, the product of $s_i$ goes to zero since each random variable is bounded between 0 and 1.
+Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic means of the microscopic entities
 \begin{eqnarray}
@@
 \lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\left[\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\right]\\
-\frac{d}{dt} \ln \gamma &=&  \mu + \frac{1}{2} \mu^2 + ...\\
-\frac{d}{dt} \ln \gamma &=& - \ln (1-\mu).
+\lim_{N\rightarrow \infty} \bar{v}(N)  &=&  \mu + \frac{1}{2} \mu^2 + ...\\
+\frac{1}{\mu(t)}\frac{d}{dt}\mu(t) &=& - \ln (1-\mu). 
 \end{eqnarray}
@@
 Note that for the right hand side to exhibit asymptotically infinite growth rate at the onset of growth, it is not only sufficient but also necessary to have higher order terms.
-This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.
+This is seen by virtue of the infinite polynomial representation of the logarithm and implies that \textit{Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model}, but rather requires all higher order interactions to be present.
 
@@
 In this way one could think of the activating field as an \textit{effectively} external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism.
+Examples of this field could be sound waves or the electromagnetic field.
 
 Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case.
-To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same result can be obtained under the exclusivity condition. 
-Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model.
+To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same logistic dependency can be obtained under the exclusivity condition.
+This could be interpreted as a causality reversal of the growth process.
+Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model (see Appendices \ref{appendix:swappinglogistic} and \ref{appendix:swappinggompertz}).
 
-This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g.
+Causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g.
 \citet{estrada2022networked}).
@@
 In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
+Thus, with the necessary requirement of higher order interactions between nodes, \textit{the Gompertz model cannot be used as a communicable disease model}.
+% While some applied work show how scale-free pairwise interactions approximate Gompertz, theoretical parity is not shown (e.g. \citet{zonta2022virus}).
 
@@
 Thus, not surprisingly Gaussian growth emerges for mutually independent events that have a linear time dependence.
-The same result was originally formulated by Gauss with space as the independent variable instead of time and with error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details).
+The same result was originally formulated by Gauss with space as the independent variable instead of time and with spatial estimation error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details).
 In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.
@@
 For many growth processes within biological organisms, it has been well established that the mediator is the electromagnetic field \cite{becker1984electromagnetic}, which suggests that the accompanying growth pattern in these processes follows the Gompertz model as the result of a ubiquitous growth-inducing field.
-For growth in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts.
+For growth in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts like collective fear.
 
@@
 % M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
-
-\section{Appendix A}
+\appendix
+\section*{Appendices}
+\addcontentsline{toc}{section}{Appendices}
+\renewcommand{\thesubsection}{\Alph{subsection}}
+\subsection{A closed-form finite-interaction model}
 \label{appendix:a}
@@
 \ref{fig:hypergeometric_plain}).
+
+\subsection{The invariance of logistic growth under causality reversal}
+\label{appendix:swappinglogistic}
+Consider a causality reversal of logistic growth such that the realized growth drives potential growth. Then Equation \ref{eq:modSIR} must be modified by swapping $x$ and $s$ on the RHS as follows,
+\begin{equation}
+\label{eq:revSIR}
+\frac{d x_i}{dt} = s_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}x_{j} \quad \forall i.
+\end{equation}
+Using the identity,
+\begin{equation}
+\label{eq:modSIR2}
+\sum_i\frac{1}{s_i}\frac{dx_i}{dt} = -\sum_i\frac{1}{s_i}\frac{ds_i}{dt} = -\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ],
+\end{equation}
+and using the equivalent connectivity rule as outlined in Section \ref{sec:micro}, we have,
+\begin{equation}
+\label{eq:rev_id}
+-\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ] = \sum_j^N x_j.
+\end{equation}
+
+And using the same convergence argument as before we get $\bar{s}(N) \rightarrow \gamma(t) = 1 - \mu(t)$
+\begin{equation}
+  -\frac{1}{\gamma(t)}\frac{d}{dt}\gamma(t) = (1-\gamma(t)),
+\end{equation}
+which solves to 
+\begin{equation}
+  \gamma(t) = \frac{1}{1 + A \exp(t)},
+\end{equation}
+which implies that
+\begin{equation}
+  \mu(t) = \frac{A \exp(t)}{1 + A \exp(t)}.
+\end{equation}
+So even when realized growth is responsible for growth thus reversing causality in the pairwise growth setting, the logistic emerges.
+Thus, we can say that the \textit{microscopic logistic model is invariant under causality reversal}.
+
+\subsection{The lack of invariance of Gompertz growth under causality reversal}
+\label{appendix:swappinggompertz}
+Similar to our procedure with the logistic model we now swap potential growth with actual growth on the RHS of the microscopic Gompertz model to simulate a causality reversal,
+\begin{equation}
+\label{eq:MicroGompRev}
+-\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ] = \sum_{j\in S_1}^N x_{j} + \sum_{j_1, j_2 \in S_2} x_{j_1}x_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} x_{j_1}x_{j_2}\dots x_{j_{N/2}}.
+\end{equation}
+Make a special note of the new minus sign on the LHS due to a change of variable as done with the logistic treatment (Eq. \ref{eq:rev_id}).
+
+Now, with our continuity argument as outlined in the main text we obtain the macroscopic version of this equation as
+\begin{equation}
+  \frac{d}{dt}\ln \gamma(t) = \ln \gamma(t),
+\end{equation}
+with $\gamma(t) = 1-\mu(t)$ and where we are now rid of the minus sign on the RHS of the regular Gompertz model (Eq. \ref{eq:Gomp} due to causality reversal.
+
+This equation solves to
+\begin{equation}
+  \mu(t) = 1 - \exp\left[ A\exp(t)\right],
+\end{equation}
+which is not of the Gompertz form and furthermore does not exhibit the initially fast growth which is characteristic of the Gompertz model (see Figure \ref{fig:causality_reversal}).
+Thus we have shown that the \textit{microscopic Gompertz model is not invariant under causality reversal} and is thus not suitable for a communicable disease setting where infected nodes infect uninfected nodes.
+
+\insertSmallPdfFig{causality_reversal}{A comparison of the growth profiles of the Gompertz and the causality-reversed Gompertz models.
+Their relative growth rates defined as the time derivative of the logarithm of the abundance ($\mu(t)$) are also shown.
+To solve for the constants involved we apply the initial condition $\mu(0) = 0.001$ (a very small positive number).}
 

### 2024-04-18 arxiv2/new_draft.tex
first modification

@@
 This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities.
-This field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, or an environmental stimulant like an electromagnetic field. 
+This field could be interpreted as an environmental stressor like air-pollution, radiation, etc, or an environmental stimulant like an electromagnetic field. 
 Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
@@
 In this paper, our primary aim is to investigate the source of growth in self-organized processes.
+In particular, we posit that the source of growth can be elucidated by considering interactions in the microscopic domain which can be aggregated to recover the traditional macroscopic growth equations.
 Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
@@
 \label{eq:Rich}
-  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X^{\theta}(t)),
+  \frac{1}{X}\frac{d}{dt} X(t) = \frac{\beta}{\theta} (1 - X^{\theta}(t)),
 \end{equation}
@@
  The emergence of Gompertz growth from higher order logistic growth can be seen as another way to encode non-linear effects that were seen as coherence in accordance with Petroni and Molski.
-In practice, the augmented logistic model has a faster growth rate decay than the familiar $\theta$-logistic growth (Figure \ref{fig:hypergeometric_plain}).
+In practice, the augmented logistic model has a slower growth rate decay than the familiar $\theta$-logistic growth (Figure \ref{fig:hypergeometric_plain}).
 If we interpret deceleration of relative growth as degree of coherence, the augmented logistic growth would be more coherent than the $\theta$-logistic model for a given starting value.
@@
 \end{equation}
-where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$.
+where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$, excluding node $i$ itself.
 For example, if the adjacency matrix has only $1$s and no $0$s, this means that every node is connected to every other node.
@@
 \begin{equation}
-\label{eq:modSIR2}
+\label{eq:modSIR22}
 \frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
@@
 \begin{equation}
-\label{eq:modSIR2}
+\label{eq:modSIR2app}
 \sum_i\frac{1}{s_i}\frac{dx_i}{dt} = -\sum_i\frac{1}{s_i}\frac{ds_i}{dt} = -\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ],

### 2024-04-18 arxiv2/new_draft.tex
address Theas comments

@@
 This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities.
-This field could be interpreted as an environmental stressor like air-pollution, radiation, etc, or an environmental stimulant like an electromagnetic field. 
+This field could be interpreted as an environmental stressor like air-pollution, electromagnetic radiation, etc.
 Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.
@@
 \begin{itemize}
-\item Logistic and Gompertz growth is connected with a novel framework based on microscopic principles.
-\item Gompertz growth is connected with ubiquitous growth-activating field justified by physical and mathematical principles.
-\item Growth processes is put in a larger contextual framework based on varying dependence in spacetime, including the Exponential, Gaussian and Poisson distributions. 
+\item Logistic and Gompertz growth is connected within a novel framework based on microscopic principles.
+\item Gompertz growth is connected with a ubiquitous growth-activating field justified by physical and mathematical principles.
+\item Growth processes are put in a larger contextual framework based on varying dependence in spacetime, including the Exponential, Gaussian and Poisson distributions. 
 \end{itemize}
@@
 Equivalently, mortality rates increase in geometric progression for arithmetic increases in age, an observation now commonly referred to as Gompertz' law of mortality \cite{Gompertz1825}.
-Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time \cite{makeham1889further}.
+Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time as a ``stressor'' \cite{makeham1889further}.
 Gompertz' law has later been known as the Gompertz model or Gompertz growth.
@@
 For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.
-
 In fact, it has been shown that cancer malignancies are mostly \textit{negative} in electric polarity with respect to the rest of the body, while nonmalignant (or benign) tumor pathology is mostly \textit{positive} in polarity \cite{langman1949technique}.
 Gompertz growth has also been widely observed in cancerous cells \cite{kozusko2003combining,laird1964dynamics}.
-One is then led to the interpretation that certain biological growth processes that exhibit Gompertz growth are coupled with an electromagnetic field of specific character.
+One is then led to the interpretation that certain biological growth processes that exhibit Gompertz growth possibly are coupled with an electromagnetic field of specific character.
 Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.
@@
 On the topic of disease modelling, the Gompertz model has been shown by Wang et al.
-\cite{wang2012richards} to be excluded from a subset of models in the communicable disease paradigm.
-They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario.
+\cite{wang2012richards} to be incompatible with a subset of models in the communicable disease paradigm.
+They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range within the this model does not cover the parameter value taken in the Gompertz scenario.
 If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near-Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}.
@@
 
-\subsection{Main Results and Roadmap}
+\subsection{Main aim}
 In this paper, our primary aim is to investigate the source of growth in self-organized processes.
 In particular, we posit that the source of growth can be elucidated by considering interactions in the microscopic domain which can be aggregated to recover the traditional macroscopic growth equations.
-Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
+Second, we will elaborate on the idea that the logistic model results from local internal interactions while the Gompertz model results from a possibly external field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth.
 We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g.
@@
 In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth.
-Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz and Poisson.
+Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz, and Poisson by considering independence between entities and other spatio-temporal dependencies on growth.
 
+\subsection{Roadmap}
 Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which offers an alternative to the traditional Richards model.
@@
 \end{equation}
-In this limit, the relative growth goes to infinity as $\lim_{t\to 0}$, which could be seen as a verification of the maximally coherent characteristic of the system.
+In this limit, the relative growth rate goes to infinity as $\lim_{t\to 0}$, which could be seen as a verification of the maximally coherent characteristic of the system.
 One is reminded of a system that is perturbed and left to reach a new equilibrium.
@@
 The right hand side ``resource'' term is not specific enough so as to specify the location of the source.
-One could argue that the since the Gompertz model has asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal (or catalyst in Markov's language) that initiates the growth and subsequently decays exponentially like a memoryless process.
+One could argue that since the Gompertz model has an asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal (or catalyst in the language of \citet{markov2019reaction}) that initiates the growth and subsequently decays exponentially like a memoryless process \citep{durrett1999essentials}.
 To further explore the source of growth we turn to the microscopic domain.
@@
 
-Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}.
+Now as we let N grow, we impose an exclusivity rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}.
 This condition translates into an adjacency matrix with each column containing a single non-zero entry.
@@
 Instead of limiting the growth equation to pairwise interactions, we allow higher order interactions in the form of triplets, quadruplets, and so on.
-Analogous to the pairwise exclusivity rule, each entity can again only participate in \textit{one} interaction group at any given order and any given direction.
+Analogous to the pairwise exclusivity rule, \textit{each entity can again only participate in one interaction group at any given order and any given direction}.
 For example, any triplet can only interact with one other triplet and have exclusive possession of its entities at the triplet level (Figure \ref{fig:order}).
@@
 In this way one could think of the activating field as an \textit{effectively} external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism.
-Examples of this field could be sound waves or the electromagnetic field.
+Examples of such a field could be sound waves or an electromagnetic field.
 
-Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case.
+Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms originate from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case.
 To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same logistic dependency can be obtained under the exclusivity condition.
@@
 
-Causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g.
+Causality reversal is indeed necessary when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g.
 \citet{estrada2022networked}).
-But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head.
-In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
+Higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head.
+In this context, the emergence of Gompertz growth would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.
 Thus, with the necessary requirement of higher order interactions between nodes, \textit{the Gompertz model cannot be used as a communicable disease model}.
@@
 
-\section{Unifying the Gaussian and Exponential curves with the Logistic and the Gompertz curves}
+\section{Unifying the Gaussian, Exponential, and Poisson models with the Logistic and the Gompertz models}
+
+\subsection{Unification with the Gaussian and Exponential models}
 
@@
 where $\beta$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$.
-Thus, not surprisingly Gaussian growth emerges for mutually independent events that have a linear time dependence.
+Thus, not surprisingly, Gaussian growth emerges for mutually independent events that have a linear time dependence.
 The same result was originally formulated by Gauss with space as the independent variable instead of time and with spatial estimation error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details).
 In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.
+
+In general, our choice of time as the dependent variable can be substituted for any suitable metric of choice.
+Another example of space as the dependent variable in the Gompertz setting is given by \citet{molski2003coherent}.
 
@@
 \end{equation}
-This exponential equation implies a complete independence on both time and of the mutual entities involved in the growth.
-In the context of measurement error this function would be the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}.
+This exponential equation implies a complete independence on both time and of the entities involved in the growth.
+In the context of measurement error this function would represent the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}.
 % Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable where the physical interpretation would be a decay process.
-
-In general, our choice of time as the dependent variable can be substituted for any suitable metric of choice.
-Another example of space as the dependent variable in the Gompertz setting is given by \citet{molski2003coherent}.
 
@@
 By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}.
-This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system.
+This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection of the Poisson distribution with a coherent system.
 Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here.
@@
 
-An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be a ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}.
-With a such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli.
+Growth of infinitely interacting systems where entities are classical in nature with well-defined position and momentum breaks the law of causality unless we allow the source of growth to be a ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}.
+With such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli.
 Examples of such models are already given by De Lauro (2014), Sacher (1956) and Markov (2019).
-\citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic in nature, and while Lauro et al.
+\citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic or random in nature, and while Lauro et al.
 do not discuss the external nature of such a source, their model does not preclude it.
-More generally, with growth as stochastically rooted it is also ubiquitously present uniformly across all entities from the very onset of growth.
+More generally, with growth as stochastically rooted the source is also ubiquitously present uniformly across all entities from the very onset of growth.
 Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Lauro and Sacher's models.
@@
 In this context the ubiquitous stressor postulated above would be the living conditions on our planet, for which time acts as a proxy.
-One could further postulate this stressor as a field surrounding the earth, e.g.
-in its electromagnetic footprint often related with the Schumann Resonance \cite{cherry2002schumann,schumann1952strahlungslosen}.
 It is indeed curious that we do not see mortality rates behave as linearly dependent on time, for which a Gaussian curve would be appropriate.
 Nor do we see any finite-ordered polynomial time dependence on mortality rates.
-Instead we see growth of mortality rates as sourced from other entities at all interaction orders, which in light of our discussion above is equivalent to mortality coming from an environmental stressor.
+Instead we see growth of mortality rates as sourced from other entities at all interaction orders, which in light of our discussion above is equivalent to mortality triggered by an environmental stressor.
 
